---
title: "03 Global gene modules and block annotation"
author: "Saumya Pothukuchi"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
#    number_sections: true #adds numbers to headers
    code_folding: hide
---


```{r}
library(Seurat)
# run conda install conda-forge::zstd command line if there is a "UserWarning: zstandard could not be imported. Running without .conda support" errro + restart R session
# unloadNamespace("anndata")
# unloadNamespace("reticulate")
# unloadNamespace("Matrix")
# unloadNamespace("Seurat")
library(reticulate)
# point reticulate at the env built from env/metacell.yml
reticulate::use_condaenv("metacell", required = TRUE)
library(anndata)
library(ggplot2)
library(ggrepel)
library(SCpubr)
```

# Configuration

```{r, results = FALSE, message = FALSE, warning = FALSE}
# Paths, factor levels and palettes all come from config/config.R so this
# notebook and the composition / DE steps cannot drift apart.
source(file.path("..", "config", "config.R"))

SUBSET_SLUG <- "global"
FILE_PREFIX <- "global"

# where this step reads from and writes to
h5ad_dir <- file.path(DIR_H5AD, SUBSET_SLUG, "metacell")
rds_dir  <- out_dirs(SUBSET_SLUG, "rds")
csv_dir  <- out_dirs(SUBSET_SLUG, "csv")
fig_dir  <- out_dirs(SUBSET_SLUG, "pdf")

# the two objects written by 02
CELLS_H5AD <- file.path(h5ad_dir, "project.clean_cells.h5ad")
MC_H5AD    <- file.path(h5ad_dir, "project.metacells.h5ad")

# Number of gene modules cut from the gene-gene correlation tree. Set here
# because it appears in every intermediate filename as well as in the cutree
# call, and the two must not drift.
NCLUST <- 100

# helper so every intermediate lands in rds_dir with a consistent stem
rdsf <- function(tag, ext = "rd")
  file.path(rds_dir, sprintf("%s_%s_%dmods.%s", FILE_PREFIX, tag, NCLUST, ext))
```

# Part 1. Cell-level QC

Plots only. Cells were already filtered upstream, so nothing here drops anything;
the point is to see whether one sample sits apart from the others before its
metacells are trusted. See docs/13_diagnostics.md.

## Output directory for QC figures

```{r}
output_dir <- fig_dir
```

## Read the cell object
```{r}
cdata = anndata::read_h5ad(CELLS_H5AD)
```


## Prepare the matrix
```{r}
colnames(cdata$X) = cdata$var_names
rownames(cdata$X) = cdata$obs_names
remvlist <- which(Matrix::colSums(cdata$X) == 0) #removes absent genes
mdatarm <- cdata$X[, -remvlist]

mdatarm
```


## Mitochondrial fraction per sample
```{r}
mydata <- data.frame(Perct_mito = cdata$obs$percent_mito, Cond = cdata$obs$sample_origin)

ggplot(mydata, aes(x = Cond, y = Perct_mito)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Sequencing depth per sample
```{r}
mydata <- data.frame(Reads = cdata$obs$total_umis, Cond = cdata$obs$sample_origin)

ggplot(mydata, aes(x = Cond, y = Reads)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Genes detected per cell
```{r}
cdata$obs$n_genes <- Matrix::rowSums(cdata$X > 0)

ggplot(cdata$obs, aes(x = sample_origin, y = n_genes)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Part 2. Gene modules

Everything below runs on the metacell object, not the cell object. Genes that
co-vary across metacells are grouped into modules, and each metacell is scored on
every module. Method and the reasoning behind each threshold:
docs/07_annotation.md and docs/06_parameters.md.

```{r}
output_dir <- fig_dir
```

## Read the metacell object

```{r}
mdata = anndata::read_h5ad(MC_H5AD)
```

## Prepare the matrix
```{r}
colnames(mdata$X) = mdata$var_names
rownames(mdata$X) = mdata$obs_names
remvlist = which(Matrix::colSums(mdata$X)==0) 
#remvlist
mdatarm = mdata$X[, -remvlist] 

#Useful in the next steps
total = CreateSeuratObject(t(mdatarm), min.cells = 0, min.features = 0) 
```

## Mean-dispersion LOESS fit and HVG selection

```{r}
ds = t(mdatarm) #The matrix is transposed so that genes become rows and metacells columns
ds_mean = Matrix::rowMeans(ds) #Calculates the mean expression of each gene across all metacells
ds_var = apply(ds, 1, var) #Calculates the variance of each gene across all metacells

varmean_df = data.frame(m = ds_mean, v = ds_var, gene = rownames(ds))   # Creation of a data frame containing for each gene, its name ("gene") as well as its mean expression ("m") and variance ("v") across all metacells
```

```{r}
x = log10(varmean_df$m) #Abundance: Applies log10 to the mean expression of each gene. Gene expression is normalized by M 0.9.5 so that the sum of gene expression within each metacell is 1, so mean expression of each gene across all metacells is going to be < 1 (mean of frequencies), and the logarithm is going to be negative, which allows a better spreading of mean expression values than with a linear scale
lv = log2(varmean_df$v / varmean_df$m) #Dispersion = log2(variance/mean). Variance rises with the mean, so ranking on variance alone selects only abundant genes.
breaks = seq(min(x), max(x),.2)   # Segments the range of values of x (=log10(mean)) for loess (LOcally Estimated Scatterplot Smoothing) curve calculation. Each range now includes genes with similar level of expression
z = sapply(split(lv, cut(x, breaks)), min, na.rm = T) #Minimum dispersion per expression bin: the technical noise floor at that abundance, i.e. a non-variable gene.
maskinf = is.infinite(z) #Remove intervales with infinite minimum dispersion
z = z[!maskinf]
b = breaks[-length(breaks)] #Keep only the corresponding interval boundaries
b = b[!maskinf]
lo = loess(z~b) #Calculation of the loess curve for z (the minimum value of dispersion for a given interval). Models the background noise as a function of the expression level.

inVarMean_MeanThresh = -9
# example: on one global run range(x) was about -10.8 to -1.6

#Aiming for 3k HVGs isntead of standard 2k
inVarMean_varmeanThresh = 3.67

n_genes <- function(thresh){
  lline2 = predict(lo, newdata = log10(varmean_df$m))
  sum(log10(varmean_df$m) > as.numeric(inVarMean_MeanThresh) & log2(varmean_df$v/varmean_df$m) > lline2 + as.numeric(thresh), na.rm = TRUE)
}

thresholds <- seq(3.65, 3.7, by = 0.01)
sapply(thresholds, n_genes)
```

```{r}
#pdf(paste0(output_dir, "mean_dispersion_loess.pdf"))
plot(x, lv, xlab = "Log10(mean)", ylab = "log2(var/mean)", panel.first = grid())   # Make a data plot with genes as dots according to their log10(mean) and dispersion as coordinates
x1 = seq(-10, 0, l = 100)   # Creates a sequence for the range of the X axis that will be used for the loess curve (smaller intervals than previously with 'breaks', about 100 intervals)
lline = predict(lo, newdata = x1)   # If range(x) This predicts a new loess curve for each x1 value
lines(x1, lline + as.numeric(inVarMean_varmeanThresh), col = 2)   # Draws a line joining all the points at LOESS + threshold
abline(v = inVarMean_MeanThresh, col = 2)   # Draws a vertical line for the other threshold (minimum mean expression to select a gene for module calculation). Prevents genes with very low expression from being selected.
lline2 = predict(lo, newdata = log10(varmean_df$m)) #Actually predicts the minimal dispersion expected for each gene
#dev.off()
```

```{r}
geneModuleMask = log10(varmean_df$m)>as.numeric(inVarMean_MeanThresh) & log2(varmean_df$v/varmean_df$m)>lline2+as.numeric(inVarMean_varmeanThresh)   # Create a mask of the gene modules that will actually be used for the calculation of gene modules
table(geneModuleMask)

effivar = rownames(varmean_df)[which(geneModuleMask==T)]   # Get HVG gene names

# carry lateral/noisy masks into the metacell-level module HVG selection
get_mask <- function(v, col){ m <- as.logical(v[[col]]); m[is.na(m)] <- FALSE; rownames(v)[m] }
mask_src  <- if (all(c("lateral_gene","noisy_gene") %in% colnames(mdata$var))) mdata$var else cdata$var
lat_noisy <- union(get_mask(mask_src, "lateral_gene"), get_mask(mask_src, "noisy_gene"))


n0 <- length(effivar)
effivar <- setdiff(effivar, lat_noisy)
cat(sprintf("module HVGs: %d -> %d after dropping %d lateral/noisy\n", n0, length(effivar), n0 - length(effivar)))

ds_select <- ds[effivar, ]

min(rowSums(ds[which(rownames(ds)%in%effivar),])) # example value from one run: ~1e-05
ds_select = ds[effivar,] # Subset the whole dataset to contain only the selected genes (as rows, and all the metacells as columns)

# geneModuleMask
# FALSE  TRUE 
```

## Gene-gene correlations and module cutting

```{r}
nclust <- NCLUST
res2 = cor(as.matrix(t(ds_select)), method = c("pearson"))   # Pearson correlation between all gene pairs (matrix is transposed, so columns are genes)
res2_dist = dist(res2, method = 'euclidean')   # Returns the distance matrix of the correlation matrix (= computes the euclidean distances between the rows of the correlation matrix) => 2 genes with very similar correlation values to each other gene will be "close"
res2_clust = hclust(res2_dist, method = 'complete')   # Compute hierarchical clustering of the genes based on the distance matrix
res2_cut = cutree(res2_clust, k = nclust)   # Cuts the tree resulting from the hierarchical clustering so as to produce k (= 100) clusters - result is a named vector indicating for each gene the cluster it belongs to (number, index-like)
```

## Gene to module table
```{r}
dfcut = as.data.frame(names(res2_cut))   # 1986 HVGs post filtering
dfcut$related_genes_module = res2_cut   # Adds a column with the corresponding cluster (= module) for each gene
colnames(dfcut) = c("X","related_genes_module")   # Renames columns for the dataframe (first one named "X" now, and second one "related_genes_module")
```

## Gene module objects

```{r}
newmod = c()   # create null vector
for(i in 1:nclust){
  newmod = append(newmod, paste0("My_Mod_",i))
}   # Creates a vector containing the names of gene modules (My_Mod_<nb>)

for(i in levels(as.factor(dfcut$related_genes_module))){   # For each module (strings here, instead of numbers in the previous loop) :
  assign(paste0("My_Mod_",i), dfcut$X[which(dfcut$related_genes_module==i)])   # Create a vector named after the module, containing the names of the genes it contains
}

allmod = c()
alllist = list()   # Create an empty vector and an empty list
for(i in 1:nclust){   # For each cluster :
  submod = data.frame(Mod=paste0(newmod[i]),Gen=paste0(eval(as.name(newmod[i]))))   # eval() extracts the name of the module (newmod[i]) as an object name in R, so basically this creates a n-row dataframe where the 1st column is always the name of the gene module, and the 2nd column contains all the names of the genes inside that module (long format data.frame)
  allmod = rbind(allmod,submod)   # This new data.frame is added to the aggregated object with all the previous ones (empty if first iteration)
  alllist[i] = list(paste0( paste0(eval(as.name(newmod[i])),collapse=",") ) )   # Creates an additional string in the alllist, containing the names of all the genes from the corresponding gene module, separated by comas
}   # allmod and alllist hold the same gene-per-module information in two shapes

genelist = data.frame(Mods = newmod, Gens = unlist(alllist))   # Combines newmod (vector of module names) and alllist (list of strings, each including all genes of a module) in a single data.frame
save(genelist, file = rdsf("genelist"))
```

## Module scores across metacells

```{r}
# First remove previous module scores if they exist
cols_remove <- grepl("^My_Mod_.*_(score|scaled)$", colnames(total@meta.data))
total@meta.data <- total@meta.data[, !cols_remove]

for(i in newmod){   # For each module :
  print(paste0("---- Working on ", i, " ----"))
  ilist = unlist(strsplit(genelist$Gens[which(genelist$Mods==i)],split = ","))   # Extracts the list of genes within the considered module (by splitting the string from genelist) and unlists it (= converts it into a vector)
  
  if(length(ilist)<2) {   # If there are less than 2 genes in this module (so only 1 basically) :
    values = ( log(1+total@assays$RNA$counts[intersect(ilist,rownames(total)),]) / colSums(log(1+total@assays$RNA$counts)) )   # Same as below but only for this one gene, so no need for colSums() at the numerator
    eval(parse(text=paste0("total@meta.data$",i,"_score = values")))
  } else {   # Otherwise, if more than 1 gene :
    values = ( colSums(log(1+total@assays$RNA$counts[intersect(ilist,rownames(total)),])) / colSums(log(1+total@assays$RNA$counts)) )   # Obtains a numeric vector : ratio between [total expression of the genes of this module in each metacell (named numeric vector)] and [total expression of all non-excluded genes (25000 from the beginning)]   
    eval(parse(text=paste0("total@meta.data$",i,"_score = values")))   # Eval here allows to enter the text as if it were a command : adds the content of the "values" object (= for each metacell, fraction of total non-excluded gene expression attributed to genes of this module) to a slot (i) in the metadata of the total Seurat object
  }
}

for(i in newmod){   # For each module :
  eval(parse(text = paste0("total@meta.data$", i, "_scaled = as.vector(scale(total@meta.data$", i, "_score, center = TRUE, scale = TRUE))")))   # Add a slot for the corresponding gene module in the Seurat metadata, which is the scaled value of the scores of the corresponding module across metacells = centers and scales (z-score) within each module
}
```

## Save the gene x metacell matrix

```{r}
dgcmtxcounts = as(total@assays$RNA$counts, "dgCMatrix")
save(dgcmtxcounts, file = rdsf("counts"))
```

## Extract metadata

```{r}
metadf = as.data.frame(total@meta.data)   # Extract the metadata as a data.frame
for(i in newmod){   # For each module :
  eval(parse(text = paste0("metadf$", i, "_score = NA")))
  eval(parse(text = paste0("metadf$", i, "_score[match(colnames(total),rownames(metadf))] = total@meta.data$", i, "_score")))   # The match() part is useful in case there is a discrepancy between the metacells in the Seurat object and in the metadata df (don't see why that would happen but I guess it did at some point)
}   # Replaces the score values in the metadata df with those of the Seurat object metadata (again, don't see why they would not correspond, but I guess some discrepancy happened at some point which required this kind of regularization)

metadf_z = metadf   # already z-scored in the module scoring chunk above
save(metadf_z, file = rdsf("metadata"))
```

# Part 3. Module score heatmap and metacell blocks

The module-score matrix is cut into `column_km` k-means groups. Those groups are
the blocks, and the blocks are what get cell type labels. `column_km` is the
resolution knob and is tuned by looking: see docs/07_annotation.md.

## Libraries and output directories
```{r}
library(heatmaply)
library(ggplot2)
library(scales)
library(ComplexHeatmap)
library(circlize)

output_dir_res <- csv_dir
output_dir_fig <- fig_dir
```

## Load objects

```{r}
load(rdsf("counts"))   # dgcmtxcounts
load(rdsf("genelist"))   # genelist
load(rdsf("metadata"))   # metadf_z
```

## Build the heatmap and cut blocks

```{r}
hm_zmod <- metadf_z[, grepl("_scaled$", colnames(metadf_z))]

thmz = t(hm_zmod)   # Transposes the selected metadata matrix : metacells are columns now, and module scores are rows
set.seed(42)   # Set seed
col_fun <- colorRamp2(
  c(-2, 0, 2),
  c("navy", "white", "firebrick")
)
#column_km = 40, means 40 metacell groups, aka split metacells into 40 groups. This is like the "res" parameter, and is adjusted for when the heatmap looks "good"
ht = draw(Heatmap(thmz, col=col_fun, column_title = "Module clustering based on corrected z score", name = "mat",
                  row_names_gp = gpar(fontsize = rel(400/length(colnames(hm_zmod)))), show_row_dend = TRUE,
                  row_dend_reorder = FALSE, cluster_rows = TRUE, row_order = NULL,
                  column_dend_reorder = FALSE, cluster_columns = TRUE, show_column_names = FALSE, column_names_gp = gpar(fontsize = rel(400/length(rownames(hm_zmod)))), column_km = 40, row_km = 0, use_raster = TRUE) )   # Draws the heatmap from the matrix
save(ht, file = rdsf("HMlist"))
```

```{r}
# pdf(file = paste0(output_dir_fig, paste0(FILE_PREFIX, "_HM_", NCLUST, "mods.pdf")), width = 12, height = 8  )
# ht
# dev.off()
```

```{r}
htmxord = thmz[ as.vector(unlist(row_order(ht))), as.vector(unlist(column_order(ht))) ]   # Reorders rows and columns of the thmz matrix according to the order on the heatmap (was reordered due to clustering)
save(htmxord, file = rdsf("htmxord"))
```


# Pull which genes are in the module

```{r}
newmod <- genelist$Mods
sapply(newmod, function(m) paste(head(unlist(strsplit(genelist$Gens[genelist$Mods==m], ",")), 8), collapse=", "))

# Get info regarding 40 metacell groups/blocks, like how many metacells are in each block.
cl <- column_order(ht)  # a list of 40 vectors of column indices
block <- integer(ncol(thmz)); for (k in seq_along(cl)) block[cl[[k]]] <- k
names(block) <- colnames(thmz)   # metacell -> block number (1..40)
table(block)

# Map block to module, to understand which modules are driving which blocks - helps in anntoations
blockmeans <- t(apply(thmz, 1, function(r) tapply(r, block, mean)))  # module × block mean z
apply(blockmeans, 2, function(col) names(sort(col, decreasing=TRUE))[1:5])  # top 5 modules per block
```
```{r}
# Tab 1: every module and its full gene list
modules_genes <- data.frame(
  module = genelist$Mods,
  genes  = gsub(",", ", ", genelist$Gens),   # full list, comma-space separated
  stringsAsFactors = FALSE
)

# Tab 2: all modules per block, ranked by descending block mean z (no top-5 cap)
block_rank <- as.data.frame(
  apply(blockmeans, 2, function(col) names(sort(col, decreasing = TRUE))),
  stringsAsFactors = FALSE
)
colnames(block_rank) <- paste0("block_", seq_len(ncol(block_rank)))
# drop the _scaled suffix to match tab 1 module names:
# block_rank[] <- lapply(block_rank, sub, pattern = "_scaled$", replacement = "")

# Both tabs, one workbook
out_xlsx <- file.path(csv_dir, paste0(FILE_PREFIX, "_module_block_map.xlsx"))
openxlsx::write.xlsx(
  list(modules_genes = modules_genes, block_module_ranking = block_rank),
  file = out_xlsx, rowNames = FALSE
)
```


# Annotation heatmap with global markers

```{r}
library(ComplexHeatmap); library(circlize)

# PBMC lineage markers. "CD4","CD40LG","FOXP3" not picked up
panel <- list(
  "T cells"      = c("CD2","CD7","CD3G","CD3D","CD3E","IL7R","CD8A","CD8B","GZMK","GZMB","GNLY","NKG7","PRF1"),
  "NK"           = c("KLRF1","NCAM1","SPON2","KLRD1","TYROBP"),
  "B cells"      = c("CD19","MS4A1","CD22","PAX5","CD79B","BANK1","EBF1"),
  "Plasma cells" = c("MZB1","XBP1","JCHAIN","TNFRSF17"),
  "Granulocytes" = c("MS4A2","CPA3","HDC","GATA2"),
  "Mo/Macs"      = c("CD14","LYZ","FCN1","CSF1R","S100A9","CD68", "NLRP3", "ACSL1", "BCL2A1", "ICAM1", "NINJ1"),
  "moDCs"        = c("ITGAX","HLA-DRA","HLA-DQA1","CD1C","FCER1A","CLEC10A"),
  "pDCs"         = c("IL3RA","TLR9","LILRA4","CLEC4C"),
  "Platelet"     = c("PPBP","PF4"),
  "Cycling"      = c("MKI67","STMN1")
)

# lineage-specific scoring signatures
# lineage-specific scoring signatures
sig <- lapply(list(
  "T cells"      = c("CD3D","CD3E","CD2"),
  "NK"           = c("KLRF1","NCAM1","SPON2"),          # no TYROBP/KLRD1: both are high in Mo/Mac
  "B cells"      = c("CD19","MS4A1","BANK1"),
  "Plasma cells" = c("MZB1","XBP1"),
  "Granulocytes" = c("MS4A2","CPA3","HDC","CD177","ALPL","CYP4F3","ADGRG3"),  # was mast/baso only
  "Mo/Macs"      = c("CD14","LYZ","FCN1","NLRP3","ACSL1"),
  "moDCs"        = c("CD1C","FCER1A","CLEC10A"),
  "pDCs"         = c("IL3RA","LILRA4","CLEC4C"),
  "Platelet"     = c("PPBP","PF4"),
  "Cycling"      = c("MKI67","STMN1")
), function(g) intersect(g, rownames(dgcmtxcounts)))
print(lengths(sig)); stopifnot(all(lengths(sig) > 0))

markers <- unlist(panel, use.names = FALSE)
row_lin <- rep(names(panel), lengths(panel))
keep    <- markers %in% rownames(dgcmtxcounts)
markers <- markers[keep]; row_lin <- factor(row_lin[keep], levels = names(panel))

veto_T  <- intersect(c("CD3D","CD3E","CD3G","CD2","TRAC"), rownames(dgcmtxcounts))
veto_NK <- intersect(c("KLRF1","NCAM1","SPON2"), rownames(dgcmtxcounts))

score_genes <- union(markers, unlist(c(sig, list(veto_T, veto_NK)), use.names = FALSE))
libsize <- Matrix::colSums(dgcmtxcounts)
lognorm <- as.matrix(log1p(t(t(dgcmtxcounts[score_genes, ]) / libsize * 1e4)))
stopifnot(identical(colnames(lognorm), names(block)))

blk   <- t(apply(lognorm, 1, function(r) tapply(r, block, mean)))
score <- sapply(sig, function(g) colMeans(blk[g, , drop = FALSE]))
z <- scale(score); z[is.na(z)] <- 0
stopifnot(identical(rownames(z), colnames(blk)), identical(colnames(z), names(sig)))

call <- colnames(z)[max.col(z, "first")]
s    <- t(apply(z, 1, sort, decreasing = TRUE))
call[s[,1] < 0] <- "Undefined"

# CD3 decides the T/NK boundary in both directions, on absolute log-CP10K
CD3_MIN <- 1.0
cd3 <- colMeans(blk[veto_T,  , drop = FALSE])
nk  <- colMeans(blk[veto_NK, , drop = FALSE])
bad_T  <- call == "T cells" & (cd3 < CD3_MIN | nk > cd3)
call[bad_T & nk >= CD3_MIN] <- "NK"
call[bad_T & nk <  CD3_MIN] <- "Undefined"
bad_NK <- call == "NK" & cd3 >= CD3_MIN & cd3 > nk
call[bad_NK] <- "T cells"
cat("T vetoed:", sum(bad_T), " NK vetoed:", sum(bad_NK), "\n")

block_anno <- setNames(call, rownames(z))
print(block_anno)
# every block, not just the flagged ones, so this does not need another pass
diag <- data.frame(block = rownames(z), call = call,
                   best = round(s[,1], 2), margin = round(s[,1] - s[,2], 2),
                   cd3 = round(cd3, 2), nk = round(nk, 2),
                   umis = round(tapply(libsize, block, median)))
print(diag[order(diag$call, -diag$best), ], row.names = FALSE)


# order + split columns by lineage
lin_order <- c(names(panel), "Undefined")
anno <- factor(block_anno[as.character(block)], levels = lin_order)
ord  <- order(anno, block); ord <- ord[!is.na(anno[ord])]

# color: row z (diverging) or absolute (quantile viridis)
scale_rows <- TRUE
if (scale_rows) {
  mat <- t(scale(t(lognorm[markers, ord]))); mat[mat > 2.5] <- 2.5; mat[mat < -2.5] <- -2.5
  col_fun <- colorRamp2(c(-2,0,2), c("#3B4CC0","#F7F7F7","#B40426")); hm_name <- "z-scaled"
} else {
  mat <- lognorm[markers, ord]
  col_fun <- colorRamp2(quantile(mat, c(0,.5,.9,.99), na.rm=TRUE), viridisLite::viridis(4)); hm_name <- "logExpr"
}

lineage_cols <- setNames(
  c("#8B1A1A","#4E9E2F","#2C6FAD","#E69F00","#2D6A4F","#9E79C4","#56B4E9","#A0522D","#CC79A7","#8C8C8C","#B2182B"),
  lin_order)
top <- HeatmapAnnotation(compartment = anno[ord], col = list(compartment = lineage_cols),
                         show_annotation_name = FALSE)

ht2 <- Heatmap(mat, name = hm_name, col = col_fun, top_annotation = top,
        row_split = row_lin, column_split = anno[ord], column_title = NULL,
        cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = FALSE,
        row_names_gp = gpar(fontsize = 7), row_title_gp = gpar(fontsize = 8), 
        row_title_rot = 0,  
        column_gap = unit(0.5,"mm"), row_gap = unit(0.5,"mm"), use_raster = TRUE)

draw(ht2)
```


```{r}
#Write cell type annotation to block mapping to same file
celltype_block_map <- data.frame(
  block = rownames(z),
  assigned_celltype = block_anno[rownames(z)],
  best_score = apply(z, 1, max),
  second_score = apply(z, 1, function(x) sort(x, decreasing = TRUE)[2]),
  score_margin = apply(z, 1, function(x) diff(sort(x, decreasing = TRUE)[1:2])),
  top_module = apply(blockmeans[, rownames(z), drop = FALSE], 2, function(x) names(which.max(x))),
  stringsAsFactors = FALSE
)

out_xlsx <- file.path(csv_dir, paste0(FILE_PREFIX, "_module_block_map.xlsx"))

openxlsx::write.xlsx(
  list(
    modules_genes = modules_genes,
    block_module_ranking = block_rank,
    block_celltype_mapping = celltype_block_map
  ),
  file = out_xlsx,
  rowNames = FALSE
)
```


```{r}
#Add block and cluster label info to seurat obj
cdata_cell <- anndata::read_h5ad(CELLS_H5AD)
counts_cell <- Matrix::t(cdata_cell$X)
counts_cell <- as(counts_cell, "CsparseMatrix")   # dgRMatrix -> dgCMatrix

colnames(counts_cell) <- rownames(cdata_cell$obs)
rownames(counts_cell) <- rownames(cdata_cell$var)

# X must be raw UMIs, not normalized
stopifnot(inherits(counts_cell, "dgCMatrix"))
stopifnot(nrow(counts_cell) == nrow(cdata_cell$var),
          ncol(counts_cell) == nrow(cdata_cell$obs))
stopifnot(all(counts_cell@x == floor(counts_cell@x)))

pbmc <- CreateSeuratObject(counts = counts_cell, meta.data = as.data.frame(cdata_cell$obs))
stopifnot(identical(colnames(pbmc), rownames(cdata_cell$obs)))

# only name block if cutree did not already
if (is.null(names(block))) {
  stopifnot(length(block) == ncol(thmz))
  names(block) <- colnames(thmz)
} else {
  stopifnot(setequal(names(block), colnames(thmz)))
}

mc_name    <- as.character(pbmc$metacell_name)
mc_int     <- as.integer(as.character(pbmc$metacell))
mc_outlier <- !is.na(mc_int) & mc_int < 0

mc_block <- as.character(unname(block[mc_name]))
mc_block[mc_outlier]  <- "Outlier"
mc_block[is.na(mc_block)] <- "Unassigned"

broad_label <- unname(block_anno[mc_block])
broad_label[mc_block %in% c("Outlier", "Unassigned")] <- mc_block[mc_block %in% c("Outlier", "Unassigned")]
broad_label[is.na(broad_label)] <- "Unassigned"

blk_idx <- match(mc_block, as.character(celltype_block_map$block))

pbmc <- AddMetaData(pbmc, data.frame(
  mc_name         = mc_name,
  mc_block        = mc_block,
  mc_outlier      = mc_outlier,
  broad_label     = broad_label,
  mc_anno_score   = celltype_block_map$best_score[blk_idx],
  mc_anno_margin  = celltype_block_map$score_margin[blk_idx],
  mc_top_module   = celltype_block_map$top_module[blk_idx],
  row.names       = colnames(pbmc),
  stringsAsFactors = FALSE
))

mc_lookup <- data.frame(
  metacell_name = names(block),
  mc_block      = unname(block),
  broad_label   = unname(block_anno[as.character(unname(block))]),
  stringsAsFactors = FALSE
)
write.csv(mc_lookup, file.path(csv_dir, paste0(FILE_PREFIX, "_metacell_block_annotations.csv")), row.names = FALSE)

cat("Metacells in module heatmap (thmz):     ", length(block), "\n")
cat("Metacells assigned to >=1 cell:         ", length(unique(mc_name[!mc_outlier])), "\n")
cat("Assigned, block-labelled metacells:     ", length(unique(mc_name[!mc_outlier & !mc_block %in% c("Outlier","Unassigned")])), "\n")
cat("Cells total:                            ", ncol(pbmc), "\n")
cat("  in metacell:                          ", sum(!mc_outlier), "\n")
cat("  MC2 outliers:                         ", sum(broad_label == "Outlier"), "\n")
cat("  Unassigned (MC not in heatmap):       ", sum(broad_label == "Unassigned"), "\n")
cat("  Undefined (weak score):               ", sum(broad_label == "Undefined"), "\n")
print(table(broad_label, useNA = "ifany"))
```

```{r}
saveRDS(pbmc, file.path(rds_dir, paste0(FILE_PREFIX, "_MC_annotated.RDS")))
```


<!-- Optional. Rerun every chunk below after relabelling blocks, so the
     workbooks and the annotated RDS pick up the new calls. -->


# regenerate the annotation heatmap after relabelling in 04

```{r}
# Optional: apply hand curation from 04 and redraw. Requires the annotation heatmap chunk to have run.
# Writes block_anno_final; block_anno is left as the automatic call so it is never silently overwritten.
# Keep in sync with BLOCK_RELABEL in 04.
# 7 -> T cells: M28/M23 naive-CM CD4 program, scores ~0 from CD3 dropout in resting cells
# 5 -> Granulocytes: M5 neutrophil (CD177, ALPL, CYP4F3, ADGRG3), no neutrophil genes in the mast-only sig
# 1 -> MK/Ery: M24 MK (P2RY12, RGS18, NFE2, MEIS1) + M32/M27 ery (HEMGN, CA2, RHAG), mast trio not specific vs MK
BLOCK_RELABEL <- c()   # e.g. c("7" = "T cells", "5" = "Granulocytes")

stopifnot(exists("block_anno"), all(names(BLOCK_RELABEL) %in% names(block_anno)))
block_anno_final <- block_anno
block_anno_final[names(BLOCK_RELABEL)] <- BLOCK_RELABEL

cat("blocks relabelled:", sum(block_anno_final != block_anno), "\n")
print(data.frame(block = names(BLOCK_RELABEL),
                 auto  = block_anno[names(BLOCK_RELABEL)],
                 final = block_anno_final[names(BLOCK_RELABEL)], row.names = NULL))

# any label introduced by the relabel that the panel does not define, kept before Undefined
new_lin   <- setdiff(unique(block_anno_final), lin_order)
lin_order <- c(names(panel), new_lin, "Undefined")

anno <- factor(block_anno_final[as.character(block)], levels = lin_order)
ord  <- order(anno, block); ord <- ord[!is.na(anno[ord])]
anno <- droplevels(anno)   # drops Undefined if nothing is left in it

if (scale_rows) {
  mat <- t(scale(t(lognorm[markers, ord]))); mat[mat > 2.5] <- 2.5; mat[mat < -2.5] <- -2.5
} else {
  mat <- lognorm[markers, ord]
}

# keyed by name so new compartments cannot shift the existing colors
lineage_cols <- c(
  "T cells"      = "#8B1A1A",
  "NK"           = "#4E9E2F",
  "B cells"      = "#2C6FAD",
  "Plasma cells" = "#E69F00",
  "Granulocytes" = "#2D6A4F",
  "Mo/Macs"      = "#9E79C4",
  "moDCs"        = "#56B4E9",
  "pDCs"         = "#A0522D",
  "Platelet"     = "#CC79A7",
  "Cycling"      = "#8C8C8C",
  "MK/Ery"       = "#7F3F98",
  "Undefined"    = "#B2182B"
)
stopifnot(!anyNA(names(lineage_cols)), all(lin_order %in% names(lineage_cols)))
lineage_cols <- lineage_cols[lin_order]

top <- HeatmapAnnotation(compartment = anno[ord], col = list(compartment = lineage_cols),
                         show_annotation_name = FALSE)
ht2c <- Heatmap(mat, name = hm_name, col = col_fun, top_annotation = top,
        row_split = row_lin, column_split = anno[ord], column_title = NULL,
        cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = FALSE,
        row_names_gp = gpar(fontsize = 7), row_title_gp = gpar(fontsize = 8),
        row_title_rot = 0,
        column_gap = unit(0.5,"mm"), row_gap = unit(0.5,"mm"), use_raster = TRUE)
draw(ht2c)
```

```{r}
#Write cell type annotation to block mapping to same file
# uses the curated labels if the optional relabel chunk ran, automatic otherwise
block_anno_use <- if (exists("block_anno_final")) block_anno_final else block_anno
label_source   <- if (exists("block_anno_final")) "curated" else "automatic"
cat("block labels in use:", label_source, "\n")

celltype_block_map <- data.frame(
  block = rownames(z),
  assigned_celltype = block_anno_use[rownames(z)],
  best_score = apply(z, 1, max),
  second_score = apply(z, 1, function(x) sort(x, decreasing = TRUE)[2]),
  score_margin = apply(z, 1, function(x) diff(sort(x, decreasing = TRUE)[1:2])),
  top_module = apply(blockmeans[, rownames(z), drop = FALSE], 2, function(x) names(which.max(x))),
  label_source = label_source,
  stringsAsFactors = FALSE
)

out_xlsx <- file.path(csv_dir, paste0(FILE_PREFIX, "_module_block_map.xlsx"))

openxlsx::write.xlsx(
  list(
    modules_genes = modules_genes,
    block_module_ranking = block_rank,
    block_celltype_mapping = celltype_block_map
  ),
  file = out_xlsx,
  rowNames = FALSE
)
```

```{r}
#Add block and cluster label info to seurat obj
# uses the curated labels if the optional relabel chunk ran, automatic otherwise
block_anno_use <- if (exists("block_anno_final")) block_anno_final else block_anno
label_source   <- if (exists("block_anno_final")) "curated" else "automatic"
cat("block labels in use:", label_source, "\n")
stopifnot(identical(celltype_block_map$label_source[1], label_source))

cdata_cell <- anndata::read_h5ad(CELLS_H5AD)
counts_cell <- Matrix::t(cdata_cell$X)
counts_cell <- as(counts_cell, "CsparseMatrix")   # dgRMatrix -> dgCMatrix
colnames(counts_cell) <- rownames(cdata_cell$obs)
rownames(counts_cell) <- rownames(cdata_cell$var)

# X must be raw UMIs, not normalized
stopifnot(inherits(counts_cell, "dgCMatrix"))
stopifnot(nrow(counts_cell) == nrow(cdata_cell$var),
          ncol(counts_cell) == nrow(cdata_cell$obs))
stopifnot(all(counts_cell@x == floor(counts_cell@x)))

pbmc <- CreateSeuratObject(counts = counts_cell, meta.data = as.data.frame(cdata_cell$obs))
stopifnot(identical(colnames(pbmc), rownames(cdata_cell$obs)))

# only name block if cutree did not already
if (is.null(names(block))) {
  stopifnot(length(block) == ncol(thmz))
  names(block) <- colnames(thmz)
} else {
  stopifnot(setequal(names(block), colnames(thmz)))
}

mc_name    <- as.character(pbmc$metacell_name)
mc_int     <- as.integer(as.character(pbmc$metacell))
mc_outlier <- !is.na(mc_int) & mc_int < 0

mc_block <- as.character(unname(block[mc_name]))
mc_block[mc_outlier]  <- "Outlier"
mc_block[is.na(mc_block)] <- "Unassigned"

broad_label <- unname(block_anno_use[mc_block])
broad_label[mc_block %in% c("Outlier", "Unassigned")] <- mc_block[mc_block %in% c("Outlier", "Unassigned")]
broad_label[is.na(broad_label)] <- "Unassigned"

blk_idx <- match(mc_block, as.character(celltype_block_map$block))

pbmc <- AddMetaData(pbmc, data.frame(
  mc_name         = mc_name,
  mc_block        = mc_block,
  mc_outlier      = mc_outlier,
  broad_label     = broad_label,
  label_source    = label_source,
  mc_anno_score   = celltype_block_map$best_score[blk_idx],
  mc_anno_margin  = celltype_block_map$score_margin[blk_idx],
  mc_top_module   = celltype_block_map$top_module[blk_idx],
  row.names       = colnames(pbmc),
  stringsAsFactors = FALSE
))

mc_lookup <- data.frame(
  metacell_name = names(block),
  mc_block      = unname(block),
  broad_label   = unname(block_anno_use[as.character(unname(block))]),
  label_source  = label_source,
  stringsAsFactors = FALSE
)
write.csv(mc_lookup, file.path(csv_dir, paste0(FILE_PREFIX, "_metacell_block_annotations.csv")), row.names = FALSE)

cat("Metacells in module heatmap (thmz):     ", length(block), "\n")
cat("Metacells assigned to >=1 cell:         ", length(unique(mc_name[!mc_outlier])), "\n")
cat("Assigned, block-labelled metacells:     ", length(unique(mc_name[!mc_outlier & !mc_block %in% c("Outlier","Unassigned")])), "\n")
cat("Cells total:                            ", ncol(pbmc), "\n")
cat("  in metacell:                          ", sum(!mc_outlier), "\n")
cat("  MC2 outliers:                         ", sum(broad_label == "Outlier"), "\n")
cat("  Unassigned (MC not in heatmap):       ", sum(broad_label == "Unassigned"), "\n")
cat("  Undefined (weak score):               ", sum(broad_label == "Undefined"), "\n")
print(table(broad_label, useNA = "ifany"))
```


```{r}
saveRDS(pbmc, file.path(rds_dir, paste0(FILE_PREFIX, "_MC_annotated.RDS")))
```


<!-- # Block-wise dot plot, all modules  -->

<!-- ```{r} -->
<!-- library(reshape2); library(ggplot2) -->
<!-- drivers <- unique(as.vector(apply(blockmeans, 2, function(c) names(sort(c, decreasing=TRUE))[1:5]))) -->
<!-- mean_z <- t(sapply(drivers, function(m) tapply(thmz[m,], block, mean))) -->
<!-- frac_on <- t(sapply(drivers, function(m) tapply(thmz[m,] > 0, block, mean))) -->
<!-- df <- melt(mean_z, value.name="meanz"); df$frac <- melt(frac_on)$value -->
<!-- df$Var2 <- factor(df$Var2) -->
<!-- ggplot(df, aes(Var2, Var1, size=frac, color=meanz)) +  -->
<!--   geom_point() +  -->
<!--   scale_color_gradient2(low="navy", mid="white", high="firebrick") +  -->
<!--   labs(x="metacell block", y="module") +  -->
<!--   theme_bw() +  -->
<!--   theme(axis.text.x=element_text(angle=90, vjust=0.5, size=6), axis.text.y=element_text(size=5)) -->
<!-- ``` -->


<!-- # Modules grouped together based on block heatmap -->

<!-- ```{r} -->
<!-- # 1. assign each module to the single block where its mean z is highest -->
<!-- mod_block <- apply(blockmeans, 1, function(r) which.max(r))   # module -> its dominant block -->
<!-- # 2. one row per block = mean z across that block's modules, evaluated over ALL metacells -->
<!-- grp_rows <- t(sapply(sort(unique(mod_block)), function(b){ -->
<!--   mods <- names(mod_block)[mod_block == b] -->
<!--   if (length(mods) == 1) thmz[mods, ] else colMeans(thmz[mods, , drop=FALSE]) -->
<!-- })) -->
<!-- rownames(grp_rows) <- paste0("blk", sort(unique(mod_block)), "_mods") -->
<!-- # 3. plot: module-group rows x all metacells, columns split by block -->
<!-- ord <- order(block) -->
<!-- anno <- factor(block_anno[as.character(block)], levels = lin_order) -->
<!-- ord <- order(anno, block) -->
<!-- grp_rows <- t(scale(t(grp_rows))) -->
<!-- grp_rows[grp_rows > 2.5] <- 2.5 -->
<!-- grp_rows[grp_rows < -2.5] <- -2.5 -->

<!-- ComplexHeatmap::Heatmap(grp_rows[, ord], cluster_columns=FALSE, cluster_rows=FALSE, -->
<!--                         show_column_names=FALSE, column_split=block[ord], -->
<!--                         col=circlize::colorRamp2(c(-2,0,2), c("#3B4CC0","#F7F7F7","#B40426")), name="mean z", -->
<!--                         row_names_gp=grid::gpar(fontsize=6), column_title_gp=grid::gpar(fontsize=5), -->
<!--                         column_gap=grid::unit(0.3,"mm")) -->
<!-- ``` -->


<!-- # UMAP of metacells from Jupyter metacell coordinates - UMAP based on find_marker_genes and has nothing to do with module gene sets -->

```{r}
# umap <- read.csv(file.path(csv_dir, paste0(FILE_PREFIX, "_metacell_umap_coords.csv")), row.names=1)
# umap$block <- factor(block[rownames(umap)])
# cent <- aggregate(cbind(x, y) ~ block, data = umap, FUN = median)
# ggplot(umap, aes(x, y, color = block)) +
#   geom_point(size = 1.2) +
#   geom_text_repel(data = cent, aes(label = block), color = "black", size = 4, seed = 42, box.padding = 0.3, max.overlaps = Inf) +
#   theme_void() + theme(legend.position = "none")
# 
# ## SCpubr UMAP
# umap <- read.csv(file.path(csv_dir, paste0(FILE_PREFIX, "_metacell_umap_coords.csv")), row.names=1)
# total <- CreateSeuratObject(dgcmtxcounts)
# umap <- umap[colnames(total), ]                      
# total$block <- factor(block[colnames(total)])        
# emb <- as.matrix(umap[, c("x","y")]); colnames(emb) <- c("mcUMAP_1","mcUMAP_2")
# total[["mcumap"]] <- CreateDimReducObject(embeddings = emb, key = "mcUMAP_", assay = DefaultAssay(total))
# do_DimPlot(total, 
#            reduction = "mcumap", 
#            group.by = "block", 
#            label = TRUE, 
#            repel = TRUE, 
#            legend.position = "right")
#SCpubr with metacell celltype annotations
#block_labels <- c("1"="Baso?", "24"="CD4 Tem", "25"="CD8 eff")   # manual calls
#total$celltype <- factor(block_labels[as.character(total$block)])
#do_DimPlot(total, reduction = "mcumap", group.by = "celltype", label = TRUE, repel = TRUE)
```


# Stacked bar plot, cell level

Quick look only. The reportable version, with donor weighting and the
`complete()` zero-fill, is step 06. See docs/09_composition.md.

```{r}
library(dplyr)
# ---- 1. per-cell metadata + metacell assignment ----
cdata <- read_h5ad(CELLS_H5AD)
cell  <- cdata$obs
print(colnames(cell))

cell <- cell[cell$metacell >= 0, ]
cell$mc_name <- as.character(cell$metacell_name)   # now every row maps to a real metacell

# ---- 2. attach block annotation (block_anno + block from the heatmap step) ----
names(block) <- colnames(thmz)
cell$block      <- block[cell$mc_name]
cell$broad_label <- block_anno[as.character(cell$block)]
cell <- cell[!is.na(cell$broad_label), ]

# ---- 3. % per label within each sample, then averaged per group ----
per_sample <- cell %>%
  dplyr::count(sample_origin, broad_label) %>%
  group_by(sample_origin) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

sample_map <- cell %>% dplyr::select(sample_origin, condition, category, category_plot) %>% distinct()
per_sample <- per_sample %>% left_join(sample_map, by = "sample_origin")

summary_avg <- per_sample %>%
  group_by(category_plot, broad_label) %>%
  summarise(mean_pct = mean(pct), .groups = "drop")

# ---- 4. plot, palette keyed by label name ----
labs_all <- sort(unique(summary_avg$broad_label))
colour_palette <- setNames(
  c("#2C6FAD", "#5DC8C8", "#8B7420", "#E3B505", "#4E9E2F", "#ADD8E6", "#9E79C4","#8B1A1A","#8C8C8C")[seq_along(labs_all)],
  labs_all)

ggplot(summary_avg, aes(category_plot, mean_pct, fill = broad_label)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = colour_palette) +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "% of total cells", x = NULL, fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid.major.x = element_blank())
```
