---
title: "06 Composition: stacked bar plots"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    code_folding: hide
---

```{r, results=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(scales)
library(openxlsx)
```

# Configuration

```{r, results = FALSE, message=FALSE, warning=FALSE}
# LEVEL selects which annotated object this notebook runs on. Nothing else in
# the file is level-specific, so a subset run is a one-word edit here.
#   "global"  the whole dataset, annotated by broad lineage in 03
#   "subset"  one lineage, annotated by state in 05
LEVEL <- "subset"

source(file.path("..", "config", "config.R"))

CFG <- switch(LEVEL,
  global = list(slug = "global", prefix = "global",      label = "broad_label",
                title = "All cell types"),
  subset = list(slug = "subset", prefix = "subset_data", label = "cluster_label",
                title = "Subset"),
  stop("LEVEL must be 'global' or 'subset'"))

SUBSET_SLUG  <- CFG$slug
FILE_PREFIX  <- CFG$prefix
LABEL_COL    <- CFG$label     # per-cell annotation column written by 03 or 05
SUBSET_TITLE <- CFG$title     # appears in every figure title

# the annotated Seurat object written at the end of 03 (global) or 05 (subset)
rds_path <- file.path(DIR_RDS, SUBSET_SLUG, "metacell",
                      paste0(FILE_PREFIX, "_MC_annotated.RDS"))

# output trees for this step, created on demand
csv_dir <- out_dirs(SUBSET_SLUG, "csv")
fig_dir <- out_dirs(SUBSET_SLUG, "pdf")
```

# Read RDS

The metacell is the annotation device, not the counting unit. Every cell carries the
label of its metacell's block, and the frequencies below are counted over cells.
Counting metacells instead would be wrong twice over: MC2 sizes metacells to a target
cell count so their number is not proportional to abundance, and a metacell can pool
cells from several samples so it has no single sample of origin.

```{r}
obj <- readRDS(rds_path)
stopifnot(LABEL_COL %in% colnames(obj@meta.data))

cat("Cells total:                            ", ncol(obj), "\n")
print(table(obj@meta.data[[LABEL_COL]], useNA = "ifany"))
cat("Cells dropped (no cell type call):      ",
    sum(obj@meta.data[[LABEL_COL]] %in% drop_labels), "\n")
```

# QC on the dropped cells

Dropping unlabelled cells is only harmless if the drop rate is flat across samples and
conditions. If one condition loses many more cells than another, every proportion below
is shifted by that alone, so this table is checked before trusting any of the plots.

```{r}
drop_qc <- obj@meta.data %>%
  mutate(dropped = .data[[LABEL_COL]] %in% drop_labels) %>%
  group_by(sample_origin, condition, timepoint) %>%
  summarise(n_cells     = n(),
            n_dropped   = sum(dropped),
            pct_dropped = 100 * mean(dropped),
            .groups = "drop") %>%
  arrange(desc(pct_dropped))

print(as.data.frame(drop_qc))

drop_qc_cond <- drop_qc %>%
  group_by(condition) %>%
  summarise(mean_pct_dropped = mean(pct_dropped),
            min_pct_dropped  = min(pct_dropped),
            max_pct_dropped  = max(pct_dropped),
            .groups = "drop")
print(as.data.frame(drop_qc_cond))
```

# Per-sample composition table

`complete()` is what makes the donor means honest: a label absent from a sample is a
real zero for that sample, not a missing row. Without it the mean for a rare label is
taken over only the donors that happen to carry it, the per-group percentages no longer
sum to 100, and every stacked bar is silently renormalised.

```{r}
meta <- obj@meta.data %>%
  mutate(cell_label = as.character(.data[[LABEL_COL]])) %>%
  filter(!cell_label %in% drop_labels)

sample_map <- meta %>%
  select(sample_origin, patient_id, condition, timepoint, category, category_plot) %>%
  distinct()
stopifnot(!anyDuplicated(sample_map$sample_origin))

per_sample_counts <- meta %>%
  dplyr::count(sample_origin, cell_label, name = "n") %>%
  tidyr::complete(sample_origin, cell_label, fill = list(n = 0)) %>%
  group_by(sample_origin) %>%
  mutate(total_cells_sample = sum(n),
         pct_within_sample  = n / total_cells_sample * 100) %>%
  ungroup() %>%
  left_join(sample_map, by = "sample_origin") %>%
  mutate(timepoint = factor(as.character(timepoint), levels = TP_LEVELS),
         condition = factor(as.character(condition), levels = COND_LEVELS))

# every sample must close to 100 percent, otherwise the stacked bars are meaningless
stopifnot(all(abs(
  tapply(per_sample_counts$pct_within_sample, per_sample_counts$sample_origin, sum) - 100
) < 1e-8))
stopifnot(!anyNA(per_sample_counts$condition))

write.xlsx(as.data.frame(per_sample_counts),
           file.path(csv_dir, paste0(FILE_PREFIX, "_per_sample_composition.xlsx")),
           rowNames = FALSE)
```

```{r}
base_palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00",
                  "#CC79A7", "#999999", "#6A3D9A", "#B2182B", "#78A850", "#4D8FAC",
                  "#A0522D", "#C4A0C8", "#2D6A4F", "#C8A882", "#E07B8A", "#00B4C8",
                  "#8B1A1A", "#9AB8D4", "#D4A017", "#7B7B3C", "#5C4A72", "#E64B35",
                  "#8491B4", "#FAD4C0", "#ff7f0e", "#00468B", "#c49c94")

labs_all <- sort(unique(per_sample_counts$cell_label))
label_cols <- setNames(
  if (length(labs_all) <= length(base_palette)) base_palette[seq_along(labs_all)]
  else grDevices::hcl.colors(length(labs_all), "Dark 3"),
  labs_all)
stopifnot(!anyNA(names(label_cols)), !anyNA(label_cols))
```

# Stacked bars

Bars are the mean across donors of each donor's own percentage, so every donor carries
equal weight regardless of how many cells were captured. Bars are stacked, not filled:
after `complete()` the means already close to 100, and `position = "fill"` would hide it
if they ever stopped doing so.

```{r}
summarise_by <- function(group_col) {
  per_sample_counts %>%
    mutate(grp = .data[[group_col]]) %>%
    group_by(grp, cell_label) %>%
    summarise(mean_pct = mean(pct_within_sample),
              sd_pct   = sd(pct_within_sample),
              sem_pct  = sd(pct_within_sample) / sqrt(n()),
              n_donors = n(),
              .groups  = "drop")
}

stacked_bar <- function(group_col, y_lab, title_str, file_tag) {
  d <- summarise_by(group_col)
  # all groups must close to 100, this is the check that complete() did its job
  stopifnot(all(abs(tapply(d$mean_pct, d$grp, sum) - 100) < 1e-6))

  p <- ggplot(d, aes(x = grp, y = mean_pct, fill = cell_label)) +
    geom_col(position = "stack", width = 0.75) +
    scale_fill_manual(values = label_cols) +
    scale_y_continuous(labels = label_percent(scale = 1), expand = expansion(mult = c(0, 0.02))) +
    labs(y = y_lab, x = NULL, fill = NULL, title = title_str) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.major.x = element_blank(),
          plot.title = element_text(size = 10))

  ggsave(file.path(fig_dir, paste0(FILE_PREFIX, "_stacked_", file_tag, ".pdf")),
         p, width = 8, height = 5)
  print(p)
  invisible(d)
}
```

## By patient

```{r}
d_category <- stacked_bar("category",
  "Mean proportion across that patient's timepoints (%)",
  paste0(SUBSET_TITLE, ": composition by patient"), "by_patient")
```

## By condition

```{r}
d_condition <- stacked_bar("condition",
  "Mean proportion across that condition's samples (%)",
  paste0(SUBSET_TITLE, ": composition by condition"), "by_condition")
```

## By sample

```{r, fig.width = 12}
d_sample <- stacked_bar("category_plot",
  "Proportion per sample (%)",
  paste0(SUBSET_TITLE, ": composition by sample"), "by_sample")
```

## By timepoint within condition

```{r, fig.width = 10}
d_cond_tp <- per_sample_counts %>%
  mutate(cond_tp = factor(paste(condition, timepoint),
                          levels = paste(rep(COND_LEVELS, each = length(TP_LEVELS)), TP_LEVELS))) %>%
  filter(!is.na(cond_tp)) %>%
  group_by(cond_tp, cell_label) %>%
  summarise(mean_pct = mean(pct_within_sample), n_donors = n(), .groups = "drop") %>%
  mutate(cond_tp = droplevels(cond_tp))
stopifnot(all(abs(tapply(d_cond_tp$mean_pct, d_cond_tp$cond_tp, sum) - 100) < 1e-6))

p <- ggplot(d_cond_tp, aes(x = cond_tp, y = mean_pct, fill = cell_label)) +
  geom_col(position = "stack", width = 0.75) +
  scale_fill_manual(values = label_cols) +
  scale_y_continuous(labels = label_percent(scale = 1), expand = expansion(mult = c(0, 0.02))) +
  labs(y = "Mean proportion across donors (%)", x = NULL, fill = NULL,
       title = paste0(SUBSET_TITLE, ": composition by condition and timepoint")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 10))
ggsave(file.path(fig_dir, paste0(FILE_PREFIX, "_stacked_by_condition_timepoint.pdf")),
       p, width = 10, height = 5)
print(p)
```

# Dodged bars, label by condition

```{r, fig.width = 10}
d_dodge <- summarise_by("condition")

condition_colours <- c("CTRL" = "#F5C342", "CASE" = "#D4494E")

p <- ggplot(d_dodge, aes(x = cell_label, y = mean_pct, fill = grp)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean_pct, ymax = mean_pct + sem_pct),
                position = position_dodge(width = 0.8), width = 0.25) +
  scale_fill_manual(values = condition_colours) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(y = "% of a sample's cells (mean across samples, +SEM)", x = NULL, fill = NULL,
       title = paste0(SUBSET_TITLE, ": abundance by condition")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(size = 10))
ggsave(file.path(fig_dir, paste0(FILE_PREFIX, "_dodged_by_condition.pdf")), p,
       width = 10, height = 5)
print(p)
```

```{r}
wb <- createWorkbook()
addWorksheet(wb, "by_patient");   writeData(wb, "by_patient",   as.data.frame(d_category))
addWorksheet(wb, "by_condition"); writeData(wb, "by_condition", as.data.frame(d_condition))
addWorksheet(wb, "by_sample");    writeData(wb, "by_sample",    as.data.frame(d_sample))
addWorksheet(wb, "by_cond_tp");   writeData(wb, "by_cond_tp",   as.data.frame(d_cond_tp))
addWorksheet(wb, "drop_qc");      writeData(wb, "drop_qc",      as.data.frame(drop_qc))
saveWorkbook(wb, file.path(csv_dir, paste0(FILE_PREFIX, "_composition_summaries.xlsx")),
             overwrite = TRUE)
```
