---
title: "07 Composition: CLR models across timepoints"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    code_folding: show
---

```{r, results=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(lme4)
# lmerTest must be attached: without it anova() on an lmer fit has no p-value column and
# emmeans falls back to asymptotic z inference, which is anti-conservative at n = 3-5
library(lmerTest)
library(emmeans)
library(ggplot2)
library(compositions)
library(purrr)
library(forcats)
library(openxlsx)
```

# Information of Used R Libraries

R: `r R.Version()$version.string`,
emmeans: `r packageVersion('emmeans')`,
lme4: `r packageVersion('lme4')`,
lmerTest: `r packageVersion('lmerTest')`,
compositions: `r packageVersion('compositions')`,
tidyverse: `r packageVersion('tidyverse')`

# Method

Cell-type proportions are compositional: they are constrained to sum to 100 within a
sample, so a real increase in one population forces an apparent decrease in every other.
The centred log-ratio maps each sample from the simplex into unconstrained Euclidean
space by dividing every part by the geometric mean of all parts in that sample and
taking logs, after which ordinary linear models are valid. A CLR value is therefore
read as abundance relative to the average population in the same sample, not as an
absolute percentage.

Within a condition, timepoints are compared with `lmer(clr ~ timepoint + (1 | patient_id))`,
which respects the repeated sampling of the same patient. Between conditions, each
timepoint is a single sample per patient, so `lm(clr ~ condition)` is the correct
unpaired model.

# Configuration

```{r, results = FALSE, message=FALSE, warning=FALSE}
# LEVEL selects which annotated object this notebook runs on. Nothing else in
# the file is level-specific, so a subset run is a one-word edit here.
#   "global"  the whole dataset, annotated by broad lineage in 03
#   "subset"  one lineage, annotated by state in 05
LEVEL <- "subset"

source(file.path("..", "config", "config.R"))

CFG <- switch(LEVEL,
  global = list(slug = "global", prefix = "global",      label = "broad_label",
                title = "All cell types"),
  subset = list(slug = "subset", prefix = "subset_data", label = "cluster_label",
                title = "Subset"),
  stop("LEVEL must be 'global' or 'subset'"))

SUBSET_SLUG  <- CFG$slug
FILE_PREFIX  <- CFG$prefix
LABEL_COL    <- CFG$label     # per-cell annotation column written by 03 or 05
SUBSET_TITLE <- CFG$title     # appears in every figure title

# the annotated Seurat object written at the end of 03 (global) or 05 (subset)
rds_path <- file.path(DIR_RDS, SUBSET_SLUG, "metacell",
                      paste0(FILE_PREFIX, "_MC_annotated.RDS"))

# output trees for this step, created on demand
csv_dir <- out_dirs(SUBSET_SLUG, "csv")
fig_dir <- out_dirs(SUBSET_SLUG, "pdf")

# a timepoint needs at least this many patients in a condition before it enters any model
MIN_PATIENTS_PER_TP <- 2
```

# Read RDS and design availability

```{r}
obj <- readRDS(rds_path)
stopifnot(LABEL_COL %in% colnames(obj@meta.data))

cat("Cells total:                            ", ncol(obj), "\n")
print(table(obj@meta.data[[LABEL_COL]], useNA = "ifany"))
cat("Cells dropped (no cell type call):      ",
    sum(obj@meta.data[[LABEL_COL]] %in% drop_labels), "\n")

avail <- obj@meta.data %>%
  distinct(patient_id, condition, timepoint) %>%
  mutate(timepoint = factor(as.character(timepoint), levels = TP_LEVELS),
         condition = factor(as.character(condition), levels = COND_LEVELS)) %>%
  dplyr::count(condition, timepoint, .drop = FALSE) %>%
  tidyr::pivot_wider(names_from = timepoint, values_from = n, values_fill = 0)
print(as.data.frame(avail))

# long form of the same table, used to gate every contrast below
avail_long <- obj@meta.data %>%
  distinct(patient_id, condition, timepoint) %>%
  dplyr::count(condition, timepoint, name = "n_patients") %>%
  mutate(condition = as.character(condition), timepoint = as.character(timepoint))

usable_tp <- function(cond) {
  keep <- avail_long %>% filter(condition == cond, n_patients >= MIN_PATIENTS_PER_TP)
  intersect(TP_LEVELS, keep$timepoint)
}
for (cnd in COND_LEVELS)
  cat("Usable timepoints,", cnd, ":", paste(usable_tp(cnd), collapse = ", "), "\n")
```

# Per-sample composition

```{r, results=FALSE, message=FALSE, warning=FALSE}
meta <- obj@meta.data %>%
  mutate(cell_label = as.character(.data[[LABEL_COL]])) %>%
  filter(!cell_label %in% drop_labels,
         condition %in% COND_LEVELS) %>%
  as.data.frame()

sample_map <- meta %>%
  select(sample_origin, patient_id, condition, timepoint) %>%
  distinct()
stopifnot(!anyDuplicated(sample_map$sample_origin))

per_sample_counts <- meta %>%
  dplyr::count(sample_origin, cell_label, name = "n") %>%
  tidyr::complete(sample_origin, cell_label, fill = list(n = 0)) %>%
  group_by(sample_origin) %>%
  mutate(total_cells_sample = sum(n),
         pct_within_sample  = n / total_cells_sample * 100) %>%
  ungroup() %>%
  left_join(sample_map, by = "sample_origin") %>%
  mutate(timepoint = factor(as.character(timepoint), levels = TP_LEVELS),
         condition = factor(as.character(condition), levels = COND_LEVELS))

stopifnot(all(abs(
  tapply(per_sample_counts$pct_within_sample, per_sample_counts$sample_origin, sum) - 100
) < 1e-8))
```

```{r}
contrast_levels <- unlist(lapply(seq_along(TP_LEVELS), function(i)
  if (i < length(TP_LEVELS)) paste(TP_LEVELS[(i + 1):length(TP_LEVELS)], "-", TP_LEVELS[i])))

order_contrasts <- function(x)
  factor(as.character(x), levels = intersect(contrast_levels, unique(as.character(x))))
```

# CLR transform

A block absent from a sample gives a zero, and log(0) is -Inf, which would poison every
model that sample enters. Zeros are replaced multiplicatively with half the smallest
observed non-zero proportion in that same sample, which is the standard compositional
treatment, and the count is reported so the scale of the correction stays visible. The
row is then reclosed to 100; CLR is invariant to closure so this changes nothing, it
only keeps the input table readable.

```{r, results=FALSE, message=FALSE, warning=FALSE}
clr_df <- per_sample_counts %>%
  select(sample_origin, condition, timepoint, patient_id, cell_label, pct_within_sample) %>%
  tidyr::pivot_wider(names_from = cell_label, values_from = pct_within_sample, values_fill = 0)

clr_input <- as.matrix(clr_df %>% select(-sample_origin, -condition, -timepoint, -patient_id))
rownames(clr_input) <- clr_df$sample_origin

n_zero <- sum(clr_input == 0)
cat("Structural zeros in the sample x cluster table:", n_zero, "of", length(clr_input), "\n")

if (n_zero > 0) {
  for (i in seq_len(nrow(clr_input))) {
    row_i <- clr_input[i, ]
    if (any(row_i == 0)) {
      repl <- min(row_i[row_i > 0]) / 2
      row_i[row_i == 0] <- repl
      clr_input[i, ] <- row_i / sum(row_i) * 100
    }
  }
  cat("Zeros replaced with half the smallest non-zero proportion of their own sample\n")
}
stopifnot(all(clr_input > 0))

clr_mat <- clr(clr_input)

# as.data.frame() on a matrix runs make.names() on the columns, which would turn
# "CD4/CD8-amb Exhausted/Act" into "CD4.CD8.amb.Exhausted.Act" and silently break every
# join back to the label set. as_tibble with minimal repair keeps the names verbatim.
clr_tab <- as.matrix(clr_mat)
colnames(clr_tab) <- colnames(clr_input)
stopifnot(all(is.finite(clr_tab)))

clr_long <- tibble::as_tibble(clr_tab, .name_repair = "minimal") %>%
  mutate(sample_origin = clr_df$sample_origin,
         condition     = as.character(clr_df$condition),
         timepoint     = as.character(clr_df$timepoint),
         patient_id    = as.character(clr_df$patient_id)) %>%
  tidyr::pivot_longer(cols = -c(sample_origin, condition, timepoint, patient_id),
                      names_to = "cluster", values_to = "clr_value")

stopifnot(setequal(unique(clr_long$cluster), unique(per_sample_counts$cell_label)))
# CLR is centred: every sample's values sum to zero
stopifnot(all(abs(tapply(clr_long$clr_value, clr_long$sample_origin, sum)) < 1e-8))
```

# Within condition, across timepoints

```{r, results=FALSE, message=FALSE, warning=FALSE}
fit_within <- function(dat) {
  dat$timepoint <- droplevels(factor(dat$timepoint, levels = TP_LEVELS))
  use_mixed <- any(table(dat$patient_id) > 1) && dplyr::n_distinct(dat$patient_id) >= 3
  f <- if (use_mixed) lmerTest::lmer(clr_value ~ timepoint + (1 | patient_id), data = dat)
       else           lm(clr_value ~ timepoint, data = dat)
  list(fit = f, model = if (use_mixed) "lmer" else "lm")
}

anova_p_of <- function(fit, term) {
  tab <- suppressMessages(anova(fit))
  col <- intersect(c("Pr(>F)", "Pr(>Chisq)"), colnames(tab))
  if (!length(col) || !(term %in% rownames(tab))) return(NA_real_)
  tab[term, col[1]]
}

# timepoints with too few patients are dropped before fitting, not after, so they cannot
# inflate the residual df or produce contrasts that lean on a single sample
clr_within <- function(cond) {
  keep_tp <- usable_tp(cond)
  clr_long %>% filter(condition == cond, timepoint %in% keep_tp)
}

n_per_tp <- function(cond) {
  avail_long %>% filter(condition == cond) %>% select(timepoint, n_patients)
}

run_means_within <- function(cond) {
  clr_within(cond) %>%
    split(.$cluster) %>%
    imap_dfr(~{
      m  <- suppressMessages(fit_within(.x))
      em <- emmeans(m$fit, ~ timepoint, infer = TRUE)
      as.data.frame(em) %>%
        mutate(cluster   = .y,
               anova_p   = anova_p_of(m$fit, "timepoint"),
               model     = m$model,
               condition = cond)
    }) %>%
    left_join(n_per_tp(cond), by = "timepoint")
}

run_within <- function(cond) {
  np <- n_per_tp(cond)
  clr_within(cond) %>%
    split(.$cluster) %>%
    imap_dfr(~{
      m  <- suppressMessages(fit_within(.x))
      em <- emmeans(m$fit, ~ timepoint)
      as.data.frame(contrast(em, method = "revpairwise", adjust = "none", infer = TRUE)) %>%
        mutate(cluster = .y, model = m$model, condition = cond,
               anova_p = anova_p_of(m$fit, "timepoint"))
    }) %>%
    group_by(cluster) %>%
    mutate(p.adjust = p.adjust(p.value, method = "BH")) %>%   # BH over the pairs, per cluster
    ungroup() %>%
    mutate(tp_later   = sub(" - .*$", "", contrast),
           tp_earlier = sub("^.* - ", "", contrast)) %>%
    left_join(np %>% rename(tp_later = timepoint,   n_later   = n_patients), by = "tp_later") %>%
    left_join(np %>% rename(tp_earlier = timepoint, n_earlier = n_patients), by = "tp_earlier") %>%
    mutate(signif = case_when(p.adjust < 0.001 ~ "***",
                              p.adjust < 0.01  ~ "**",
                              p.adjust < 0.05  ~ "*",
                              TRUE ~ ""),
           est_label = sprintf("%+.2f%s", estimate, signif),
           contrast  = order_contrasts(contrast)) %>%
    arrange(contrast, cluster)
}

means_within_ctrl <- run_means_within("CTRL")
means_within_case   <- run_means_within("CASE")
within_ctrl <- run_within("CTRL")
within_case   <- run_within("CASE")
```

# Between conditions, per timepoint

```{r, results=FALSE, message=FALSE, warning=FALSE}
# a timepoint is only comparable if both conditions clear MIN_PATIENTS_PER_TP there
tp_both <- intersect(usable_tp("CTRL"), usable_tp("CASE"))
cat("Timepoints comparable between conditions:", paste(tp_both, collapse = ", "), "\n")

fit_between <- function(dat) {
  if (dplyr::n_distinct(dat$condition) < 2) return(NULL)
  f  <- lm(clr_value ~ condition, data = dat)
  em <- emmeans(f, ~ condition)
  as.data.frame(contrast(em, method = "revpairwise", adjust = "none", infer = TRUE)) %>%
    mutate(cluster   = as.character(dat$cluster)[1],
           timepoint = as.character(dat$timepoint)[1],
           n_CTRL    = sum(dat$condition == "CTRL"),
           n_CASE      = sum(dat$condition == "CASE"))
}

between_tp <- clr_long %>%
  filter(timepoint %in% tp_both) %>%
  split(list(.$cluster, .$timepoint), drop = TRUE) %>%
  map_dfr(fit_between) %>%
  group_by(cluster) %>%
  mutate(p.adjust = p.adjust(p.value, method = "BH")) %>%   # BH over timepoints, per cluster
  ungroup() %>%
  mutate(signif = case_when(p.adjust < 0.001 ~ "***",
                            p.adjust < 0.01  ~ "**",
                            p.adjust < 0.05  ~ "*",
                            TRUE ~ ""),
         est_label = sprintf("%+.2f%s", estimate, signif),
         timepoint = factor(timepoint, levels = TP_LEVELS)) %>%
  arrange(timepoint, cluster)
```

# Save

```{r, results=FALSE, message=FALSE, warning=FALSE}
write_block <- function(wb, sheet, title, df, start_row) {
  if (!(sheet %in% names(wb))) addWorksheet(wb, sheet)
  writeData(wb, sheet, title, startRow = start_row, startCol = 1)
  addStyle(wb, sheet, createStyle(textDecoration = "bold", fontSize = 12),
           rows = start_row, cols = 1)
  mergeCells(wb, sheet, cols = 1:ncol(df), rows = start_row)
  writeData(wb, sheet, as.data.frame(df), startRow = start_row + 1, startCol = 1,
            headerStyle = createStyle(textDecoration = "bold", border = "bottom"))
  setColWidths(wb, sheet, cols = 1:ncol(df), widths = "auto")
  start_row + nrow(df) + 4
}

within_ctrl_t0 <- within_ctrl %>% filter(grepl("- T0$", contrast))
within_case_t0   <- within_case   %>% filter(grepl("- T0$", contrast))

wb <- createWorkbook()
r <- write_block(wb, "within_CTRL", "Within CTRL: CLR timepoint comparison, T0 as reference", within_ctrl_t0, 1)
r <- write_block(wb, "within_CTRL", "Within CTRL: CLR timepoint comparison, all pairwise timepoints", within_ctrl, r)
r <- write_block(wb, "within_CASE",   "Within CASE: CLR timepoint comparison, T0 as reference", within_case_t0, 1)
r <- write_block(wb, "within_CASE",   "Within CASE: CLR timepoint comparison, all pairwise timepoints", within_case, r)
r <- write_block(wb, "CASE_vs_CTRL_by_tp", "CASE vs CTRL: CLR comparison at each timepoint", between_tp, 1)
r <- write_block(wb, "means", "Estimated marginal CLR means, within CTRL", means_within_ctrl, 1)
r <- write_block(wb, "means", "Estimated marginal CLR means, within CASE", means_within_case, r)
r <- write_block(wb, "availability", "Subjects per condition and timepoint", avail, 1)
saveWorkbook(wb, file.path(csv_dir, paste0(FILE_PREFIX, "_CLR_timepoint.xlsx")),
             overwrite = TRUE)
```

# Forest plots

```{r}
tp_cols <- c("T0" = "purple1", "T1" = "#4C78A8", "T2" = "#F58518",
             "T3" = "#54A24B", "T4" = "#D4494E")

forest_theme <- theme_minimal() +
  theme(strip.text        = element_text(size = 5, color = "black"),
        panel.spacing     = unit(0.4, "lines"),
        panel.spacing.y   = unit(0, "lines"),
        plot.margin       = margin(5, 5, 5, 5),
        axis.text.y       = element_text(size = 5, color = "black"),
        axis.text.x       = element_text(size = 5, angle = 45, hjust = 1, color = "black"),
        axis.title.x      = element_text(size = 6, color = "black"),
        axis.title.y      = element_text(size = 6, color = "black"),
        axis.ticks        = element_line(color = "black"),
        axis.ticks.length = unit(0.15, "cm"),
        plot.title        = element_text(size = 6),
        legend.position   = "bottom",
        legend.key.size   = unit(0.4, "cm"),
        legend.text       = element_text(size = 4),
        legend.title      = element_text(size = 4.5),
        panel.grid.major  = element_blank(),
        panel.grid.minor  = element_blank())

forest_base <- function(d, colour_var) {
  ggplot(d, aes(x = estimate, y = cluster, color = .data[[colour_var]])) +
    geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL),
                  position = position_dodge(width = 0.55), height = 0.2) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(position = position_dodge(width = 0.55), size = 2) +
    geom_text(aes(x = upper.CL + 0.25, label = est_label, group = .data[[colour_var]]),
              position = position_dodge(width = 0.55), vjust = 0.3, size = 1.8,
              color = "black", show.legend = FALSE) +
    scale_y_discrete(expand = expansion(mult = c(0.001, 0.001))) +
    scale_x_continuous(breaks = seq(-100, 100, by = 1)) +
    coord_cartesian(clip = "off")
}

order_clusters <- function(d)
  mutate(d, cluster = factor(as.character(cluster),
                             levels = rev(sort(unique(as.character(cluster))))))
```

## Relative to T0, within condition

```{r, fig.width = 14}
plots_t0 <- list()
for (cnd in COND_LEVELS) {
  d <- (if (cnd == "CTRL") within_ctrl_t0 else within_case_t0) %>%
    order_clusters() %>%
    mutate(timepoint = factor(sub(" - T0$", "", as.character(contrast)), levels = TP_LEVELS)) %>%
    filter(!is.na(timepoint))
  if (!nrow(d)) next
  p <- forest_base(d, "timepoint") +
    scale_color_manual(values = tp_cols) +
    labs(x = "CLR difference relative to T0", y = "Clusters", color = "Timepoint",
         title = paste0(SUBSET_TITLE, " (", cnd, "): All timepoint changes from T0")) +
    forest_theme + theme(axis.line = element_line(color = "black"))
  
  plots_t0[[cnd]] <- p
  print(p)
}
```

## All pairwise timepoints, within condition

```{r, fig.width = 14}
plots_pairwise <- list()
for (cnd in COND_LEVELS) {
  d <- (if (cnd == "CTRL") within_ctrl else within_case) %>% order_clusters()
  if (!nrow(d)) next
  p <- forest_base(d, "contrast") +
    labs(x = "CLR difference for later timepoint compared to earlier", y = "Clusters",
         title = paste0(SUBSET_TITLE, " (", cnd, "): all pairwise timepoint contrasts")) +
    facet_wrap(~ contrast, ncol = 3) +
    forest_theme +
    theme(legend.position = "none",
          panel.border = element_rect(color = "black", fill = NA, linewidth = 0.3),
          axis.line = element_blank())
  
  plots_pairwise[[cnd]] <- p
  print(p)
}
```

## CASE vs CTRL, per timepoint

```{r, fig.width = 14}
d <- between_tp %>% order_clusters() %>% mutate(timepoint = droplevels(timepoint))
p_between <- forest_base(d, "timepoint") +
  scale_color_manual(values = tp_cols) +
  labs(x = "CLR difference for CASE relative to CTRL", y = "Clusters",
       title = paste0(SUBSET_TITLE, ": CASE vs CTRL at each timepoint")) +
  facet_wrap(~ timepoint, nrow = 1) +
  forest_theme +
  theme(legend.position = "none",
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.3),
        axis.line = element_blank())

print(p_between)
```
