---
title: "08 Pseudobulk differential expression"
author: "Saumya Pothukuchi"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    code_folding: hide
---

```{r, include=FALSE}
options(width = 60)
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output, sprintf('style="max-height: %s;"', options$max.height))
    hook_output(x, options)
  })
})
```

```{r, results = FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(Seurat)
library(Matrix)
library(DESeq2)
library(edgeR)
library(limma)
library(variancePartition)
library(BiocParallel)
library(ashr)
library(ggplot2)
library(ggrepel)
library(openxlsx)
```

### Information of Used R Libraries

R: `r R.Version()$version.string`,
Seurat: `r packageVersion('Seurat')`,
DESeq2: `r packageVersion('DESeq2')`,
edgeR: `r packageVersion('edgeR')`,
limma: `r packageVersion('limma')`,
variancePartition: `r packageVersion('variancePartition')`,
ashr: `r packageVersion('ashr')`

# Why pseudobulk

The replicate is the patient, not the cell. Counts are summed to one profile per
replicate unit before any test is run, and the model is fitted on those units.

Which unit is the replicate depends on the contrast, and the four families below each
use a different one. The table of contrast against unit against design, and why testing
single cells directly does not control the false positive rate, are in
docs/10_differential_expression.md.

# Configuration

```{r, results = FALSE, message=FALSE, warning=FALSE}
# LEVEL selects which annotated object this notebook runs on. Nothing else in
# the file is level-specific, so a subset run is a one-word edit here.
#   "global"  the whole dataset, annotated by broad lineage in 03
#   "subset"  one lineage, annotated by state in 05
LEVEL <- "subset"

source(file.path("..", "config", "config.R"))

CFG <- switch(LEVEL,
  global = list(slug = "global", prefix = "global",      label = "broad_label",
                title = "All cell types"),
  subset = list(slug = "subset", prefix = "subset_data", label = "cluster_label",
                title = "Subset"),
  stop("LEVEL must be 'global' or 'subset'"))

SUBSET_SLUG  <- CFG$slug
FILE_PREFIX  <- CFG$prefix
LABEL_COL    <- CFG$label     # per-cell annotation column written by 03 or 05
SUBSET_TITLE <- CFG$title     # appears in every figure title

# the annotated Seurat object written at the end of 03 (global) or 05 (subset)
rds_path <- file.path(DIR_RDS, SUBSET_SLUG, "metacell",
                      paste0(FILE_PREFIX, "_MC_annotated.RDS"))

# One folder per contrast family. Keeping them apart means a family can be
# rerun, or dropped, without touching the others, and the aggregate workbook
# at the end just globs the folder.
deg_root <- file.path(out_dirs(SUBSET_SLUG, "csv"), "DEGs")
DIR_COND <- file.path(deg_root, "condition")                            # family 1
DIR_WITH <- file.path(deg_root, "within_condition_timepoint")           # family 2
DIR_BTWN <- file.path(deg_root, "between_condition_timepoint")          # family 3
DIR_INTX <- file.path(deg_root, "condition_by_timepoint_interaction")   # family 4
DIR_OMNI <- file.path(DIR_INTX, "omnibus")                              # unsigned F tests
fig_dir  <- file.path(out_dirs(SUBSET_SLUG, "pdf"), "DEGs")
for (d in c(DIR_COND, DIR_WITH, DIR_BTWN, DIR_INTX, DIR_OMNI, fig_dir))
  dir.create(d, recursive = TRUE, showWarnings = FALSE)

# set FALSE to run only the pooled ("all") analysis and skip the per-cluster loop
RUN_CLUSTER_WISE <- TRUE
# the interaction is fitted pooled first; flip to TRUE for the per-cluster pass
RUN_CLUSTER_WISE_INTERACTION <- FALSE

MIN_CELLS           <- 20   # a pseudobulk unit needs this many cells or it is dropped
MIN_UNITS_PER_GROUP <- 2    # DESeq2 cannot estimate within-group dispersion below this
MIN_COUNT           <- 10   # filterByExpr, per-unit count floor
MIN_TOTAL_COUNT     <- 15   # filterByExpr, across-unit count floor
MIN_GENES_TESTED    <- 200  # below this the fit is not worth reporting

# volcano thresholds, presentation only, they do not affect the tables
VOLC_P_COL   <- "p_val"
VOLC_P_CUT   <- 0.05
VOLC_LFC_CUT <- 0.25
```

# Load

```{r, results=FALSE, message=FALSE, warning=FALSE}
obj <- readRDS(rds_path)
obj <- JoinLayers(obj)
stopifnot(LABEL_COL %in% colnames(obj@meta.data))

cts_all <- GetAssayData(obj, assay = "RNA", layer = "counts")
md_all  <- obj@meta.data
md_all$cell_label <- as.character(md_all[[LABEL_COL]])
md_all$patient_id <- as.character(md_all$patient_id)
md_all$condition  <- factor(as.character(md_all$condition), levels = COND_LEVELS)
md_all$timepoint  <- factor(as.character(md_all$timepoint), levels = TP_LEVELS)
md_all$sample_origin <- as.character(md_all$sample_origin)

stopifnot(identical(colnames(cts_all), rownames(md_all)))
stopifnot(!anyNA(md_all$condition), !anyNA(md_all$timepoint))
# counts layer must be raw UMIs
stopifnot(all(cts_all@x == floor(cts_all@x)))

cells_base <- rownames(md_all)[!md_all$cell_label %in% drop_labels]
cat("Cells retained:", length(cells_base), "of", nrow(md_all), "\n")

# one sample per patient per timepoint is what makes the between-condition models unpaired
stopifnot(!anyDuplicated(unique(md_all[, c("sample_origin", "patient_id", "timepoint")])$sample_origin))
print(table(md_all$condition, md_all$timepoint))
```


# Helpers

```{r, results=FALSE, message=FALSE, warning=FALSE}
slugify <- function(x) gsub("^_+|_+$", "", gsub("[^A-Za-z0-9]+", "_", x))

units_to_run <- c("all",
  if (RUN_CLUSTER_WISE) sort(setdiff(unique(md_all$cell_label), drop_labels)) else NULL)

cells_of <- function(unit)
  if (unit == "all") cells_base else cells_base[md_all[cells_base, "cell_label"] == unit]

# Sum raw counts per level of group_col. Version-proof, no AggregateExpression ambiguity.
pb_sum <- function(cells, group_col, min_cells = MIN_CELLS) {
  if (!length(cells)) return(NULL)
  grp  <- factor(as.character(md_all[cells, group_col]))
  ok   <- names(which(table(grp) >= min_cells))
  keep <- grp %in% ok
  if (!any(keep)) return(NULL)
  cells <- cells[keep]; grp <- droplevels(grp[keep])
  # built directly rather than via sparse.model.matrix: a one-level factor has no
  # contrasts, and after the min-cell filter a rare cluster can legitimately leave one unit
  mm <- Matrix::sparseMatrix(i = seq_along(grp), j = as.integer(grp), x = 1,
                             dims = c(length(grp), nlevels(grp)),
                             dimnames = list(NULL, levels(grp)))
  round(as.matrix(cts_all[, cells, drop = FALSE] %*% mm))   # genes x units
}

# One row per pseudobulk unit. keep_cols must be constant inside a unit or the mapping
# is not well defined, so that is asserted rather than assumed.
pb_coldata <- function(cells, group_col, keep_cols, units) {
  md <- md_all[cells, c(group_col, keep_cols), drop = FALSE]
  g  <- as.character(md[[group_col]])
  for (k in keep_cols)
    stopifnot(all(tapply(as.character(md[[k]]), g, function(v) length(unique(v))) == 1))
  out <- md[!duplicated(g), keep_cols, drop = FALSE]
  rownames(out) <- g[!duplicated(g)]
  droplevels(out[units, , drop = FALSE])
}

MANIFEST <- list()

# DESeq2 on a pseudobulk matrix. Returns NULL when the design cannot support the
# contrast, so a skipped comparison is visible in the manifest instead of a bad table.
run_pb_deseq <- function(pb, coldata, design, contrast) {
  if (is.null(pb) || ncol(pb) < 2) return(NULL)
  coldata <- droplevels(coldata[colnames(pb), , drop = FALSE])
  if (anyNA(coldata)) return(NULL)

  grp <- coldata[[contrast[1]]]
  if (!all(contrast[2:3] %in% levels(grp))) return(NULL)
  if (min(sum(grp == contrast[2]), sum(grp == contrast[3])) < MIN_UNITS_PER_GROUP) return(NULL)

  mm <- model.matrix(design, coldata)
  if (qr(mm)$rank < ncol(mm)) return(NULL)          # singular or fully nested design
  if (nrow(mm) - ncol(mm) < 1)  return(NULL)        # no residual degrees of freedom

  keep <- edgeR::filterByExpr(pb, design = mm, lib.size = colSums(pb),
                              min.count = MIN_COUNT, min.total.count = MIN_TOTAL_COUNT)
  if (sum(keep) < MIN_GENES_TESTED) return(NULL)

  dds  <- DESeqDataSetFromMatrix(pb[keep, , drop = FALSE], colData = coldata, design = design)
  dds  <- DESeq(dds, quiet = TRUE)
  res  <- results(dds, contrast = contrast)
  resS <- lfcShrink(dds, contrast = contrast, type = "ashr", quiet = TRUE)

  data.frame(
    gene       = rownames(res),
    baseMean   = res$baseMean,
    avg_log2FC = resS$log2FoldChange,   # shrunken, for volcano
    lfcSE      = res$lfcSE,
    stat       = res$stat,              # unshrunken Wald stat, for GSEA ranking
    p_val      = res$pvalue,
    p_val_adj  = res$padj,
    row.names  = NULL
  ) %>% arrange(p_val)
}

# Single entry point: runs the contrast, writes the csv, logs the manifest row.
run_and_save <- function(pb, coldata, design, contrast, family, dir_out, unit, tag) {
  n_ref <- n_test <- NA_integer_
  if (!is.null(pb)) {
    cd <- droplevels(coldata[colnames(pb), , drop = FALSE])
    if (contrast[1] %in% colnames(cd)) {
      n_test <- sum(cd[[contrast[1]]] == contrast[2])
      n_ref  <- sum(cd[[contrast[1]]] == contrast[3])
    }
  }
  res <- tryCatch(run_pb_deseq(pb, coldata, design, contrast), error = function(e) NULL)

  if (is.null(res)) {
    MANIFEST[[length(MANIFEST) + 1]] <<- data.frame(
      family = family, unit = unit, contrast = tag, n_test = n_test, n_ref = n_ref,
      n_genes_tested = NA_integer_, n_padj05 = NA_integer_, status = "skipped")
    return(invisible(NULL))
  }

  f <- file.path(dir_out, paste0(FILE_PREFIX, "__", slugify(unit), "__", tag, ".csv"))
  write.csv(res, f, row.names = FALSE)
  MANIFEST[[length(MANIFEST) + 1]] <<- data.frame(
    family = family, unit = unit, contrast = tag, n_test = n_test, n_ref = n_ref,
    n_genes_tested = nrow(res), n_padj05 = sum(res$p_val_adj < 0.05, na.rm = TRUE),
    status = if (min(n_test, n_ref) < 3) "exploratory" else "ok")
  invisible(res)
}
```

# Family 1: CASE vs CTRL, patient level

Condition varies between patients, so the patient is the replicate and all of a patient's
timepoints are pooled into one profile. This is the headline contrast.

```{r, message=FALSE, warning=FALSE}
for (u in units_to_run) {
  cells <- cells_of(u)
  pb <- pb_sum(cells, "patient_id")
  if (is.null(pb)) next
  cd <- pb_coldata(cells, "patient_id", c("condition"), colnames(pb))
  cd$condition <- factor(as.character(cd$condition), levels = COND_LEVELS)
  run_and_save(pb, cd, ~ condition, c("condition", "CASE", "CTRL"),
               "condition", DIR_COND, u, "CASE_vs_CTRL")
}
```

# Family 2: timepoint contrasts within a condition, paired

Every pairwise timepoint contrast is run inside each condition. Only patients sampled at
**both** timepoints of a given pair are kept, so `~ patient_id + timepoint` is a genuine
paired design: a patient present at one timepoint only cannot inform the difference, and
leaving it in makes the design matrix rank deficient in exactly the cases that matter,
which is any timepoint carried by a single patient.

```{r, message=FALSE, warning=FALSE}
tp_pairs <- do.call(rbind, lapply(seq_len(length(TP_LEVELS) - 1), function(i)
  do.call(rbind, lapply((i + 1):length(TP_LEVELS), function(j)
    data.frame(earlier = TP_LEVELS[i], later = TP_LEVELS[j], stringsAsFactors = FALSE)))))

for (u in units_to_run) {
  cells_u <- cells_of(u)
  if (!length(cells_u)) next
  for (cnd in COND_LEVELS) {
    cells_c <- cells_u[md_all[cells_u, "condition"] == cnd]
    if (!length(cells_c)) next
    md_c <- md_all[cells_c, ]
    for (k in seq_len(nrow(tp_pairs))) {
      a <- tp_pairs$earlier[k]; b <- tp_pairs$later[k]
      pats <- intersect(unique(md_c$patient_id[md_c$timepoint == a]),
                        unique(md_c$patient_id[md_c$timepoint == b]))
      if (length(pats) < MIN_UNITS_PER_GROUP) next
      sel <- cells_c[md_c$timepoint %in% c(a, b) & md_c$patient_id %in% pats]
      pb  <- pb_sum(sel, "sample_origin")
      if (is.null(pb)) next
      cd <- pb_coldata(sel, "sample_origin", c("patient_id", "timepoint"), colnames(pb))
      cd$patient_id <- droplevels(factor(cd$patient_id))
      cd$timepoint  <- droplevels(factor(as.character(cd$timepoint), levels = TP_LEVELS))
      # a patient must still be complete after the min-cell filter or the pairing breaks
      complete <- names(which(table(cd$patient_id) == 2))
      cd <- cd[cd$patient_id %in% complete, , drop = FALSE]
      pb <- pb[, rownames(cd), drop = FALSE]
      if (ncol(pb) < 2 * MIN_UNITS_PER_GROUP) next
      cd$patient_id <- droplevels(cd$patient_id)
      run_and_save(pb, cd, ~ patient_id + timepoint, c("timepoint", b, a),
                   "within_condition_timepoint", DIR_WITH, u, paste0(cnd, "__", b, "_vs_", a))
    }
  }
}
```

# Family 3: CASE vs CTRL at each timepoint, unpaired

Each patient contributes one sample per timepoint, so there is nothing to block on and
`~ condition` is the correct design. Timepoints where either arm has fewer than
`MIN_UNITS_PER_GROUP` patients are skipped by the guard rather than fitted.

```{r, message=FALSE, warning=FALSE}
for (u in units_to_run) {
  cells_u <- cells_of(u)
  if (!length(cells_u)) next
  for (tp in TP_LEVELS) {
    sel <- cells_u[md_all[cells_u, "timepoint"] == tp]
    if (!length(sel)) next
    pb <- pb_sum(sel, "sample_origin")
    if (is.null(pb)) next
    cd <- pb_coldata(sel, "sample_origin", c("condition", "patient_id"), colnames(pb))
    cd$condition <- factor(as.character(cd$condition), levels = COND_LEVELS)
    run_and_save(pb, cd, ~ condition, c("condition", "CASE", "CTRL"),
                 "between_condition_timepoint", DIR_BTWN, u, paste0("CASE_vs_CTRL__at_", tp))
  }
}
```

# Family 4: condition x timepoint interaction

The first three families never ask whether one arm *changes differently* from the other.
The interaction term is the difference of those differences, which is the trajectory
question.

DESeq2 cannot fit it: patient is nested inside condition, so a patient's fixed effect
already encodes their condition and the design matrix is rank deficient. Patient enters
as a random intercept instead, which means a mixed model per gene.
`variancePartition::dream` does that, with `voomWithDreamWeights` for precision weights
and `eBayes` for variance moderation. Background: docs/10_differential_expression.md.

Two outputs. The **omnibus** test is a moderated F over all interaction coefficients
jointly, and lands in a subfolder because an F statistic is unsigned and cannot rank a
GSEA list. The **per-timepoint** tables give one signed coefficient each.

```{r, message=FALSE, warning=FALSE}
# a timepoint enters the interaction only if both arms clear MIN_UNITS_PER_GROUP there
pat_tp <- md_all %>%
  distinct(patient_id, condition, timepoint) %>%
  dplyr::count(condition, timepoint, name = "n_pat")
tp_intx <- intersect(TP_LEVELS, Reduce(intersect, lapply(COND_LEVELS, function(cnd)
  as.character(pat_tp$timepoint[pat_tp$condition == cnd & pat_tp$n_pat >= MIN_UNITS_PER_GROUP]))))
TP_REF <- if (length(tp_intx)) tp_intx[1] else NA_character_
cat("Timepoints in the interaction model:", paste(tp_intx, collapse = ", "),
    "| reference:", TP_REF, "\n")

run_interaction <- function(cells, unit) {
  log_skip <- function() {
    MANIFEST[[length(MANIFEST) + 1]] <<- data.frame(
      family = "condition_by_timepoint_interaction", unit = unit, contrast = "all_interaction",
      n_test = NA_integer_, n_ref = NA_integer_, n_genes_tested = NA_integer_,
      n_padj05 = NA_integer_, status = "skipped")
  }
  if (length(tp_intx) < 2) { log_skip(); return(invisible(NULL)) }

  sel <- cells[md_all[cells, "timepoint"] %in% tp_intx]
  pb  <- pb_sum(sel, "sample_origin")
  if (is.null(pb)) { log_skip(); return(invisible(NULL)) }

  cd <- pb_coldata(sel, "sample_origin", c("patient_id", "condition", "timepoint"), colnames(pb))
  cd$condition  <- factor(as.character(cd$condition), levels = COND_LEVELS)
  cd$timepoint  <- factor(as.character(cd$timepoint), levels = tp_intx)
  cd$patient_id <- droplevels(factor(cd$patient_id))

  # the min-cell filter can empty a condition x timepoint cell, so recheck after it
  ok <- all(table(cd$condition, cd$timepoint) >= MIN_UNITS_PER_GROUP)
  if (!ok || nlevels(cd$patient_id) < 3) { log_skip(); return(invisible(NULL)) }

  mm <- model.matrix(~ condition * timepoint, cd)
  if (qr(mm)$rank < ncol(mm) || nrow(mm) - ncol(mm) < 1) { log_skip(); return(invisible(NULL)) }

  dge  <- edgeR::DGEList(pb[, rownames(cd), drop = FALSE])
  keep <- edgeR::filterByExpr(dge, design = mm, min.count = MIN_COUNT,
                              min.total.count = MIN_TOTAL_COUNT)
  if (sum(keep) < MIN_GENES_TESTED) { log_skip(); return(invisible(NULL)) }
  dge <- edgeR::calcNormFactors(dge[keep, , keep.lib.sizes = FALSE])

  form <- ~ condition * timepoint + (1 | patient_id)
  fit <- tryCatch({
    vobj <- variancePartition::voomWithDreamWeights(dge, form, cd, quiet = TRUE,
                                                    BPPARAM = BiocParallel::SerialParam())
    variancePartition::eBayes(
      variancePartition::dream(vobj, form, cd, BPPARAM = BiocParallel::SerialParam()))
  }, error = function(e) NULL)
  if (is.null(fit)) { log_skip(); return(invisible(NULL)) }

  intx <- grep(":timepoint", colnames(coef(fit)), value = TRUE)
  if (!length(intx)) { log_skip(); return(invisible(NULL)) }

  log_row <- function(contrast, n_gen, n_sig, status)
    MANIFEST[[length(MANIFEST) + 1]] <<- data.frame(
      family = "condition_by_timepoint_interaction", unit = unit, contrast = contrast,
      n_test = sum(cd$condition == "CASE"), n_ref = sum(cd$condition == "CTRL"),
      n_genes_tested = n_gen, n_padj05 = n_sig,
      status = if (min(table(cd$condition)) < 3) "exploratory" else status)

  # omnibus, moderated F across every interaction coefficient
  tt <- limma::topTable(fit, coef = intx, number = Inf, sort.by = "F")
  omni <- data.frame(gene = rownames(tt), AveExpr = tt$AveExpr, F_stat = tt$F,
                     p_val = tt$P.Value, p_val_adj = tt$adj.P.Val, row.names = NULL)
  write.csv(omni, file.path(DIR_OMNI,
    paste0(FILE_PREFIX, "__", slugify(unit), "__interaction_omnibus.csv")), row.names = FALSE)
  log_row(paste0("omnibus_vs_", TP_REF), nrow(omni),
          sum(omni$p_val_adj < 0.05, na.rm = TRUE), "ok")

  # one signed table per interaction coefficient, rankable for GSEA
  for (cf in intx) {
    tp <- sub("^.*:timepoint", "", cf)
    t2 <- limma::topTable(fit, coef = cf, number = Inf, sort.by = "P")
    out <- data.frame(gene = rownames(t2), AveExpr = t2$AveExpr,
                      avg_log2FC = t2$logFC, lfcSE = t2$logFC / t2$t, stat = t2$t,
                      p_val = t2$P.Value, p_val_adj = t2$adj.P.Val, row.names = NULL)
    write.csv(out, file.path(DIR_INTX, paste0(FILE_PREFIX, "__", slugify(unit),
      "__CASE_vs_CTRL__", tp, "_vs_", TP_REF, ".csv")), row.names = FALSE)
    log_row(paste0("CASE_vs_CTRL__", tp, "_vs_", TP_REF), nrow(out),
            sum(out$p_val_adj < 0.05, na.rm = TRUE), "ok")
  }
  invisible(NULL)
}

intx_units <- c("all", if (RUN_CLUSTER_WISE_INTERACTION) setdiff(units_to_run, "all") else NULL)
for (u in intx_units) {
  cells <- cells_of(u)
  if (length(cells)) run_interaction(cells, u)
}
```

# Run manifest

Every attempted contrast is listed with the number of pseudobulk units on each side, so a
comparison that was skipped is distinguishable from one that was never asked for, and a
result resting on two units per arm is labelled `exploratory` rather than read as though
it had the same standing as the headline contrast.

```{r, message=FALSE, warning=FALSE}
manifest <- bind_rows(MANIFEST)
write.xlsx(manifest, file.path(deg_root, paste0(FILE_PREFIX, "_DEG_run_manifest.xlsx")),
           rowNames = FALSE)
print(as.data.frame(manifest %>% filter(unit == "all")))
cat("\nContrasts run:", sum(manifest$status != "skipped"),
    " skipped:", sum(manifest$status == "skipped"), "\n")
```

# Biotype annotation

```{r, message=FALSE, warning=FALSE}
library(stringr); library(readr)

gencode_gtf <- file.path(PROJ, "..", "reference", "gencode.v32.annotation.gtf.gz")
deg_files <- list.files(c(DIR_COND, DIR_WITH, DIR_BTWN, DIR_INTX, DIR_OMNI),
                        pattern = "\\.csv$", full.names = TRUE)

if (file.exists(gencode_gtf)) {
  g <- rtracklayer::import(gencode_gtf)
  genes <- as.data.frame(S4Vectors::mcols(g)) %>%
    filter(type == "gene") %>%
    select(gene_id, gene_name, gene_type) %>%
    distinct(gene_name, .keep_all = TRUE)

  for (f in deg_files) {
    degs <- read_csv(f, show_col_types = FALSE) %>%
      left_join(genes, by = c("gene" = "gene_name")) %>%
      mutate(
        biotype_class = case_when(
          str_detect(gene_type, "pseudogene") ~ "pseudogene",
          str_detect(gene_type, regex("IG|TR", ignore_case = TRUE)) ~ "IG_TR_gene",
          str_detect(gene_type, "lncRNA") ~ "lncRNA",
          str_detect(gene_type, "uncharacterized|unknown") ~ "uncharacterized",
          str_detect(gene_type, "protein_coding") ~ "protein_coding",
          TRUE ~ "other"),
        remove_flag = biotype_class %in% c("pseudogene", "lncRNA", "IG_TR_gene", "uncharacterized"))
    write_csv(degs, f)   # annotate in place, GSEA reads the same files
  }
} else {
  warning("gencode gtf not found, DEG csvs left unannotated: ", gencode_gtf)
}
```

# Volcano plots

```{r, results='asis', message=FALSE, warning=FALSE}
volcanoplot <- function(de, title_str) {
  if ("remove_flag" %in% names(de))
    de <- de %>% filter(is.na(remove_flag) | !remove_flag)
  de <- de %>% filter(!grepl("^MT-", gene), !grepl("^RP[SL]", gene))
  pv <- de[[VOLC_P_COL]]
  de$diffexpressed <- "NO"
  de$diffexpressed[de$avg_log2FC < -VOLC_LFC_CUT & pv < VOLC_P_CUT] <- "DOWN"
  de$diffexpressed[de$avg_log2FC >  VOLC_LFC_CUT & pv < VOLC_P_CUT] <- "UP"
  c1 <- de %>% filter(diffexpressed == "UP")   %>% arrange(desc(avg_log2FC)) %>% head(10)
  c2 <- de %>% filter(diffexpressed == "DOWN") %>% arrange(avg_log2FC)       %>% head(10)
  de$delabel <- ifelse(de$gene %in% unique(c(c1$gene, c2$gene)), de$gene, NA)
  ggplot(de, aes(avg_log2FC, -log10(.data[[VOLC_P_COL]]), col = diffexpressed, label = delabel)) +
    geom_vline(xintercept = c(-VOLC_LFC_CUT, VOLC_LFC_CUT), col = "gray", linetype = "dashed") +
    geom_hline(yintercept = -log10(VOLC_P_CUT), col = "gray", linetype = "dashed") +
    geom_point(alpha = 0.5, size = 0.8) +
    geom_text_repel(size = 2, max.overlaps = Inf, show.legend = FALSE) +
    scale_color_manual(values = c("UP" = "#bb0c00", "DOWN" = "#00AFBB", "NO" = "gray")) +
    labs(x = "shrunken avg log2FC", y = paste0("-log10 ", VOLC_P_COL), title = title_str) +
    theme_bw() + theme(plot.title = element_text(hjust = 0.5, size = 9))
}

pdf(file.path(fig_dir, paste0(FILE_PREFIX, "_volcanoes.pdf")), width = 6, height = 5)
for (f in deg_files) {
  tag <- sub("\\.csv$", "", basename(f))
  de  <- read.csv(f)
  if (!nrow(de) || !("avg_log2FC" %in% names(de))) next   # omnibus F tables are unsigned
  cat("##", tag, "\n")
  p <- volcanoplot(de, paste0(SUBSET_TITLE, ": ", tag))
  print(p)
  cat("\n\n")
}
invisible(dev.off())
```

# Aggregate workbook

```{r, message=FALSE, warning=FALSE}
for (fam in list(c("condition", DIR_COND), c("within_condition_timepoint", DIR_WITH),
                 c("between_condition_timepoint", DIR_BTWN),
                 c("condition_by_timepoint_interaction", DIR_INTX),
                 c("condition_by_timepoint_interaction_omnibus", DIR_OMNI))) {
  fs <- list.files(fam[2], pattern = "\\.csv$", full.names = TRUE)
  if (!length(fs)) next
  sheets <- lapply(fs, read.csv)
  nm <- sub(paste0("^", FILE_PREFIX, "__"), "", sub("\\.csv$", "", basename(fs)))
  
  # Shorten to 27 chars first, then make unique so '_1', '_2' fit under the 31-char limit
  nm_short <- substr(nm, 1, 27)
  names(sheets) <- make.unique(nm_short, sep = "_")
  
  write.xlsx(sheets, file.path(deg_root,
    paste0(FILE_PREFIX, "_pseudobulk_DEGs_", fam[1], ".xlsx")), rowNames = FALSE)
}
```

# Volcano plots and aggregate workbooks

```{r, message=FALSE, warning=FALSE}
FAMS <- list(
  c("condition",                                   DIR_COND),
  c("within_condition_timepoint",                  DIR_WITH),
  c("between_condition_timepoint",                 DIR_BTWN),
  c("condition_by_timepoint_interaction",          DIR_INTX),
  c("condition_by_timepoint_interaction_omnibus",  DIR_OMNI))

# The filename already encodes unit and contrast. Lifting them into columns keeps the
# sheet name down to the cluster, which is the only part that has to be unique per sheet.
annot_cols <- function(f, fam) {
  p <- strsplit(sub(paste0("^", FILE_PREFIX, "__"), "",
                    sub("\\.csv$", "", basename(f))), "__", fixed = TRUE)[[1]]
  out <- switch(fam,
    condition                   = data.frame(comparison = "CASE_vs_CTRL",
                                             timepoint  = "pooled", reference = NA_character_),
    within_condition_timepoint  = data.frame(comparison = p[2],
                                             timepoint  = sub("_vs_.*$", "", p[3]),
                                             reference  = sub("^.*_vs_", "", p[3])),
    between_condition_timepoint = data.frame(comparison = "CASE_vs_CTRL",
                                             timepoint  = sub("^at_", "", p[3]),
                                             reference  = NA_character_),
    condition_by_timepoint_interaction =
                                  data.frame(comparison = "CASE_vs_CTRL",
                                             timepoint  = sub("_vs_.*$", "", p[3]),
                                             reference  = sub("^.*_vs_", "", p[3])),
    condition_by_timepoint_interaction_omnibus =
                                  data.frame(comparison = "CASE_vs_CTRL",
                                             timepoint  = "omnibus", reference = NA_character_))
  cbind(cluster = if (p[1] == "all") "overall" else p[1], out)
}

load_family <- function(fam, dir_out) {
  fs <- list.files(dir_out, pattern = "\\.csv$", full.names = TRUE)
  if (!length(fs)) return(NULL)
  d <- bind_rows(lapply(fs, function(f) cbind(annot_cols(f, fam), read.csv(f))))
  d$timepoint <- factor(d$timepoint, levels = intersect(
    c(TP_LEVELS, "pooled", "omnibus"), unique(d$timepoint)))
  d %>% arrange(cluster != "overall", cluster, comparison, timepoint)
}
```

```{r, results='asis', message=FALSE, warning=FALSE}
volcanoplot <- function(de, title_str) {
  if ("remove_flag" %in% names(de))
    de <- de %>% filter(is.na(remove_flag) | !remove_flag)
  de <- de %>% filter(!grepl("^MT-", gene), !grepl("^RP[SL]", gene))
  pv <- de[[VOLC_P_COL]]
  de$diffexpressed <- "NO"
  de$diffexpressed[de$avg_log2FC < -VOLC_LFC_CUT & pv < VOLC_P_CUT] <- "DOWN"
  de$diffexpressed[de$avg_log2FC >  VOLC_LFC_CUT & pv < VOLC_P_CUT] <- "UP"
  c1 <- de %>% filter(diffexpressed == "UP")   %>% arrange(desc(avg_log2FC)) %>% head(10)
  c2 <- de %>% filter(diffexpressed == "DOWN") %>% arrange(avg_log2FC)       %>% head(10)
  de$delabel <- ifelse(de$gene %in% unique(c(c1$gene, c2$gene)), de$gene, NA)
  ggplot(de, aes(avg_log2FC, -log10(.data[[VOLC_P_COL]]), col = diffexpressed, label = delabel)) +
    geom_vline(xintercept = c(-VOLC_LFC_CUT, VOLC_LFC_CUT), col = "gray", linetype = "dashed") +
    geom_hline(yintercept = -log10(VOLC_P_CUT), col = "gray", linetype = "dashed") +
    geom_point(alpha = 0.5, size = 0.8) +
    geom_text_repel(size = 2, max.overlaps = Inf, show.legend = FALSE) +
    scale_color_manual(values = c("UP" = "#bb0c00", "DOWN" = "#00AFBB", "NO" = "gray")) +
    labs(x = "log2 fold change", y = paste0("-log10 ", VOLC_P_COL), title = title_str) +
    theme_bw() + theme(plot.title = element_text(hjust = 0.5, size = 9))
}

# One pdf per family, pages in the same cluster order as that family's workbook sheets
for (fam in FAMS) {
  d <- load_family(fam[1], fam[2])
  if (is.null(d) || !("avg_log2FC" %in% names(d))) next   # omnibus F tables are unsigned
  pdf(file.path(fig_dir, paste0(FILE_PREFIX, "_volcanoes_", fam[1], ".pdf")),
      width = 6, height = 5)
  for (cl in unique(d$cluster)) {
    for (cmp in unique(d$comparison[d$cluster == cl])) {
      sub_d <- d %>% filter(cluster == cl, comparison == cmp)
      for (tp in unique(sub_d$timepoint)) {
        de <- sub_d %>% filter(timepoint == tp)
        if (!nrow(de)) next
        ref <- de$reference[1]
        ttl <- paste0(cl, " | ", cmp,
                      if (tp != "pooled") paste0(" | ", tp) else "",
                      if (!is.na(ref)) paste0(" vs ", ref) else "")
        cat("##", ttl, "\n")
        print(volcanoplot(de, paste0(SUBSET_TITLE, ": ", ttl)))
        cat("\n\n")
      }
    }
  }
  invisible(dev.off())
}
```

```{r, message=FALSE, warning=FALSE}
# One workbook per family, one sheet per cluster. Contrast identity lives in the
# cluster / comparison / timepoint / reference columns, not in the sheet name.
for (fam in FAMS) {
  d <- load_family(fam[1], fam[2])
  if (is.null(d)) next
  sheets <- split(d, factor(d$cluster, levels = unique(d$cluster)))
  names(sheets) <- make.unique(substr(names(sheets), 1, 28), sep = "_")
  write.xlsx(sheets, file.path(deg_root,
    paste0(FILE_PREFIX, "_pseudobulk_DEGs_", fam[1], ".xlsx")), rowNames = FALSE)
}
```

