---
title: "09 Pseudobulk GSEA"
author: "Saumya Pothukuchi"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    code_folding: hide
---

```{r, include=FALSE}
options(width = 60)
local({
  hook_output <- knitr::knit_hooks$get('output')
  knitr::knit_hooks$set(output = function(x, options) {
    if (!is.null(options$max.height)) options$attr.output <- c(
      options$attr.output, sprintf('style="max-height: %s;"', options$max.height))
    hook_output(x, options)
  })
})
```

```{r, results = FALSE, message = FALSE, warning=FALSE}
library(openxlsx)
library(clusterProfiler)
library(msigdbr)
library(org.Hs.eg.db)
library(tidyverse)
library(fgsea)
library(msigdbdf)
```

### Information of Used R Libraries

R: `r R.Version()$version.string`,
clusterProfiler: `r packageVersion('clusterProfiler')`,
msigdbr: `r packageVersion('msigdbr')`,
fgsea: `r packageVersion('fgsea')`

# Method

Genes are ranked by the DESeq2 Wald statistic (`stat`), or by the dream moderated t for the
interaction family, rather than by fold change. The statistic is the effect divided by its
standard error, so a large but noisy fold change in a low-count gene does not outrank a
moderate, well-estimated one. The shrunken `avg_log2FC` in the same tables is for the
volcano plots, where the question is effect size, not evidence. Ranking a pseudobulk
contrast by log2FC alone puts low-expression genes at both ends of the list and the
enrichment follows them there.

The omnibus interaction tables carry `F_stat` and no signed statistic, so they cannot be
ranked. They are scanned anyway and logged as `not rankable` so that every DEG csv is
accounted for in the manifest rather than silently absent.

# Configuration

Mirrors the DEG configuration exactly: same family names, same directory layout, same
`FILE_PREFIX__unit__contrast` filename grammar. One input csv gives one output xlsx with
the same stem plus `__GSEA`, in the matching family folder.

```{r, results = FALSE, message=FALSE, warning=FALSE}
# LEVEL selects which annotated object this notebook runs on. Nothing else in
# the file is level-specific, so a subset run is a one-word edit here.
#   "global"  the whole dataset, annotated by broad lineage in 03
#   "subset"  one lineage, annotated by state in 05
LEVEL <- "subset"

source(file.path("..", "config", "config.R"))

CFG <- switch(LEVEL,
  global = list(slug = "global", prefix = "global",      label = "broad_label",
                title = "All cell types"),
  subset = list(slug = "subset", prefix = "subset_data", label = "cluster_label",
                title = "Subset"),
  stop("LEVEL must be 'global' or 'subset'"))

SUBSET_SLUG  <- CFG$slug
FILE_PREFIX  <- CFG$prefix
LABEL_COL    <- CFG$label     # per-cell annotation column written by 03 or 05
SUBSET_TITLE <- CFG$title     # appears in every figure title

# the annotated Seurat object written at the end of 03 (global) or 05 (subset)
rds_path <- file.path(DIR_RDS, SUBSET_SLUG, "metacell",
                      paste0(FILE_PREFIX, "_MC_annotated.RDS"))

# mirrors the 08 layout exactly, so one DEG csv maps to one GSEA xlsx
deg_root  <- file.path(out_dirs(SUBSET_SLUG, "csv"), "DEGs")
gsea_root <- file.path(out_dirs(SUBSET_SLUG, "csv"), "GSEA")
fig_dir   <- file.path(out_dirs(SUBSET_SLUG, "pdf"), "GSEA")

# family, subfolder, and whether the family can be ranked. The omnibus family is
# scanned for the manifest only: an unsigned F statistic cannot rank a GSEA list.
FAMS <- list(
  c("condition",                          "condition",                          "rank"),
  c("within_condition_timepoint",         "within_condition_timepoint",         "rank"),
  c("between_condition_timepoint",        "between_condition_timepoint",        "rank"),
  c("condition_by_timepoint_interaction", "condition_by_timepoint_interaction", "rank"),
  c("condition_by_timepoint_interaction_omnibus",
    file.path("condition_by_timepoint_interaction", "omnibus"),                 "skip"))

deg_dir_of  <- function(fam) file.path(deg_root,  fam[2])
gsea_dir_of <- function(fam) file.path(gsea_root, fam[2])

for (d in c(vapply(FAMS, gsea_dir_of, character(1)), fig_dir))
  dir.create(d, recursive = TRUE, showWarnings = FALSE)

RANK_COL  <- "stat"   # falls back to avg_log2FC if the column is absent
MIN_GENES <- 200      # a ranked list shorter than this is not worth testing
Q_CUT     <- 0.05     # significance cut for the summary plots only
TOP_N     <- 10

# the DEG csvs carry biotype flags; set TRUE to rank on protein-coding genes only
GSEA_DROP_FLAGGED <- FALSE
# mito and ribosomal genes are dropped from the volcanoes in 08, do the same here
GSEA_DROP_MT_RIBO <- TRUE
```

```{r, warning=FALSE, results = FALSE, message = FALSE}
set.seed(2024)
human_hallmark <- msigdbr(species = "Homo sapiens", collection = "H")
human_gobp     <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP")
human_kegg     <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:KEGG_LEGACY")
human_reactome <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:REACTOME")
human_gene_sets <- rbind(human_hallmark, human_gobp, human_kegg, human_reactome)
term2gene <- as.data.frame(human_gene_sets[, c("gs_name", "gene_symbol")])
cat("Gene sets loaded:", dplyr::n_distinct(term2gene$gs_name), "\n")
```

# Helpers

```{r, warning=FALSE, results = FALSE, message = FALSE}
# Same parser as the DEG notebook, so a GSEA row and its DEG row carry identical
# cluster / comparison / timepoint / reference values.
annot_cols <- function(f, fam) {
  p <- strsplit(sub(paste0("^", FILE_PREFIX, "__"), "",
                    sub("\\.csv$", "", basename(f))), "__", fixed = TRUE)[[1]]
  out <- switch(fam,
    condition                   = data.frame(comparison = "CASE_vs_CTRL",
                                             timepoint  = "pooled", reference = NA_character_),
    within_condition_timepoint  = data.frame(comparison = p[2],
                                             timepoint  = sub("_vs_.*$", "", p[3]),
                                             reference  = sub("^.*_vs_", "", p[3])),
    between_condition_timepoint = data.frame(comparison = "CASE_vs_CTRL",
                                             timepoint  = sub("^at_", "", p[3]),
                                             reference  = NA_character_),
    condition_by_timepoint_interaction =
                                  data.frame(comparison = "CASE_vs_CTRL",
                                             timepoint  = sub("_vs_.*$", "", p[3]),
                                             reference  = sub("^.*_vs_", "", p[3])),
    condition_by_timepoint_interaction_omnibus =
                                  data.frame(comparison = "CASE_vs_CTRL",
                                             timepoint  = "omnibus", reference = NA_character_))
  cbind(cluster = if (p[1] == "all") "overall" else p[1], out)
}

contrast_title <- function(a) {
  paste0(a$cluster, " | ", a$comparison,
         if (!a$timepoint %in% c("pooled")) paste0(" | ", a$timepoint) else "",
         if (!is.na(a$reference)) paste0(" vs ", a$reference) else "")
}

# fgsea indexes the ranked vector by name, so a duplicated symbol silently overwrites its
# twin and the set membership lookup goes wrong. Duplicates are resolved to the strongest
# statistic before ranking rather than left to chance.
build_ranks <- function(df) {
  rank_col <- if (RANK_COL %in% names(df)) RANK_COL else "avg_log2FC"
  if (!rank_col %in% names(df)) return(numeric(0))
  d <- df %>% filter(!is.na(gene), gene != "", is.finite(.data[[rank_col]]))
  if (GSEA_DROP_FLAGGED && "remove_flag" %in% names(d))
    d <- d %>% filter(is.na(remove_flag) | !remove_flag)
  if (GSEA_DROP_MT_RIBO)
    d <- d %>% filter(!grepl("^MT-", gene), !grepl("^RP[SL]", gene))
  d <- d %>%
    arrange(desc(abs(.data[[rank_col]]))) %>%
    distinct(gene, .keep_all = TRUE)
  v <- setNames(d[[rank_col]], d$gene)
  sort(v, decreasing = TRUE)
}

gsea_human <- function(v) {
  clusterProfiler::GSEA(
    geneList = v, TERM2GENE = term2gene,
    minGSSize = 15, maxGSSize = 500, pvalueCutoff = 1,
    eps = 0, seed = TRUE, verbose = FALSE)
}

# xlsx stem for a given DEG csv: identical stem, __GSEA suffix, matching family folder
out_path_for <- function(f, fam)
  file.path(gsea_dir_of(fam), sub("\\.csv$", "__GSEA.xlsx", basename(f)))
```

# Run

Every csv in every family folder is visited. A row is written to the manifest whether the
list was tested, too short, unrankable, or returned nothing, so the manifest is a complete
one to one map from DEG csv to GSEA xlsx.

```{r, message=FALSE, warning=FALSE}
gsea_log <- list()

log_row <- function(fam, f, a, n_ranked, n_sets, n_sig, status, out_f = NA_character_)
  gsea_log[[length(gsea_log) + 1]] <<- data.frame(
    family = fam[1], cluster = a$cluster, comparison = a$comparison,
    timepoint = a$timepoint, reference = a$reference,
    rank_col = NA_character_, n_ranked = n_ranked, n_sets = n_sets, n_sig = n_sig,
    status = status, source_csv = basename(f), output_xlsx = out_f,
    stringsAsFactors = FALSE)

for (fam in FAMS) {
  in_dir <- deg_dir_of(fam)
  if (!dir.exists(in_dir)) next
  # non-recursive: the omnibus subfolder is its own family and is scanned separately
  csv_files <- sort(list.files(in_dir, pattern = "\\.csv$", full.names = TRUE))

  for (csv_path in csv_files) {
    a  <- annot_cols(csv_path, fam[1])
    df <- read.csv(csv_path, stringsAsFactors = FALSE)

    if (fam[3] == "skip") {
      log_row(fam, csv_path, a, NA_integer_, NA_integer_, NA_integer_, "not rankable")
      next
    }
    if (!("gene" %in% names(df))) {
      log_row(fam, csv_path, a, NA_integer_, NA_integer_, NA_integer_, "no gene column")
      next
    }

    rank_col <- if (RANK_COL %in% names(df)) RANK_COL else "avg_log2FC"
    v <- build_ranks(df)
    if (length(v) < MIN_GENES) {
      r <- log_row(fam, csv_path, a, length(v), NA_integer_, NA_integer_, "skipped")
      gsea_log[[length(gsea_log)]]$rank_col <- rank_col
      next
    }

    gsea_res <- tryCatch(gsea_human(v), error = function(e) NULL)
    if (is.null(gsea_res) || !nrow(gsea_res@result)) {
      log_row(fam, csv_path, a, length(v), 0L, 0L, "no result")
      gsea_log[[length(gsea_log)]]$rank_col <- rank_col
      next
    }

    res_df <- cbind(a, family = fam[1], rank_col = rank_col,
                    source_csv = basename(csv_path), gsea_res@result, row.names = NULL)
    out_f <- out_path_for(csv_path, fam)
    write.xlsx(res_df, out_f, rowNames = FALSE)

    log_row(fam, csv_path, a, length(v), nrow(res_df),
            sum(res_df$qvalue < Q_CUT, na.rm = TRUE), "ok", basename(out_f))
    gsea_log[[length(gsea_log)]]$rank_col <- rank_col
  }
}

gsea_log <- bind_rows(gsea_log)
gsea_log$timepoint <- factor(gsea_log$timepoint, levels = intersect(
  c(TP_LEVELS, "pooled", "omnibus"), unique(gsea_log$timepoint)))
gsea_log <- gsea_log %>% arrange(family, cluster != "overall", cluster, comparison, timepoint)

write.xlsx(gsea_log, file.path(gsea_root, paste0(FILE_PREFIX, "_GSEA_run_manifest.xlsx")),
           rowNames = FALSE)
print(as.data.frame(gsea_log %>% filter(cluster == "overall")))
cat("\nLists tested:", sum(gsea_log$status == "ok"),
    " skipped:", sum(gsea_log$status != "ok"), "\n")
```

# Top pathways per comparison

One pdf per family, pages in the same cluster and contrast order as that family's
workbook sheets.

```{r, results='asis', message=FALSE, warning=FALSE}
plot_top_pathways <- function(res_df, tag, top_n = TOP_N) {
  sig <- res_df %>% filter(qvalue < Q_CUT)
  if (!nrow(sig)) return(NULL)
  top <- bind_rows(sig %>% arrange(desc(NES)) %>% head(top_n),
                   sig %>% arrange(NES)       %>% head(top_n)) %>%
    distinct(ID, .keep_all = TRUE) %>%
    mutate(dir = ifelse(NES > 0, "up", "down"), ID = str_trunc(ID, 60))
  ggplot(top, aes(reorder(ID, NES), NES, fill = dir)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = c("up" = "#bb0c00", "down" = "#00AFBB")) +
    labs(x = NULL, y = "NES", title = tag) +
    theme_bw() + theme(axis.text.y = element_text(size = 6),
                       plot.title = element_text(size = 8))
}

# Reads every xlsx of a family back and stacks it, with the annotation columns already
# written into each file, so ordering here needs no second parse of the filenames.
load_family <- function(fam) {
  fs <- list.files(gsea_dir_of(fam), pattern = "__GSEA\\.xlsx$", full.names = TRUE)
  if (!length(fs)) return(NULL)
  d <- bind_rows(lapply(fs, read.xlsx))
  d$timepoint <- factor(d$timepoint, levels = intersect(
    c(TP_LEVELS, "pooled", "omnibus"), unique(d$timepoint)))
  d %>% arrange(cluster != "overall", cluster, comparison, timepoint)
}

for (fam in FAMS) {
  if (fam[3] == "skip") next
  d <- load_family(fam)
  if (is.null(d)) next
  pdf(file.path(fig_dir, paste0(FILE_PREFIX, "_GSEA_top_pathways_", fam[1], ".pdf")),
      width = 9, height = 6)
  for (cl in unique(d$cluster)) {
    for (cmp in unique(d$comparison[d$cluster == cl])) {
      sub_d <- d %>% filter(cluster == cl, comparison == cmp)
      for (tp in unique(sub_d$timepoint)) {
        res <- sub_d %>% filter(timepoint == tp)
        if (!nrow(res)) next
        ttl <- contrast_title(res[1, c("cluster", "comparison", "timepoint", "reference")])
        p <- plot_top_pathways(res, paste0(SUBSET_TITLE, ": ", ttl))
        if (is.null(p)) next
        cat("##", ttl, "\n")
        print(p)
        cat("\n\n")
      }
    }
  }
  invisible(dev.off())
}
```

# Aggregate workbooks

One workbook per family, one sheet per cluster, matching the DEG workbooks. Contrast
identity lives in the cluster / comparison / timepoint / reference columns and the
`source_csv` column, not in the sheet name.

```{r, message=FALSE, warning=FALSE}
for (fam in FAMS) {
  if (fam[3] == "skip") next
  d <- load_family(fam)
  if (is.null(d)) next
  sheets <- split(d, factor(d$cluster, levels = unique(d$cluster)))
  names(sheets) <- make.unique(substr(names(sheets), 1, 28), sep = "_")
  write.xlsx(sheets, file.path(gsea_root,
    paste0(FILE_PREFIX, "_pseudobulk_GSEA_", fam[1], ".xlsx")), rowNames = FALSE)
}
```

```{r}
# Significance filter
# |NES| > 1.5 and FDR q < 0.05. Reads the finished GSEA xlsx back rather than the in-memory
# result, so this chunk is optional and re-runnable and never touches the unfiltered files.

NES_CUT <- 1.5
Q_SIG   <- 0.05

# tracking columns are pushed behind the clusterProfiler result so the enrichment is the
# first thing in view
META_COLS <- c("cluster", "comparison", "timepoint", "reference", "family",
               "rank_col", "source_csv")
meta_last <- function(d)
  d[, c(setdiff(names(d), META_COLS), intersect(META_COLS, names(d))), drop = FALSE]

# read.xlsx returns NULL for a header-only sheet and drops all-NA columns by default,
# both of which break bind_rows downstream; normalise here once
read_gsea <- function(f) {
  x <- tryCatch(read.xlsx(f, skipEmptyCols = FALSE), error = function(e) NULL)
  if (is.null(x) || !nrow(x)) return(NULL)
  for (k in c("cluster", "comparison", "timepoint", "reference", "family",
              "rank_col", "source_csv"))
    x[[k]] <- if (is.null(x[[k]])) NA_character_ else as.character(x[[k]])
  x
}

sig_log <- list()

for (fam in FAMS) {
  if (fam[3] == "skip") next
  fs <- list.files(gsea_dir_of(fam), pattern = "__GSEA\\.xlsx$", full.names = TRUE)
  if (!length(fs)) next

  for (f in fs) {
    d <- read_gsea(f)
    out_f <- sub("__GSEA\\.xlsx$", "__GSEA_filtered.xlsx", f)

    # identity comes from the filename, not from the file contents, so it is available
    # even when the source or the filtered result has no rows
    a <- annot_cols(sub("__GSEA\\.xlsx$", ".csv", f), fam[1])

    if (is.null(d)) {
      d_sig <- NULL
      n_sets <- 0L
    } else {
     d_sig  <- d %>% filter(!is.na(NES), !is.na(qvalue), abs(NES) > NES_CUT, qvalue < Q_SIG) %>%
      arrange(desc(NES)) %>% meta_last()
      n_sets <- nrow(d)
      # written even when empty, so every unfiltered file keeps a one to one partner
      write.xlsx(d_sig, out_f, rowNames = FALSE)
    }
    if (is.null(d)) write.xlsx(data.frame(), out_f, rowNames = FALSE)

    sig_log[[length(sig_log) + 1]] <- data.frame(
      family = fam[1], cluster = a$cluster, comparison = a$comparison,
      timepoint = a$timepoint, reference = as.character(a$reference),
      n_sets = n_sets,
      n_sig  = if (is.null(d_sig)) 0L else nrow(d_sig),
      n_up   = if (is.null(d_sig)) 0L else sum(d_sig$NES > 0),
      n_down = if (is.null(d_sig)) 0L else sum(d_sig$NES < 0),
      source_xlsx = basename(f), output_xlsx = basename(out_f),
      stringsAsFactors = FALSE)
  }
}

# aggregate workbooks, same split and sheet naming as the unfiltered pass
load_family_sig <- function(fam) {
  fs <- list.files(gsea_dir_of(fam), pattern = "__GSEA_filtered\\.xlsx$", full.names = TRUE)
  if (!length(fs)) return(NULL)
  ds <- lapply(fs, read_gsea)
  ds <- ds[!vapply(ds, is.null, logical(1))]
  if (!length(ds)) return(NULL)
  d <- bind_rows(ds)
  d$timepoint <- factor(d$timepoint, levels = intersect(
    c(TP_LEVELS, "pooled", "omnibus"), unique(d$timepoint)))
   d %>% arrange(cluster != "overall", cluster, comparison, timepoint, desc(NES)) %>% meta_last()
}

for (fam in FAMS) {
  if (fam[3] == "skip") next
  d <- load_family_sig(fam)
  if (is.null(d)) next
  sheets <- split(d, factor(d$cluster, levels = unique(d$cluster)))
  names(sheets) <- make.unique(substr(names(sheets), 1, 28), sep = "_")
  write.xlsx(sheets, file.path(gsea_root,
    paste0(FILE_PREFIX, "_pseudobulk_GSEA_", fam[1], "_filtered.xlsx")), rowNames = FALSE)
}

sig_log <- bind_rows(sig_log)
write.xlsx(sig_log, file.path(gsea_root,
  paste0(FILE_PREFIX, "_GSEA_significant_manifest.xlsx")), rowNames = FALSE)
print(as.data.frame(sig_log %>% filter(cluster == "overall")))
cat("\nFiles filtered:", nrow(sig_log),
    " with at least one significant set:", sum(sig_log$n_sig > 0), "\n")
```

