---
title: "10 Clonal repertoire (TCR)"
author: "Saumya Pothukuchi"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    code_folding: hide
---

<!--
Structure follows the rnabio.org TCR workflow. Attaches the repertoire and the
external flag, produces the descriptive tables and figures, writes one workbook
and one RDS. Diversity, tracking, Startrac and the test families are in 11,
which reads that RDS.

Figure conventions, applied throughout:
  - one palette, defined once in the config chunk
  - sample-wise axes rotated 45 degrees
  - sample-wise comparisons facetted by condition
  - clone size on a sequential ramp (ordinal), cell state on the categorical
    palette, condition on its own two colours

Several distribution figures are rebuilt from the scRepertoire export tables
rather than printed from scRepertoire directly, because the returned ggplot
carries no `condition` column and cannot be facetted without editing the
object. Same numbers, controllable figure.

Clone calling, bin scales and chain dropout: docs/12_clonal.md.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(width = 60)
```

```{r libs, results = FALSE}
library(dplyr)
library(tidyr)
library(Seurat)
library(scRepertoire)
library(ggplot2)
library(patchwork)
library(openxlsx)
library(ggridges)

stopifnot(packageVersion("scRepertoire") >= "2.0.0")
```

R: `r R.Version()$version.string`,
Seurat: `r packageVersion('Seurat')`,
scRepertoire: `r packageVersion('scRepertoire')`

# Configuration

```{r config, results = FALSE}
# Paths, factor levels and palettes come from config/config.R.
source(file.path("..", "config", "config.R"))

SUBSET_SLUG  <- "subset"
FILE_PREFIX  <- "subset_data"
STATE_COL    <- "mc_state"     # optional finer state column, if 05 wrote one
SUBSET_TITLE <- "Subset"
VDJ_KIND     <- "vdj_t"        # "vdj_b" for BCR, everything else is unchanged

# the annotated object written by 05
rds_path   <- file.path(DIR_RDS, SUBSET_SLUG, "metacell",
                        paste0(FILE_PREFIX, "_MC_annotated.RDS"))
clone_root <- file.path(out_dirs(SUBSET_SLUG, "csv"), "clonal")
fig_dir    <- file.path(out_dirs(SUBSET_SLUG, "pdf"), "clonal")
for (d in c(clone_root, fig_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

# "strict" is V/J gene plus the nucleotide CDR3 on both chains. The most
# conservative call scRepertoire offers, and the right default within a patient:
# two cells only share a clone if their receptors are identical at nucleotide level.
CLONE_CALL <- "strict"

# Count bins. Tuned to PBMC, where a 20-cell clone is already large. Prefer the
# occurrence bins below for anything compared across samples of different depth.
FREQ_BINS   <- c(0, 1, 2, 5, 20, Inf)
FREQ_LABELS <- c("count 1", "count 2", "count 3-5", "count 6-20", "count >20")

set.seed(123)

# Cell Ranger V(D)J tree and the library-to-sample map, both from config/
VDJ_RAW_DIR <- file.path(PROJ, "data", "vdj")
VDJ_BATCHES <- c("batch1", "batch2", "batch3", "batch4")

# library folder token -> sample id. Library naming is rarely derivable from the
# sample id by pattern, so the mapping is written out explicitly in a csv and
# read here rather than hardcoded.
vdj_map <- read.csv(file.path(PROJ, "config", "vdj_library_map.csv"),
                    stringsAsFactors = FALSE)
VDJ_TOKEN_TO_SAMPLE <- setNames(vdj_map$sample_origin, vdj_map$library_token)
```

## Palette and theme

One place to change every figure in both notebooks.

The clone-size ramp is taken from Mhanna et al., *Cell Rep Methods* 2024,
Fig 1C: cool blues for rare clonotypes, then cream, orange and red as occurrence
rises. It reads better than a single-hue ramp because the warm end is where the
biology is, and it is ordinal so it needs no legend lookup to interpret. Every
ordered clone-size scale in both notebooks uses it.

Condition colours follow the Nature Medicine house style in three of the four
papers: a steel blue and a brick red, used consistently and never swapped.

```{r palette, results = FALSE}
PAL_COND  <- c(CTRL = "#4C7FB8", CASE = "#A02C2C")   # steel blue / brick red
PAL_TP    <- c(T0 = "#D9D9D9", T1 = "#A6BDDB", T2 = "#67A9CF",
               T3 = "#3690C0", T4 = "#02648A")     # ordered, light to dark
PAL_FLAG   <- c(`FALSE` = "#C9CCD1", `TRUE` = "#A02C2C")

# Expanded / Unique, the >=2-cell dichotomy both Nature Medicine papers lead with
PAL_CLONAL <- c(Unique = "#4FBF6E", Expanded = "#3F4E8C")

# ordinal ramp for clone size: cool = rare, warm = expanded (Cell Rep Methods Fig 1C).
# The cool end is darker than Fig 1C's: at bar-slice size the near-white blues
# were invisible and read as empty legend keys.
PAL_SEQ <- c("#BBD7EA", "#7FB2D4", "#3D7FB0", "#FBD277", "#EE8B3C", "#B62B2B")

# categorical, for cell states. Muted, prints well, 12 before it recycles.
PAL_CAT <- c("#4C72B0", "#DD8452", "#55A868", "#C44E52", "#8172B3", "#937860",
             "#DA8BC3", "#8C8C8C", "#CCB974", "#64B5CD", "#4A6FE3", "#A0522D")

# occurrence breaks, verbatim from Cell Rep Methods Fig 1C. A clonotype is
# binned by the percentage of the repertoire it occupies, not by cell count, so
# samples of different depth are directly comparable.
OCC_BREAKS <- c(0, 1e-4, 1e-3, 1e-2, 1e-1, 1, 100)
OCC_LABELS <- c("(0,1e-04]", "(1e-04,0.001]", "(0.001,0.01]",
                "(0.01,0.1]", "(0.1,1]", "(1,100]")
PAL_OCC    <- setNames(PAL_SEQ, OCC_LABELS)

# map a sequential ramp onto however many levels a factor actually has.
# scRepertoire names cloneSize levels things like "Rare (0 < X <= 1e-04)", so
# they are matched positionally rather than by name.
# rev = TRUE when the factor's levels run largest-first, so the warm end of the
# ramp always lands on the most expanded level whichever way the levels are ordered.
seq_pal <- function(x, rev = FALSE) {
  lv <- if (is.factor(x)) levels(droplevels(x)) else sort(unique(x[!is.na(x)]))
  cols <- grDevices::colorRampPalette(PAL_SEQ)(length(lv))
  if (rev) cols <- rev(cols)
  setNames(cols, lv)
}
cat_pal <- function(x) {
  lv <- if (is.factor(x)) levels(droplevels(x)) else sort(unique(x[!is.na(x)]))
  setNames(rep(PAL_CAT, length.out = length(lv)), lv)
}

theme_tcr <- function(base_size = 11) {
  theme_bw(base_size) +
    theme(panel.grid.minor  = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.border      = element_rect(colour = "grey35", linewidth = 0.5),
          strip.background  = element_rect(fill = "grey92", colour = "grey35"),
          strip.text        = element_text(face = "bold", size = rel(0.85)),
          axis.text.x       = element_text(angle = 45, hjust = 1, vjust = 1),
          legend.key.size   = unit(0.4, "cm"),
          plot.title        = element_text(face = "bold", size = rel(1.0)),
          plot.subtitle     = element_text(colour = "grey30", size = rel(0.85)))
}

# CTRL and CASE in their own boxes, each only as wide as its own sample count
facet_cond <- function(...) facet_grid(..., scales = "free_x", space = "free_x")
```

# Load the annotated object

```{r load, results = FALSE}
obj <- readRDS(rds_path)
obj <- JoinLayers(obj)

stopifnot(LABEL_COL %in% colnames(obj@meta.data))
stopifnot(all(c("category", "category_plot") %in% colnames(obj@meta.data)))

obj$patient_id    <- as.character(obj$patient_id)
obj$sample_origin <- as.character(obj$sample_origin)
obj$condition     <- factor(as.character(obj$condition), levels = COND_LEVELS)
obj$timepoint     <- factor(as.character(obj$timepoint), levels = TP_LEVELS)
obj$cell_label    <- as.character(obj@meta.data[[LABEL_COL]])
if (STATE_COL %in% colnames(obj@meta.data))
  obj$cell_state <- as.character(obj@meta.data[[STATE_COL]])

CAT_LEVELS     <- levels(factor(obj$category))
CATPLOT_LEVELS <- levels(factor(obj$category_plot))
obj$category      <- factor(as.character(obj$category),      levels = CAT_LEVELS)
obj$category_plot <- factor(as.character(obj$category_plot),  levels = CATPLOT_LEVELS)

# StartracDiversity (stats notebook) reads Idents(). Left unset it reports
# majorCluster = "SeuratProject" and tran = 0 everywhere.
Idents(obj) <- factor(obj$cell_label)
stopifnot(!anyNA(obj$condition), !anyNA(obj$timepoint), !anyNA(obj$category_plot))
```

```{r load-report}
cat("Cells in object:", ncol(obj), "\n")
print(table(obj$condition, obj$timepoint))

sample_meta <- obj@meta.data %>%
  distinct(sample_origin, category_plot, category, patient_id, condition, timepoint) %>%
  arrange(category_plot)
stopifnot(!anyDuplicated(sample_meta$sample_origin))
sampleNames <- sample_meta$sample_origin
```

# Normalisation and UMAP

`DimPlot(obj, group.by = "cloneSize")` failed with *Unable to find a DimReduc
matching one of 'umap', 'tsne', or 'pca'*, and `clonalNetwork(reduction = "umap")`
was failing for the same reason and being swallowed by `safe()`.

There is no embedding anywhere in this project to carry over. MC2 ran directly on
the merged object, cells were subset on the global block annotations, and MC2 ran
again on the subset; the blocks come from ComplexHeatmap column k-means, so no
step ever needed one. So it is computed here, on this object.

Worth knowing when reading the panels: this embedding is independent of the
metacell blocks. `cluster_label` comes from the module heatmap, not from these
coordinates, so the labels will not necessarily form clean islands, and where
they disagree the heatmap is the annotation of record.

```{r normalise-umap}
obj <- NormalizeData(obj, verbose = FALSE)
obj <- FindVariableFeatures(obj, nfeatures = 2000, verbose = FALSE)
obj <- ScaleData(obj, verbose = FALSE)
obj <- RunPCA(obj, npcs = 30, verbose = FALSE)
ElbowPlot(obj, reduction = "pca", ndims = 50)

obj <- RunUMAP(obj, dims = 1:30, verbose = FALSE)
```

# Locate the contig files

```{r locate}
vdj_paths <- unlist(lapply(VDJ_BATCHES, function(b)
  Sys.glob(file.path(VDJ_RAW_DIR, b, "cellranger_output", "*", "per_sample_outs", "*",
                     VDJ_KIND, "filtered_contig_annotations.csv"))))

# token match, the same map 01 uses. Visit numbering is not derivable by pattern.
vdj_path_of <- list()
for (p in vdj_paths) {
  hits <- VDJ_TOKEN_TO_SAMPLE[vapply(names(VDJ_TOKEN_TO_SAMPLE),
                                     function(tk) grepl(tk, p, fixed = TRUE), logical(1))]
  if (length(hits) == 1) vdj_path_of[[unname(hits)]] <- c(vdj_path_of[[unname(hits)]], p)
}
if (length(names(vdj_path_of)[lengths(vdj_path_of) > 1]))
  stop("more than one ", VDJ_KIND, " file matched a sample token")

vdj_samples <- intersect(sampleNames, names(vdj_path_of))
missing_vdj <- setdiff(sampleNames, names(vdj_path_of))
cat("Files found:", length(vdj_paths), "| samples resolved:",
    length(vdj_samples), "of", length(sampleNames), "\n")
if (length(missing_vdj)) cat("WARNING no", VDJ_KIND, "for:",
                             paste(missing_vdj, collapse = ", "), "\n")
stopifnot(length(vdj_samples) > 0)

# the token map errors if one token matches two files, but not if a token
# matches the WRONG file. Printing the resolved table is the only way to catch
# that. batch feeds the public-clone check in the stats notebook.
path_map <- data.frame(
  sample_origin = vdj_samples,
  contig_path   = vapply(vdj_samples, function(s) vdj_path_of[[s]][1], character(1),
                         USE.NAMES = FALSE), stringsAsFactors = FALSE) %>%
  mutate(batch = sub("/.*$", "", sub(paste0("^", VDJ_RAW_DIR, "/"), "", contig_path))) %>%
  left_join(sample_meta, by = "sample_origin") %>%
  dplyr::select(category_plot, sample_origin, patient_id, condition, timepoint,
                batch, contig_path) %>%
  arrange(category_plot)
print(as.data.frame(path_map %>% dplyr::select(-contig_path)))

sample_meta <- sample_meta %>%
  left_join(path_map %>% dplyr::select(sample_origin, batch), by = "sample_origin")
obj$batch <- sample_meta$batch[match(obj$sample_origin, sample_meta$sample_origin)]
```

# Combine contigs

```{r combine}
# Not filtering contigs on productive/is_cell again: 01 preprocessing already
# did doublet filtering based on TCR-B.
contig_list <- lapply(vdj_samples, function(s)
  read.csv(vdj_path_of[[s]][1], stringsAsFactors = FALSE))
names(contig_list) <- vdj_samples

contig_list <- lapply(contig_list, function(df) {
  for (cl in c("productive", "is_cell", "high_confidence", "full_length"))
    if (cl %in% names(df)) df[[cl]] <- as.logical(toupper(df[[cl]]))
  df$sample <- NULL
  df
})

# filterMulti keeps the highest-UMI contig per chain rather than discarding the
# barcode, which is the right call after the 01 dual-TRB doublet filter.
# removeNA = FALSE keeps single-chain barcodes: they are still clones.
combined <- combineTCR(contig_list,
                       samples     = vdj_samples,
                       removeNA    = FALSE,
                       removeMulti = FALSE,
                       filterMulti = TRUE)

# combineTCR writes barcodes as SAMPLE_BARCODE. 01 wrote cell names as BARCODE_SAMPLE.
combined <- lapply(combined, function(d) {
  d$barcode <- sub("^(.*?)_(.*)$", "\\2_\\1", d$barcode); d
})
names(combined) <- vdj_samples
```

# Filter contigs to cells present in the object

From the rnabio workflow. Contig files contain barcodes that
never reached this object: cells that failed RNA QC, called doublets, and cells
that landed in another lineage subset. Left in, every repertoire summary
describes the raw contig file rather than the T cells being analysed, and the
clone-size denominators stop agreeing with the QC table.

Filtering here rather than only inside `combineExpression` gives one denominator
everywhere. `clonalFrequency` and `cloneSize` shift slightly against the
unfiltered run; that is the intended effect.

```{r filter-to-object}
in_obj <- split(colnames(obj), obj$sample_origin)

pre_n    <- vapply(combined, nrow, integer(1))
combined <- lapply(vdj_samples, function(s)
  combined[[s]][combined[[s]]$barcode %in% in_obj[[s]], , drop = FALSE])
names(combined) <- vdj_samples

# a sample losing every barcode means the reformat above is wrong, not that VDJ failed
keep <- vapply(combined, nrow, integer(1)) > 0
if (any(!keep)) cat("dropping samples with no matching barcode:",
                    paste(names(keep)[!keep], collapse = ", "), "\n")
combined    <- combined[keep]
vdj_samples <- names(combined)
stopifnot(length(vdj_samples) > 0)

post_n <- vapply(combined, nrow, integer(1))
cat(sprintf("Contig barcodes: %d before filter, %d after (%.1f%% retained)\n",
            sum(pre_n), sum(post_n), 100 * sum(post_n) / sum(pre_n)))
```

```{r vdj-qc}
## TCR matching is never 100%: 10x VDJ recovery is 75-90%, low-TCR-mRNA cells
## fail contig assembly, and NK / innate-like cells make no productive TCR reads.
vdj_qc <- bind_rows(lapply(vdj_samples, function(s) {
  d <- combined[[s]]
  data.frame(sample_origin     = s,
             n_cells_subset    = length(in_obj[[s]]),
             n_contigs_pre     = unname(pre_n[s]),
             n_cells_with_vdj  = nrow(d),
             vdj_recovery_pct  = round(100 * nrow(d) / max(length(in_obj[[s]]), 1), 2),
             n_paired          = sum(!is.na(d$TCR1) & !is.na(d$TCR2)),
             paired_pct_of_vdj = round(100 * mean(!is.na(d$TCR1) & !is.na(d$TCR2)), 2),
             stringsAsFactors  = FALSE)
})) %>%
  left_join(sample_meta, by = "sample_origin") %>%
  dplyr::select(category_plot, sample_origin, patient_id, condition, timepoint,
                batch, everything()) %>%
  arrange(category_plot)

stopifnot(all(vdj_qc$n_paired <= vdj_qc$n_cells_with_vdj))
print(as.data.frame(vdj_qc %>% dplyr::select(-sample_origin, -patient_id)))
```

```{r label-copy}
# scRepertoire labels its plots and export tables with the names of this list.
# combineExpression matches on barcode and is run on `combined`; everything the
# reader sees is run on this copy, so every figure comes out labelled CASE3_T0.
cat_of  <- setNames(as.character(sample_meta$category_plot), sample_meta$sample_origin)
orig_of <- setNames(sample_meta$sample_origin, as.character(sample_meta$category_plot))
combined_lab <- combined
names(combined_lab) <- unname(cat_of[names(combined)])
combined_lab <- combined_lab[order(factor(names(combined_lab), levels = CATPLOT_LEVELS))]
stopifnot(!anyNA(names(combined_lab)))

# scRepertoire export tables carry the list name in whichever column it chose.
# Find it and add sample_origin so the table still joins to everything else.
add_origin <- function(d) {
  d <- as.data.frame(d)
  if (!"category_plot" %in% names(d) && !is.null(rownames(d)) &&
      all(rownames(d) %in% names(orig_of))) d$category_plot <- rownames(d)
  if (!"category_plot" %in% names(d)) {
    hit <- names(d)[vapply(d, function(v) is.character(v) || is.factor(v), logical(1))]
    hit <- hit[vapply(hit, function(cn)
      mean(as.character(d[[cn]]) %in% names(orig_of)) > 0.9, logical(1))]
    if (length(hit)) names(d)[names(d) == hit[1]] <- "category_plot"
  }
  if ("category_plot" %in% names(d))
    d$sample_origin <- unname(orig_of[as.character(d$category_plot)])
  rownames(d) <- NULL
  d
}
```

# Attach the repertoire to the object

`group.by = "none"` computes clone frequency inside each sample, which is what a
timepoint-level expansion statement needs. Patient-level frequency is added
separately below.

Bins stay on the proportion scale. rnabio uses `proportion = FALSE` with counts
`Single=1, Small=5, Medium=20, Large=100, Hyperexpanded=500`; see the note at
the end of this file for why that does not transfer to PBMC.

```{r combine-expression}
obj <- combineExpression(combined, obj,
                         cloneCall  = CLONE_CALL,
                         chain      = "both",
                         group.by   = "none",
                         proportion = TRUE,
                         cloneSize  = c(Rare = 1e-4, Small = 0.001, Medium = 0.01,
                                        Large = 0.1, Hyperexpanded = 1),
                         filterNA   = FALSE)

obj$has_vdj <- !is.na(obj$CTstrict)
obj$has_tcr <- obj$has_vdj

# verify the group.by = "none" claim rather than trusting the comment
.chk <- obj@meta.data %>% filter(!is.na(CTstrict)) %>%
  dplyr::count(sample_origin, CTstrict, name = "n_in_sample") %>%
  left_join(obj@meta.data %>% filter(!is.na(CTstrict)) %>%
              distinct(sample_origin, CTstrict, clonalFrequency),
            by = c("sample_origin", "CTstrict"))
cat("clonalFrequency is per-sample:", all(.chk$n_in_sample == .chk$clonalFrequency), "\n")

obj$count_bin <- factor(as.character(cut(obj$clonalFrequency, breaks = FREQ_BINS,
                                         labels = FREQ_LABELS, right = TRUE)),
                        levels = FREQ_LABELS)
obj$count_bin[!obj$has_vdj] <- NA

# The >=2-cell dichotomy. Both Nature Medicine papers lead with this rather than
# with binned clone sizes, and it is the one clone-size statement that involves
# no arbitrary cutoff at all: a clone is either seen once or more than once.
obj$expanded      <- obj$has_vdj & obj$clonalFrequency > 1
obj$clonal_status <- factor(ifelse(!obj$has_vdj, NA_character_,
                                   ifelse(obj$expanded, "Expanded", "Unique")),
                            levels = c("Unique", "Expanded"))

# occurrence bin: the clone's share of its own sample's repertoire, as a percent.
# Cell Rep Methods Fig 1C. Depth-independent by construction.
obj$occ_pct <- 100 * obj$clonalProportion
obj$occ_bin <- cut(obj$occ_pct, breaks = OCC_BREAKS, labels = OCC_LABELS,
                   include.lowest = TRUE, right = TRUE)

# patient-level frequency, the denominator for anything tracking a clone across visits
md <- obj@meta.data
pat_freq <- md %>% filter(!is.na(CTstrict)) %>%
  dplyr::count(patient_id, CTstrict, name = "patient_clone_n") %>%
  group_by(patient_id) %>%
  mutate(patient_clone_prop = patient_clone_n / sum(patient_clone_n)) %>% ungroup()
idx <- match(paste(md$patient_id, md$CTstrict),
             paste(pat_freq$patient_id, pat_freq$CTstrict))
obj$patient_clone_n    <- pat_freq$patient_clone_n[idx]
obj$patient_clone_prop <- pat_freq$patient_clone_prop[idx]
obj$patient_count_bin  <- factor(as.character(cut(obj$patient_clone_n, breaks = FREQ_BINS,
                                                  labels = FREQ_LABELS, right = TRUE)),
                                 levels = FREQ_LABELS)
obj$patient_count_bin[!obj$has_vdj] <- NA

cat("Cells with a TCR:", sum(obj$has_vdj), "of", ncol(obj),
    sprintf("(%.1f%%)\n", 100 * mean(obj$has_vdj)))
print(table(obj$cloneSize, useNA = "ifany"))
print(table(obj$count_bin, useNA = "ifany"))
```

## Chain dropout, and a beta-only clone id

`CTstrict` concatenates both chains, so a cell that lost its alpha chain gets a
different string from its own sisters and is counted as a separate clone. That
inflates `n_clones`, deflates the largest clone sizes, biases `clonality`
downwards and under-detects persistence, at a rate that differs several-fold
between samples. Measured here and carried as `chain_status` / `clone_trb` so
the stats notebook can run the sensitivity analysis without re-reading contigs.
Orphans are not merged into their beta-matched parent: that would invent
lineage calls the data does not support.

```{r chain-dropout}
aa_parts  <- strsplit(ifelse(is.na(obj$CTaa), "NA_NA", obj$CTaa), "_", fixed = TRUE)
has_alpha <- vapply(aa_parts, function(v) length(v) >= 1 && !v[1] %in% c("NA", ""), logical(1))
has_beta  <- vapply(aa_parts, function(v) length(v) >= 2 && !v[2] %in% c("NA", ""), logical(1))
obj$chain_status <- ifelse(!obj$has_vdj, NA_character_,
                    ifelse(has_alpha & has_beta, "paired",
                    ifelse(has_beta, "beta_only", ifelse(has_alpha, "alpha_only", "unknown"))))

# beta-only clone id: TRBV/TRBJ plus the beta nucleotide CDR3
trb_map <- bind_rows(lapply(combined, function(d)
  data.frame(barcode = d$barcode,
             clone_trb = ifelse(is.na(d$TCR2) | is.na(d$cdr3_nt2), NA_character_,
                                paste(d$TCR2, d$cdr3_nt2, sep = "_")),
             stringsAsFactors = FALSE)))
obj$clone_trb <- trb_map$clone_trb[match(colnames(obj), trb_map$barcode)]

dropout_qc <- obj@meta.data %>%
  filter(has_vdj) %>%
  group_by(category_plot, sample_origin) %>%
  summarise(n_vdj = n(),
            pct_paired     = round(100 * mean(chain_status == "paired"), 2),
            pct_beta_only  = round(100 * mean(chain_status == "beta_only"), 2),
            pct_alpha_only = round(100 * mean(chain_status == "alpha_only"), 2),
            .groups = "drop") %>%
  left_join(sample_meta %>% dplyr::select(sample_origin, condition, timepoint, batch),
            by = "sample_origin") %>%
  arrange(desc(pct_beta_only))
print(as.data.frame(dropout_qc))
```

# external label mapping from the extcalls object

Moved here from the extcalls notebook and written into the RDS below, so
`flag_pos` travels with the object instead of being re-derived downstream.

```{r flag-transfer}
cols_to_transfer <- c(
  "flag_pos", "flag_call", "scGate_multi",
  "is.pure_A", "is.pure_B",
  "Tcell_UCell", "CD4Tcell_UCell", "CD8T_UCell",
  "Talphabeta_UCell", "Tgammadelta_UCell",
  "Tgd1_UCell", "Tgd2_UCell", "Treg_UCell",
  "scDblFinder_class", "scDblFinder_score",
  "KEEP_CELL")

extcalls_data <- readRDS(rds_path)
cols_present   <- intersect(cols_to_transfer, colnames(extcalls_data@meta.data))
extcalls_meta <- extcalls_data@meta.data[, cols_present, drop = FALSE]
rm(extcalls_data); invisible(gc())

# do not overwrite the QC columns already on obj
for (cl in c("scDblFinder_class", "scDblFinder_score"))
  if (cl %in% colnames(extcalls_meta))
    colnames(extcalls_meta)[colnames(extcalls_meta) == cl] <- paste0("extcalls_", cl)

# extcalls barcodes are SAMPLE_BARCODE, ours are BARCODE_SAMPLE
rownames(extcalls_meta) <- sub("^(.*?)_(.*)$", "\\2_\\1", rownames(extcalls_meta))

as_flag_logical <- function(x) {
  if (is.factor(x)) x <- as.character(x)
  if (is.character(x)) x <- x %in% c("TRUE", "True", "true", "1")
  x <- as.logical(x); x[is.na(x)] <- FALSE; x
}
extcalls_meta$flag_pos <- as_flag_logical(extcalls_meta$flag_pos)
```

```{r car-accounting-global}
n_cert     <- nrow(extcalls_meta)
flag_bc     <- rownames(extcalls_meta)[extcalls_meta$flag_pos]
n_flag_ext <- length(flag_bc)
cat(sprintf("Extcalls cells:                 %d\n", n_cert))
cat(sprintf("Flag+ in extcalls:               %d (%.2f%% of extcalls)\n",
            n_flag_ext, 100 * n_flag_ext / n_cert))

# optional: the global annotated object, used only to report where flagged
# cells landed outside this subset. Skipped with a warning if absent.
global_path <- file.path(DIR_RDS, "global", "metacell", "global_MC_annotated.RDS")
if (file.exists(global_path)) {
  global_md <- readRDS(global_path)@meta.data
  in_glob   <- intersect(flag_bc, rownames(global_md))
  cat(sprintf("Flag+ found in global MC object:  %d (%.2f%% of Flag+)\n",
              length(in_glob), 100 * length(in_glob) / n_flag_ext))
  flag_broad <- as.data.frame(table(broad_label = global_md[in_glob, "broad_label"]),
                             responseName = "n_flag") %>%
    filter(n_flag > 0) %>%
    mutate(pct_of_car = round(100 * n_flag / sum(n_flag), 2)) %>%
    arrange(desc(n_flag))
  cat("\nFlag+ cells by global broad_label:\n"); print(flag_broad)
  rm(global_md); invisible(gc())
} else {
  warning("global MC object not found, broad_label breakdown skipped: ", global_path)
}
```

```{r car-apply}
shared_barcodes <- intersect(rownames(extcalls_meta), colnames(obj))
cat(sprintf("\nBarcode transfer: %d of %d subset cells matched (%.2f%%)\n",
            length(shared_barcodes), ncol(obj),
            100 * length(shared_barcodes) / ncol(obj)))

obj <- AddMetaData(obj, metadata = extcalls_meta[shared_barcodes, , drop = FALSE])
obj$flag_pos <- as_flag_logical(obj$flag_pos)
stopifnot(is.logical(obj$flag_pos), !anyNA(obj$flag_pos))

n_flag_sub <- sum(obj$flag_pos)
cat(sprintf("\nFlag+ landing in subset:        %d (%.2f%% of all Flag+, %.2f%% of T cells)\n",
            n_flag_sub, 100 * n_flag_sub / n_flag_ext, 100 * n_flag_sub / ncol(obj)))
cat(sprintf("Flag+ lost outside subset:      %d\n", n_flag_ext - n_flag_sub))

flag_by_label <- obj@meta.data %>%
  group_by(cell_label) %>%
  summarise(n_cells = n(), n_flag = sum(flag_pos),
            pct_flag_of_label = round(100 * mean(flag_pos), 2), .groups = "drop") %>%
  mutate(pct_of_flag_in_T = round(100 * n_flag / sum(n_flag), 2)) %>%
  arrange(desc(n_flag))
cat("\nFlag+ cells by cluster_label:\n"); print(as.data.frame(flag_by_label))

# A NEW column. cluster_label, cell_label and mc_state are left untouched.
# "pos"/"neg" rather than +/- because a "+" in a factor level breaks formula
# parsing, file names and several ggplot label paths.
obj$cluster_label_flag <- paste0(obj$cell_label,
                                ifelse(obj$flag_pos, "_flagpos", "_flagneg"))
cat("\ncluster_label_flag levels:", length(unique(obj$cluster_label_flag)), "\n")

n_flag_dropped <- sum(flag_by_label$n_flag[flag_by_label$cell_label %in% drop_labels])
cat(sprintf("\nFlag+ in dropped labels (%s): %d (%.2f%% of Flag+ in subset)\n",
            paste(drop_labels, collapse = "/"), n_flag_dropped,
            100 * n_flag_dropped / max(n_flag_sub, 1)))
cat(sprintf("Flag+ retained for plotting:      %d\n", n_flag_sub - n_flag_dropped))
```

```{r save-rds}
# written to its own file, not back over the 05 output: 11 reads this one and
# the annotated object stays reproducible from 05 alone
saveRDS(obj, file.path(DIR_RDS, SUBSET_SLUG, "metacell",
                       paste0(FILE_PREFIX, "_MC_TCR.RDS")))
```

```{r analysis-frames, results = FALSE}
# drop_labels is applied to every figure. Without it Outlier appears with
# ~10,000 cells in clonalOccupy.
#
# subset(), not obj[, cells]. The bracket form drops the reductions in Seurat v5,
# which is what silently disabled clonalNetwork in the previous run.
# Intersected with the cells that actually carry embeddings: DimPlot cbinds the
# embedding matrix onto the metadata and fails with "arguments imply differing
# number of rows" if the two disagree.
keep_cells <- colnames(obj)[!obj$cell_label %in% drop_labels]
keep_cells <- intersect(keep_cells, rownames(Embeddings(obj, "umap")))
obj_plot   <- subset(obj, cells = keep_cells)
Idents(obj_plot) <- factor(obj_plot$cell_label)
stopifnot(nrow(Embeddings(obj_plot, "umap")) == ncol(obj_plot))

md_v <- obj@meta.data %>% filter(has_vdj, !cell_label %in% drop_labels)
md_v$cell_label <- factor(md_v$cell_label, levels = sort(unique(md_v$cell_label)))

PAL_STATE <- cat_pal(md_v$cell_label)

# scRepertoire orders cloneSize largest-first (Hyperexpanded ... Rare), while
# count_bin and occ_bin are cut() output and run smallest-first. Without this the
# two scales point opposite ways and Rare comes out red.
cs_levels     <- levels(droplevels(md_v$cloneSize))
PAL_CLONESIZE <- seq_pal(md_v$cloneSize, rev = grepl("^Hyperexpanded", cs_levels[1]))
PAL_COUNTBIN  <- seq_pal(md_v$count_bin)
cat("cloneSize levels as ordered:", paste(cs_levels, collapse = " | "), "\n")

# patients get their own palette, distinct from the cell-state one: greens for
# CTRL, purples for CASE, so the annotation track cannot be confused with a state.
ctrl_cats <- grep("^CTRL", CAT_LEVELS, value = TRUE)
ec_cats   <- grep("^CASE",   CAT_LEVELS, value = TRUE)
PAL_PATIENT <- setNames(
  c(grDevices::colorRampPalette(c("#00441B", "#A1D99B"))(length(ctrl_cats)),
    grDevices::colorRampPalette(c("#3F007D", "#BCBDDC"))(length(ec_cats))),
  c(ctrl_cats, ec_cats))
```

<!--
Figure choices below follow published repertoire panels: the occurrence-break
stacked bar and clone-size ramp from Cell Rep Methods 2024 Fig 1C, per-clone
phenotype composition bars with a patient annotation track, clone-size
ridgelines per patient, and the patient x state x visit sharing dot plot.

Expanded vs unique (>= 2 cells) is the headline expansion statement because it
rests on no chosen cutoff. Both binned scales are kept anyway, since the
per-state composition view needs them, and the rank-abundance and ridgeline
panels are shown alongside as the binning-free versions.

Clones are never pooled across patients. With a handful of patients and uneven
visit counts, pooling lets the deepest patient set the answer. See
docs/12_clonal.md.
-->

# Expanded versus unique

The one expansion statement that rests on no chosen cutoff. Both Nature Medicine
papers open with it before any binning.

```{r fig-expanded-unique, fig.width = 12, fig.height = 5}
clonal_state <- md_v %>%
  dplyr::count(condition, cell_label, clonal_status, name = "n_cells") %>%
  group_by(condition, cell_label) %>%
  mutate(pct = 100 * n_cells / sum(n_cells), n_tot = sum(n_cells)) %>%
  ungroup() %>% filter(n_tot >= 20)

n_lab <- md_v %>% distinct(condition, patient_id) %>% dplyr::count(condition) %>%
  mutate(lab = sprintf("%s (n = %d patients)", condition, n))
clonal_state <- clonal_state %>%
  left_join(n_lab %>% dplyr::select(condition, lab), by = "condition")

p_expunique <- ggplot(clonal_state, aes(cell_label, pct, fill = clonal_status)) +
  geom_col(width = 0.85) +
  facet_grid(~ lab) +
    geom_text(data = clonal_state %>% distinct(lab, cell_label, n_tot),
        aes(x = cell_label, y = 103, label = n_tot),
        inherit.aes = FALSE, size = 2.2, colour = "grey30") +
  coord_cartesian(ylim = c(0, 108), clip = "off") +
  scale_fill_manual(values = PAL_CLONAL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% of TCR+ cells", fill = "Clonality",
       title = paste0(SUBSET_TITLE, ": clonal expansion landscape"),
       subtitle = "Expanded = clone of 2 or more cells. States with <20 TCR+ cells are dropped.") +
  theme_tcr()
p_expunique
```

# Occurrence spectrum

Cell Rep Methods Fig 1C. Each clonotype is binned by the percentage of its own
sample's repertoire that it occupies, and the bars show what fraction of the
repertoire sits in each occurrence band. Because the bin is a share rather than
a cell count, samples of very different depth sit on the same scale, which is
the objection to the count bins and the reason this panel is worth having.

```{r fig-occurrence, fig.width = 12, fig.height = 5.5}
occ_tbl <- md_v %>%
  filter(!is.na(occ_bin)) %>%
  dplyr::count(category_plot, condition, timepoint, occ_bin, name = "n_cells") %>%
   group_by(category_plot) %>%
  mutate(pct = 100 * n_cells / sum(n_cells), n_tot = sum(n_cells)) %>% ungroup()

p_occ_spec <- ggplot(occ_tbl, aes(category_plot, pct, fill = occ_bin)) +
  geom_col(width = 0.85) +
  facet_cond(~ condition) +
  geom_text(data = occ_tbl %>% distinct(category_plot, condition, n_tot),
      aes(x = category_plot, y = 103, label = n_tot),
      inherit.aes = FALSE, size = 2.2, colour = "grey30") +
  coord_cartesian(ylim = c(0, 108), clip = "off") +
  scale_fill_manual(values = PAL_OCC, drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "Cumulative frequency (% of TCR+ cells)",
       fill = "Occurrence (%)",
       title = paste0(SUBSET_TITLE, ": clonotype occurrence spectrum"),
       subtitle = "Each clone binned by its share of its own sample. Depth-independent.") +
  theme_tcr()
p_occ_spec
```

# Per-clone phenotype composition

Nat Med 2023 ED Fig 3A, which is the panel worth stealing outright. One bar per
clone, stacked by the states its cells occupy, clones sorted along a composition
gradient inside each arm. The sorting is what makes it legible: it turns a wall
of bars into a wave, and a difference between arms shows up as a different
dominant colour rather than as a number.

The strip above the bars is one tile per clone coloured by patient, so a pattern
driven by a single patient is visible immediately rather than being buried.

Only expanded clones are drawn: a singlet clone is one cell and its "composition"
is a single state by construction, which would flatten the gradient.

```{r fig-clone-composition, fig.width = 13, fig.height = 7}
MIN_CLONE_CELLS <- 3   # a 2-cell clone can only ever be 100/0 or 50/50

clone_state <- md_v %>%
  filter(clonalFrequency >= MIN_CLONE_CELLS) %>%
  group_by(condition, patient_id, category, CTstrict) %>%
  mutate(clone_n = n()) %>%
  dplyr::count(condition, patient_id, category, CTstrict, clone_n, cell_label,
               name = "n_cells") %>%
  group_by(condition, patient_id, CTstrict) %>%
  mutate(frac = n_cells / sum(n_cells)) %>% ungroup() %>%
  mutate(clone_key = paste(patient_id, CTstrict, sep = "|"))

# order clones by their dominant state, then by how dominant it is. This is the
# gradient; the alternative (ordering by one chosen state) presupposes an answer.
state_order <- levels(md_v$cell_label)
clone_order <- clone_state %>%
  group_by(condition, clone_key) %>%
  slice_max(frac, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dom = factor(cell_label, levels = state_order)) %>%
  arrange(condition, dom, desc(frac)) %>%
  mutate(ord = row_number()) %>%
  dplyr::select(clone_key, ord, dom)

clone_state <- clone_state %>% left_join(clone_order, by = "clone_key")

cat(sprintf("Clones drawn (>= %d cells): %d CTRL, %d CASE\n", MIN_CLONE_CELLS,
            n_distinct(clone_state$clone_key[clone_state$condition == "CTRL"]),
            n_distinct(clone_state$clone_key[clone_state$condition == "CASE"])))

p_comp_main <- ggplot(clone_state,
                      aes(reorder(clone_key, ord), 100 * frac, fill = cell_label)) +
  geom_col(width = 1) +
  facet_cond(~ condition) +
  scale_fill_manual(values = PAL_STATE) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Individual clones", y = "% of clone's cells", fill = "Cell state",
       subtitle = paste("bar = one clone (>=3 cells);",
                        "striped bar = lineage spanning across different states;",
                        "Clones are sorted by dominant state, so each colour's",
                        "run width = clonal space that state dominates")) +
  theme_tcr() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p_comp_track <- ggplot(clone_state %>% distinct(condition, clone_key, ord, category),
                       aes(reorder(clone_key, ord), 1, fill = category)) +
  geom_tile() +
  facet_cond(~ condition) +
  # PAL_PATIENT, not the cell-state palette: the track and the bars sit on the
  # same x axis, so sharing colours between them makes the figure unreadable.
  scale_fill_manual(values = PAL_PATIENT) +
  scale_y_continuous(expand = c(0, 0)) +
  guides(fill = guide_legend(nrow = 1, title = "Patient")) +
  labs(x = NULL, y = NULL,
       title = paste0(SUBSET_TITLE, ": phenotype composition of individual clones")) +
  theme_void() +
  theme(strip.text = element_blank(),
        legend.position = "top",
        legend.key.size = unit(0.35, "cm"),
        plot.title = element_text(face = "bold", size = rel(1.0)),
        plot.margin = margin(2, 2, 0, 2))

p_clone_comp <- p_comp_track / p_comp_main + plot_layout(heights = c(0.06, 1))
p_clone_comp
```

# Clone size distribution per patient

Nat Med 2023 Fig 1h. Density of clone sizes on a log axis, one row per patient,
with a tick for every individual clone so the tail is honest rather than
smoothed away. Patients are blocked by arm.

```{r fig-ridgeline, fig.width = 9, fig.height = 7, eval = have_ridges}
clone_sizes <- md_v %>%
  distinct(condition, category, patient_id, CTstrict, patient_clone_n) %>%
  filter(!is.na(patient_clone_n)) %>%
  mutate(category = factor(category, levels = rev(CAT_LEVELS)))

p_ridge <- ggplot(clone_sizes, aes(patient_clone_n, category, fill = condition)) +
  ggridges::geom_density_ridges(
    jittered_points = TRUE, position = ggridges::position_points_jitter(height = 0),
    point_shape = "|", point_size = 1.6, point_alpha = 0.5,
    alpha = 0.75, scale = 1.6, linewidth = 0.3) +
  scale_x_log10() +
  scale_fill_manual(values = PAL_COND) +
  labs(x = "Clone size (cells per clone, per patient)", y = NULL, fill = NULL,
       title = paste0(SUBSET_TITLE, ": clone size distribution per patient"),
       subtitle = "Ticks are individual clones.") +
  theme_tcr() + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
p_ridge
```

# Where the big clones sit

Nat Med 2025 Fig 2i and ED2j. Patient on y, cell state on x, one facet per
visit, point size the clone's cell count and point colour the arm. Four
variables in one grid, no wasted space, and it keeps patient identity visible
instead of averaging it away.

```{r fig-sharing-dots, fig.width = 13, fig.height = 6}
TOP_N_CLONES <- 5   # per patient per state per visit

top_clone_dots <- md_v %>%
  dplyr::count(condition, category, patient_id, timepoint, cell_label, CTstrict,
               name = "n_cells") %>%
  group_by(category, timepoint, cell_label) %>%
  slice_max(n_cells, n = TOP_N_CLONES, with_ties = FALSE) %>%
  ungroup() %>%
  filter(n_cells >= 2) %>%
  mutate(category = factor(category, levels = rev(CAT_LEVELS)))

p_dots <- ggplot(top_clone_dots,
                 aes(cell_label, category, size = n_cells, colour = condition)) +
  geom_point(alpha = 0.75) +
  facet_grid(~ timepoint) +
  scale_colour_manual(values = PAL_COND) +
  scale_size_continuous(range = c(1, 7)) +
  labs(x = NULL, y = NULL, size = "Clone size", colour = NULL,
       title = paste0(SUBSET_TITLE, ": largest clones by patient, state and visit"),
       subtitle = sprintf("Top %d expanded clones per patient x state x visit.", TOP_N_CLONES)) +
  theme_tcr()
p_dots
```

# UMAP by clone size

```{r dimplot, fig.width = 15, fig.height = 6}
p_umap_clone <- DimPlot(obj_plot, group.by = "cloneSize", order = TRUE) +
  scale_colour_manual(values = PAL_CLONESIZE, na.value = "grey88") +
  labs(title = "Clone size (proportion scale)") + theme_tcr() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
p_umap_state <- DimPlot(obj_plot, group.by = "cluster_label", label = TRUE,
                        label.size = 3, repel = TRUE) +
  scale_colour_manual(values = PAL_STATE) +
  labs(title = "Cell state") + theme_tcr() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 0, hjust = 0.5))
p_umap_clone + p_umap_state
```

```{r dimplot-2, fig.width = 15, fig.height = 6}
p_umap_count <- DimPlot(obj_plot, group.by = "count_bin", order = TRUE) +
  scale_colour_manual(values = PAL_COUNTBIN, na.value = "grey88") +
  labs(title = "Clone size (cell count)") + theme_tcr() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
p_umap_car <- DimPlot(obj_plot, group.by = "flag_pos", order = TRUE) +
  scale_colour_manual(values = PAL_FLAG) +
  labs(title = "flag by expression") + theme_tcr() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
p_umap_count + p_umap_car
```

```{r dimplot-pdf, results = FALSE}
pdf(file.path(fig_dir, paste0(FILE_PREFIX, "_clonal_umap.pdf")), width = 15, height = 6)
print(p_umap_clone + p_umap_state)
print(p_umap_count + p_umap_car)
print(DimPlot(obj_plot, group.by = "cloneSize", split.by = "condition", order = TRUE) +
        scale_colour_manual(values = PAL_CLONESIZE, na.value = "grey88") +
        labs(title = "Clone size, CTRL vs CASE") + theme_tcr() +
        theme(axis.text.x = element_text(angle = 0, hjust = 0.5)))
invisible(dev.off())
```

# Clone size distributions

Rebuilt from the scRepertoire export tables so each panel can be facetted by
condition and drawn on the shared palette. Same numbers, controllable figure.

```{r export-tables, results = FALSE}
quant_tbl <- add_origin(clonalQuant(combined_lab, cloneCall = CLONE_CALL, chain = "both",
                                    scale = TRUE, exportTable = TRUE))
homeo_tbl <- add_origin(clonalHomeostasis(combined_lab, cloneCall = CLONE_CALL,
                                          exportTable = TRUE))
prop_tbl  <- add_origin(clonalProportion(combined_lab, cloneCall = CLONE_CALL,
                                         clonalSplit = c(1, 10, 25, 100, 500, 1000, 1e5),
                                         exportTable = TRUE))

abund_tbl <- add_origin(clonalAbundance(combined_lab, cloneCall = CLONE_CALL,
                                        exportTable = TRUE))
len_tbl   <- add_origin(clonalLength(combined_lab, cloneCall = "aa", chain = "both",
                                     exportTable = TRUE))

# long form for plotting: one row per sample x bin, with condition attached
to_long <- function(d, name_to) {
  num <- names(d)[vapply(d, is.numeric, logical(1))]
  d %>% dplyr::select(sample_origin, all_of(num)) %>%
    pivot_longer(all_of(num), names_to = name_to, values_to = "value") %>%
    mutate(!!name_to := factor(.data[[name_to]], levels = num)) %>%
    left_join(sample_meta, by = "sample_origin") %>%
    mutate(category_plot = factor(category_plot, levels = CATPLOT_LEVELS))
}
```

```{r fig-homeostasis, fig.width = 11, fig.height = 5}
homeo_long <- to_long(homeo_tbl, "bin")
p_homeo <- ggplot(homeo_long, aes(category_plot, 100 * value, fill = bin)) +
  geom_col(width = 0.8) +
  facet_cond(~ condition) +
  scale_fill_manual(values = seq_pal(homeo_long$bin)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% of repertoire", fill = "clone size",
       title = paste0(SUBSET_TITLE, ": clonal space homeostasis"),
       subtitle = "Proportion scale. How much of each repertoire the expanded clones occupy.") +
  theme_tcr()
p_homeo
```

```{r fig-proportion, fig.width = 11, fig.height = 5}
prop_long <- to_long(prop_tbl, "rank_band")
p_prop <- ggplot(prop_long, aes(category_plot, 100 * value, fill = rank_band)) +
  geom_col(width = 0.8) +
  facet_cond(~ condition) +
  scale_fill_manual(values = seq_pal(prop_long$rank_band)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% of repertoire", fill = "clone rank",
       title = paste0(SUBSET_TITLE, ": repertoire held by the top N ranked clones")) +
  theme_tcr()
p_prop
```

```{r fig-quant, fig.width = 11, fig.height = 4.5}
# quant_tbl already carries category_plot from add_origin(). Joining all of
# sample_meta suffixed it to category_plot.x / .y, so the mutate below could not
# find it. Only the columns quant_tbl lacks are pulled in.
quant_plot <- quant_tbl %>%
  left_join(sample_meta %>% dplyr::select(sample_origin, condition),
            by = "sample_origin") %>%
  mutate(category_plot = factor(category_plot, levels = CATPLOT_LEVELS))
yv <- if ("scaled" %in% names(quant_plot)) "scaled" else "values"
p_quant <- ggplot(quant_plot, aes(category_plot, .data[[yv]], fill = condition)) +
  geom_col(width = 0.8) +
  facet_cond(~ condition) +
  scale_fill_manual(values = PAL_COND, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "unique clones per 100 cells",
       title = paste0(SUBSET_TITLE, ": unique clones, scaled to cells sequenced")) +
  theme_tcr()
p_quant
```

```{r fig-rank-abundance, fig.width = 10, fig.height = 5}
# rank-abundance on log-log. A steeper head means a more clonally dominated
# repertoire, and it does not depend on any binning choice.
rank_ab <- md_v %>%
  dplyr::count(category_plot, category, condition, timepoint, CTstrict, name = "n") %>%
  group_by(category_plot) %>%
  arrange(desc(n), .by_group = TRUE) %>%
  mutate(rank = row_number(), prop = n / sum(n)) %>%
  ungroup()
# coloured by patient, not condition: with one line per sample the arm is already
# given by the facet, so the colour is free to say which patient a line belongs to.
p_rank <- ggplot(rank_ab, aes(rank, 100 * prop, group = category_plot, colour = category)) +
  geom_line(alpha = 0.85, linewidth = 0.5) +
  facet_grid(~ condition) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = PAL_PATIENT) +
  labs(x = "clone rank", y = "% of TCR+ cells", colour = "Patient",
       title = paste0(SUBSET_TITLE, ": rank-abundance"),
       subtitle = "One line per sample, coloured by patient. A heavier head is a more dominated repertoire.") +
  theme_tcr() + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
p_rank
```

```{r fig-occupancy, fig.width = 11, fig.height = 5}
occ_samp <- md_v %>%
  dplyr::count(category_plot, condition, cloneSize, name = "n_cells") %>%
    group_by(category_plot) %>%
  mutate(pct = 100 * n_cells / sum(n_cells), n_tot = sum(n_cells)) %>% ungroup()


p_occ_samp <- ggplot(occ_samp, aes(category_plot, pct, fill = cloneSize)) +
  geom_col(width = 0.8) +
  facet_cond(~ condition) +
  geom_text(data = occ_samp %>% distinct(category_plot, condition, n_tot),
      aes(x = category_plot, y = 103, label = n_tot),
      inherit.aes = FALSE, size = 2.2, colour = "grey30") +
  coord_cartesian(ylim = c(0, 108), clip = "off") +
  scale_fill_manual(values = PAL_CLONESIZE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% of TCR+ cells", fill = "clone size",
       title = paste0(SUBSET_TITLE, ": clone size occupancy by sample")) +
  theme_tcr()
p_occ_samp
```

```{r fig-occupancy-state, fig.width = 12, fig.height = 5}
occ_state <- md_v %>%
  dplyr::count(condition, cell_label, cloneSize, name = "n_cells") %>%
   group_by(condition, cell_label) %>%
  mutate(pct = 100 * n_cells / sum(n_cells), n_tot = sum(n_cells)) %>% ungroup()


p_occ_state <- ggplot(occ_state, aes(cell_label, pct, fill = cloneSize)) +
  geom_col(width = 0.8) +
  facet_grid(~ condition) +
  geom_text(data = occ_state %>% distinct(condition, cell_label, n_tot),
        aes(x = cell_label, y = 103, label = n_tot),
        inherit.aes = FALSE, size = 2.2, colour = "grey30") +
  coord_cartesian(ylim = c(0, 108), clip = "off") +
  scale_fill_manual(values = PAL_CLONESIZE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% of TCR+ cells", fill = "clone size",
       title = paste0(SUBSET_TITLE, ": clone size occupancy by state")) +
  theme_tcr()
p_occ_state
```

# Expansion stratified by state and visit

The panel this is modelled on stratifies expansion by cell state and by group,
so a reader sees at a glance which states carry the expanded clones in which
arm. Two versions, because the aggregation question is real:

  - **per sample**, one point per sample, timepoints kept separate. This is the
    honest version: pooling T0 with T4 inside a patient averages over the
    biology being asked about.
  - **composition over time**, clone size stacked within each state at each
    visit. Denser, and shows the direction of travel rather than a snapshot.

Nothing here is aggregated to the patient. Patient-level aggregation happens
once, in the stats notebook, where it is the unit of the test and is flagged.

```{r fig-expansion-by-state, fig.width = 12, fig.height = 5.5}
exp_by_state_sample <- md_v %>%
  group_by(category_plot, category, condition, timepoint, cell_label) %>%
  summarise(n_cells = n(), pct_expanded = 100 * mean(clonalFrequency > 1),
            .groups = "drop") %>%
  filter(n_cells >= 20)   # a state with a handful of cells gives a meaningless percentage

p_exp_state <- ggplot(exp_by_state_sample,
                      aes(cell_label, pct_expanded, colour = condition)) +
  geom_boxplot(aes(fill = condition), alpha = 0.15, outlier.shape = NA,
               linewidth = 0.4, show.legend = FALSE) +
  geom_point(position = position_jitter(width = 0.15, height = 0), size = 1.6, alpha = 0.85) +
  facet_grid(~ condition) +
  scale_colour_manual(values = PAL_COND, guide = "none") +
  scale_fill_manual(values = PAL_COND, guide = "none") +
  labs(x = NULL, y = "% of TCR+ cells in expanded clones",
       title = paste0(SUBSET_TITLE, ": clonal expansion by cell state"),
       subtitle = "One point per sample. States with <20 TCR+ cells in a sample are dropped.") +
  theme_tcr()
p_exp_state
```

```{r fig-composition-time, fig.width = 12, fig.height = 8}
bin_by_state <- md_v %>%
  dplyr::count(condition, timepoint, cell_label, cloneSize, name = "n_cells") %>%
  group_by(condition, timepoint, cell_label) %>%
  mutate(pct = 100 * n_cells / sum(n_cells), n_tot = sum(n_cells)) %>%
  ungroup() %>%
  filter(n_tot >= 20)

p_comp_time <- ggplot(bin_by_state, aes(timepoint, pct, fill = cloneSize)) +
  geom_col(width = 0.85) +
  facet_grid(cell_label ~ condition, switch = "y") +
  geom_text(data = bin_by_state %>% distinct(condition, timepoint, cell_label, n_tot),
      aes(x = timepoint, y = 106, label = n_tot),
      inherit.aes = FALSE, size = 1.8, colour = "grey30") +
  coord_cartesian(ylim = c(0, 115), clip = "off") +
  scale_fill_manual(values = PAL_CLONESIZE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% of TCR+ cells", fill = "clone size",
       title = paste0(SUBSET_TITLE, ": clone size composition over time, by state"),
       subtitle = "Timepoints kept separate. Cells of a state x visit with <20 TCR+ cells are dropped.") +
  theme_tcr() +
  theme(strip.text.y.left = element_text(angle = 0, hjust = 1, size = rel(0.75)),
        strip.placement = "outside",
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = rel(0.8)))
p_comp_time
```

```{r fig-expansion-time, fig.width = 11, fig.height = 5}
exp_traj <- md_v %>%
  group_by(category, patient_id, condition, timepoint) %>%
  summarise(pct_expanded = 100 * mean(clonalFrequency > 1), n_cells = n(), .groups = "drop")

p_exp_traj <- ggplot(exp_traj, aes(timepoint, pct_expanded,
                                   group = category, colour = category)) +
  geom_line(alpha = 0.85, linewidth = 0.6) + geom_point(size = 2) +
  facet_grid(~ condition) +
  coord_cartesian(ylim = c(0, 108), clip = "off") +
  scale_colour_manual(values = PAL_PATIENT) +
  labs(x = NULL, y = "% of TCR+ cells in expanded clones", colour = "Patient",
       title = paste0(SUBSET_TITLE, ": expansion over time")) +
  theme_tcr() + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
p_exp_traj
```

```{r bins-tables}
count_bins_by_state <- md_v %>%
  dplyr::count(condition, timepoint, cell_label, count_bin, name = "n_cells") %>%
  group_by(condition, timepoint, cell_label) %>%
  mutate(pct = round(100 * n_cells / sum(n_cells), 2)) %>% ungroup()

state_expansion <- md_v %>%
  group_by(condition, cell_label) %>%
  summarise(n_cells = n(), n_clones = n_distinct(CTstrict),
            pct_expanded = round(100 * mean(clonalFrequency > 1), 2),
            median_clone_size = median(clonalFrequency), .groups = "drop")
print(as.data.frame(state_expansion))
```

# Gene usage and clone export

```{r gene-usage}
gene_usage <- bind_rows(lapply(c("TRBV", "TRBJ", "TRAV", "TRAJ"), function(g)
  cbind(gene_class = g,
        add_origin(vizGenes(combined_lab, x.axis = g, plot = "heatmap",
                            exportTable = TRUE)))))

exported_clones <- add_origin(exportClones(combined_lab, write.file = FALSE))
```

# Flag+ clones

```{r car-clones, fig.width = 11, fig.height = 5}
car_clones <- md_v %>%
  group_by(category, patient_id, condition, CTstrict) %>%
  summarise(n_cells = n(), n_flag = sum(flag_pos),
            pct_flag = round(100 * mean(flag_pos), 2),
            visits = paste(sort(unique(as.character(timepoint))), collapse = ","),
            states = paste(sort(unique(cell_label)), collapse = ","),
            .groups = "drop") %>%
  filter(n_flag > 0) %>% arrange(desc(n_flag))

car_expansion <- md_v %>%
  group_by(condition, timepoint, flag_pos) %>%
  summarise(n_cells = n(), n_clones = n_distinct(CTstrict),
            pct_expanded = round(100 * mean(clonalFrequency > 1), 2), .groups = "drop")
print(as.data.frame(car_expansion))

# clone size composition of Flag+ against Flag- cells
car_comp <- md_v %>%
  filter(!(timepoint == "T0" & flag_pos)) %>%   # no flag pre-infusion; these are detector background
  dplyr::count(condition, timepoint, flag_pos, cloneSize, name = "n_cells") %>%
  group_by(condition, timepoint, flag_pos) %>%
  mutate(pct = 100 * n_cells / sum(n_cells), n_tot = sum(n_cells)) %>%
  ungroup() %>% filter(n_tot >= 20) %>%
  mutate(flag = factor(ifelse(flag_pos, "Flag+", "Flag-"), levels = c("Flag-", "Flag+")))


p_car_comp <- ggplot(car_comp, aes(flag, pct, fill = cloneSize)) +
  geom_col(width = 0.8) +
  facet_grid(condition ~ timepoint) +
  geom_text(data = car_comp %>% distinct(condition, timepoint, flag, n_tot),
        aes(x = flag, y = 103, label = n_tot),
        inherit.aes = FALSE, size = 2.4, colour = "grey30") +
  coord_cartesian(ylim = c(0, 108), clip = "off") +
  scale_fill_manual(values = PAL_CLONESIZE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% of TCR+ cells", fill = "clone size",
       title = paste0(SUBSET_TITLE, ": clone size composition, Flag+ vs Flag-"),
       subtitle = "Cells of a flag x visit group with <20 TCR+ cells are dropped.") +
  theme_tcr() + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
p_car_comp
```

## Clone size by state and flag status, over time

`p_comp_time` with flag positivity added. `cluster_label_flag` (state + `_flagpos` /
`_flagneg`) is on the object for any other use; the figure splits it back into a
state row and a flag column rather than putting 2 x n_states labels down one axis,
because that is the same information and stays readable.

Flag- is the internal control for Flag+ within the same state and visit, which is
the comparison the fragmented label is for.

```{r fig-composition-car-state, fig.width = 13, fig.height = 9}
bin_state_car <- md_v %>%
  filter(!(timepoint == "T0" & flag_pos)) %>%
  mutate(flag = factor(ifelse(flag_pos, "flagpos", "flagneg"),
                      levels = c("flagneg", "flagpos"))) %>%
  dplyr::count(condition, flag, timepoint, cell_label, cloneSize, name = "n_cells") %>%
  group_by(condition, flag, timepoint, cell_label) %>%
  mutate(pct = 100 * n_cells / sum(n_cells), n_tot = sum(n_cells)) %>%
  ungroup() %>%
  filter(n_tot >= 20)

p_comp_car_state <- ggplot(bin_state_car, aes(timepoint, pct, fill = cloneSize)) +
  geom_col(width = 0.85) +
  facet_grid(cell_label ~ condition + flag, switch = "y") +
    geom_text(data = bin_state_car %>% distinct(condition, flag, timepoint, cell_label, n_tot),
      aes(x = timepoint, y = 106, label = n_tot),
      inherit.aes = FALSE, size = 1.7, colour = "grey30") +
  coord_cartesian(ylim = c(0, 115), clip = "off") +
  scale_fill_manual(values = PAL_CLONESIZE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(x = NULL, y = "% of TCR+ cells", fill = "clone size",
       title = paste0(SUBSET_TITLE, ": clone size by state and flag status, over time"))+#,
       # subtitle = "Timepoints kept separate. A state x visit x flag group with <20 TCR+ cells is dropped, so gaps are absent data, not zero expansion.") +
  theme_tcr() +
  theme(strip.text.y.left = element_text(angle = 0, hjust = 1, size = rel(0.7)),
        strip.placement = "outside",
        axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.7)))
p_comp_car_state
```

```{r clonal-network, fig.width = 10, fig.height = 7}
net_tbl <- clonalNetwork(obj_plot, reduction = "umap", group.by = "cell_label",
                         filter.clones = NULL, filter.identity = NULL,
                         cloneCall = CLONE_CALL, exportTable = TRUE)
net_tbl <- as.data.frame(net_tbl)
cat("clonalNetwork exportTable columns:", paste(names(net_tbl), collapse = ", "), "\n")

wt <- intersect(c("weight", "value", "prop"), names(net_tbl))[1]
p_net <- ggplot(net_tbl, aes(to, from, fill = .data[[wt]])) +
  geom_tile(colour = "grey90", linewidth = 0.2) +
  scale_fill_gradientn(colours = PAL_SEQ) +
  coord_equal() +
  labs(x = "also found in", y = "clones from", fill = wt,
       title = paste0(SUBSET_TITLE, ": clonal sharing between states"),
       subtitle = "clonalNetwork numbers drawn as a matrix") +
  theme_tcr() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = rel(0.75)),
        axis.text.y = element_text(size = rel(0.75)),
        panel.grid = element_blank())
p_net
```

# Write the figures

```{r figures-pdf, results = FALSE}
to_pdf <- function(file, width, height, exprs) {
  pdf(file.path(fig_dir, file), width = width, height = height)
  on.exit(invisible(dev.off()), add = TRUE)
  for (nm in names(exprs)) {
    tryCatch(print(exprs[[nm]]),
             error = function(e) message("figure skipped: ", nm, " -- ",
                                         conditionMessage(e)))
  }
}

to_pdf(paste0(FILE_PREFIX, "_clonal_distributions.pdf"), 12, 6, list(
  quant = p_quant, homeostasis = p_homeo, proportion = p_prop,
  rank_abundance = p_rank, occupancy_sample = p_occ_samp,
  occupancy_state = p_occ_state, occurrence = p_occ_spec,
  expanded_unique = p_expunique, expansion_state = p_exp_state,
  expansion_time = p_exp_traj, car_composition = p_car_comp))

to_pdf(paste0(FILE_PREFIX, "_clone_size_composition.pdf"), 13, 9, list(
  composition_time = p_comp_time, composition_car_state = p_comp_car_state,
  clone_composition = p_clone_comp))

to_pdf(paste0(FILE_PREFIX, "_clone_landscape.pdf"), 13, 7, c(
  list(clone_dots = p_dots),
  list(ridgeline = p_ridge)))

to_pdf(paste0(FILE_PREFIX, "_clonal_network.pdf"), 10, 7, list(network_matrix = p_net))
```

# Write the workbook

Every table is listed once, here, after all of it has been computed. Nothing
accumulates in a global as the notebook runs, so re-running a chunk cannot
corrupt the workbook and the sheet list is readable in one screen.

```{r write}
sheet <- function(d, desc) {
  if (is.null(d)) return(NULL)
  d <- as.data.frame(d)
  if (!nrow(d)) return(NULL)
  list(d = d, desc = desc)
}

BOOK <- list(
  contig_file_map = sheet(path_map,
    "Which contig file was used for each sample, and its sequencing batch. Verify by eye once."),
  vdj_recovery = sheet(vdj_qc,
    "Cells recovered with a TCR per sample, after filtering contigs to object barcodes."),
  chain_dropout = sheet(dropout_qc,
    "Alpha-chain dropout per sample. High pct_beta_only inflates n_clones and deflates clonality."),
  car_by_broad_label = sheet(if (exists("flag_broad")) flag_broad else NULL,
    "Where Flag+ cells land across the whole dataset, before subsetting to T cells."),
  car_by_cluster_label = sheet(flag_by_label,
    "Flag+ cells per annotated T cell state."),
  expanded_vs_unique = sheet(clonal_state %>% mutate(pct = round(pct, 2)),
    "Percent of TCR+ cells in expanded (>=2 cell) versus unique clones, per state per condition."),
  occurrence_spectrum = sheet(occ_tbl %>% mutate(pct = round(pct, 3)),
    "Cumulative frequency of clonotypes per occurrence band per sample. Cell Rep Methods Fig 1C."),
  clone_phenotype_composition = sheet(
    clone_state %>% mutate(frac = round(frac, 4)) %>%
      dplyr::select(condition, category, patient_id, CTstrict, clone_n,
                    cell_label, n_cells, frac),
    "State composition of every clone with >=3 cells. The data behind the per-clone composition bars."),
  top_clones_by_state_visit = sheet(top_clone_dots,
    "Largest expanded clones per patient, state and visit. The data behind the sharing dot plot."),
  unique_clones = sheet(quant_tbl,
    "Unique clones per sample, scaled by cells sequenced."),
  homeostasis = sheet(homeo_tbl,
    "Fraction of repertoire held by rare/small/medium/large/hyperexpanded clones, per sample."),
  proportion = sheet(prop_tbl,
    "Fraction of repertoire held by the top N ranked clones, per sample."),
  abundance = sheet(abund_tbl,
    "Every clone with its size, per sample."),
  cdr3_length = sheet(len_tbl,
    "CDR3 amino-acid length distribution."),
  rank_abundance = sheet(rank_ab,
    "Every clone with its within-sample rank and proportion. The data behind the rank-abundance curve."),
  expansion_by_state_sample = sheet(exp_by_state_sample,
    "Percent of TCR+ cells in expanded clones, per state per sample. One point per sample in the expansion figure."),
  count_bins_by_state = sheet(count_bins_by_state,
    "Cells per clone-size COUNT bin per state. Count scale, not scRepertoire's proportion scale."),
  prop_bins_by_state = sheet(bin_by_state %>% mutate(pct = round(pct, 2)),
    "The same table on scRepertoire's PROPORTION scale (cloneSize). Kept separate on purpose."),
  state_expansion_by_cond = sheet(state_expansion,
    "Which states carry expanded clones, pooled within a condition. Descriptive, not tested."),
  gene_usage = sheet(gene_usage,
    "V and J gene usage per sample. Also corroborates the public-clone batch question."),
  exported_clones = sheet(exported_clones,
    "Full scRepertoire clone export, one row per cell."),
  car_positive_clones = sheet(car_clones,
    "Clones containing at least one Flag+ cell, with the visits and states they span."),
  car_expansion = sheet(car_expansion,
    "Expansion of Flag+ versus Flag- cells at each visit."),
  car_clone_size_composition = sheet(car_comp %>% mutate(pct = round(pct, 2)),
    "Clone size composition of Flag+ versus Flag- cells at each visit."),
  clone_size_state_by_car = sheet(bin_state_car %>% mutate(pct = round(pct, 2)),
    "Clone size composition per state per visit, split by flag status. Gaps are groups with <20 TCR+ cells.")
)

wb <- createWorkbook()
for (nm in names(BOOK)) {
  s <- BOOK[[nm]]
  if (is.null(s)) { cat("empty, not written:", nm, "\n"); next }
  stopifnot(nchar(nm) <= 31)
  addWorksheet(wb, nm)
  writeData(wb, nm, s$desc, startRow = 1, startCol = 1, colNames = FALSE)
  writeData(wb, nm, s$d,    startRow = 3, rowNames = FALSE)
  freezePane(wb, nm, firstActiveRow = 4)
}
f <- file.path(clone_root, "subset_data_clonal_TCR.xlsx")
saveWorkbook(wb, f, overwrite = TRUE)
cat("wrote", basename(f), "-", sum(!vapply(BOOK, is.null, logical(1))), "sheets\n")
```

<!--
NOTE ON cloneSize CUTOFFS

Neither set of cutoffs is derived from anything; they are display conventions
and the two scales are not comparable.

  rnabio, count scale (proportion = FALSE)
    Single 1, Small <=5, Medium <=20, Large <=100, Hyperexpanded <=500
  here / scRepertoire default, proportion scale (proportion = TRUE)
    Rare <=1e-4, Small <=1e-3, Medium <=1e-2, Large <=0.1, Hyperexpanded <=1

Count bins are confounded by depth: a 30-cell clone is 3% of a 1,000-cell
sample and 0.3% of a 10,000-cell sample, and lands in the same bin. Samples
here differ several-fold in TCR+ cell number, so the proportion scale is kept.
The rank-abundance figure is the same information with no binning at all.

Other rnabio defaults not used:
  cloneCall = "aa"    aa merges distinct rearrangements that share a CDR3aa.
                      "strict" (V/J + nt CDR3) is the conservative call within
                      a patient. aa is used only for public clones, in 11.
  group.by = "sample" equivalent to "none" given one list element per sample.
  column renaming     only needed when TCR and BCR are attached to one object.

Full reasoning: docs/12_clonal.md.
-->
