---
title: "11 Clonal repertoire: diversity, tracking and statistics"
author: "Saumya Pothukuchi"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    code_folding: hide
---

<!--
Reads the RDS written by 10 and does everything that notebook left out:
rarefied diversity, repertoire tracking across visits, clone sharing between
states, Startrac, public clones, and the four test families.

Nothing here touches the contig files. chain_status and clone_trb were computed
upstream and travel on the object.

Metric set: Renyi entropy profiles (Shannon is the single point at alpha = 1),
rarefaction curves observed and extrapolated, Gini, Tversky asymmetric sharing
between states, and clone-count biplots between paired visits. Why each one and
what it is sensitive to: docs/12_clonal.md.

Clones are never pooled across patients. Patient stays the unit throughout.

Writes one workbook. The run manifest is the only place a skipped analysis is
distinguishable from an empty one.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(width = 60)
```

```{r libs, results = FALSE}
library(dplyr)
library(tidyr)
library(Seurat)
library(scRepertoire)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(lme4)
library(lmerTest)
library(openxlsx)
```

R: `r R.Version()$version.string`,
Seurat: `r packageVersion('Seurat')`,
scRepertoire: `r packageVersion('scRepertoire')`,
lme4: `r packageVersion('lme4')`,
lmerTest: `r packageVersion('lmerTest')`

# What the replicate is

Every metric is first reduced to one number per sample, then tested with patient
as the unit, with patient as a random intercept for anything that spans
timepoints. Repertoire metrics depend strongly
on how many cells were sequenced, so diversity is computed on repeated
rarefactions to a common depth rather than on raw counts, and samples below
`MIN_VDJ_CELLS` are dropped rather than reported.

With small arms the smallest attainable Wilcoxon p is bounded by the group sizes
alone, and only under perfect separation. The condition family is therefore
reported with effect sizes and an explicit `powered` flag rather than as a test
that can succeed.

Every figure axis is labelled with `category_plot`.
`sample_origin` stays in every table as the join key, never as a figure axis.

# Configuration

```{r config, results = FALSE}
# Same config as 10, so the two notebooks cannot disagree about levels or paths.
source(file.path("..", "config", "config.R"))

SUBSET_SLUG  <- "subset"
FILE_PREFIX  <- "subset_data"
SUBSET_TITLE <- "Subset"

# the object 10 wrote, with the repertoire already attached
rds_path   <- file.path(DIR_RDS, SUBSET_SLUG, "metacell",
                        paste0(FILE_PREFIX, "_MC_TCR.RDS"))
clone_root <- file.path(out_dirs(SUBSET_SLUG, "csv"), "clonal")
fig_dir    <- file.path(out_dirs(SUBSET_SLUG, "pdf"), "clonal")
for (d in c(clone_root, fig_dir)) dir.create(d, recursive = TRUE, showWarnings = FALSE)

# Diversity is depth-dependent, so samples are rarefied to a common cell count
# before any index is computed. Both numbers are judgement calls: raising
# MIN_VDJ_CELLS keeps the estimates tighter but drops samples, and RAREFY_TO
# should sit at or below the smallest retained sample.
MIN_VDJ_CELLS       <- 750   # a sample below this cannot support a diversity estimate
RAREFY_TO           <- 600   # NA means the smallest retained sample
MIN_CELLS_UNIT      <- 20    # per state, per sample, for the state-wise tables
MIN_UNITS_PER_GROUP <- 2
N_BOOTS             <- 100
DROPOUT_FLAG_PCT    <- 10    # warn if a sample exceeds this % alpha-less cells
RENYI_ALPHA <- c(0, 0.25, 0.5, 1, 2, 4, 8, 16, 32, 64, Inf)
set.seed(123)

MANIFEST <- list()

log_row <- function(family, unit, item, n_test, n_ref, n_units, status, note = NA_character_)
  MANIFEST[[length(MANIFEST) + 1]] <<- data.frame(
    family = family, unit = unit, item = item, n_test = n_test, n_ref = n_ref,
    n_units = n_units, status = status, note = note, stringsAsFactors = FALSE)

safe <- function(expr, family, unit, item) {
  tryCatch(expr, error = function(e) {
    log_row(family, unit, item, NA_integer_, NA_integer_, NA_integer_, "skipped",
            substr(conditionMessage(e), 1, 200))
    NULL
  })
}
```

## Palette and theme

Identical to the main notebook, so the two sets of figures read as one set.
Clone-size ramp from Mhanna et al., Cell Rep Methods 2024, Fig 1C.

```{r palette, results = FALSE}
PAL_COND   <- c(CTRL = "#4C7FB8", CASE = "#A02C2C")
PAL_CLONAL <- c(Unique = "#4FBF6E", Expanded = "#3F4E8C")
PAL_SEQ    <- c("#BBD7EA", "#7FB2D4", "#3D7FB0", "#FBD277", "#EE8B3C", "#B62B2B")
PAL_CAT    <- c("#4C72B0", "#DD8452", "#55A868", "#C44E52", "#8172B3", "#937860",
                "#DA8BC3", "#8C8C8C", "#CCB974", "#64B5CD", "#4A6FE3", "#A0522D")

cat_pal <- function(x) {
  lv <- if (is.factor(x)) levels(droplevels(x)) else sort(unique(x[!is.na(x)]))
  setNames(rep(PAL_CAT, length.out = length(lv)), lv)
}

theme_tcr <- function(base_size = 11) {
  theme_bw(base_size) +
    theme(panel.grid.minor   = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.border       = element_rect(colour = "grey35", linewidth = 0.5),
          strip.background   = element_rect(fill = "grey92", colour = "grey35"),
          strip.text         = element_text(face = "bold", size = rel(0.85)),
          axis.text.x        = element_text(angle = 45, hjust = 1, vjust = 1),
          legend.key.size    = unit(0.4, "cm"),
          plot.title         = element_text(face = "bold", size = rel(1.0)),
          plot.subtitle      = element_text(colour = "grey30", size = rel(0.85)))
}
facet_cond <- function(...) facet_grid(..., scales = "free_x", space = "free_x")
```

# Load the TCR object

```{r load}
obj <- readRDS(rds_path)

need <- c("CTstrict", "CTaa", "clone_trb", "chain_status", "has_vdj", "cell_label",
          "clonalFrequency", "patient_clone_prop", "batch", "category_plot",
          "clonal_status")
stopifnot(all(need %in% colnames(obj@meta.data)))
Idents(obj) <- factor(obj$cell_label)

cat("Cells in object:", ncol(obj), "| with a TCR:", sum(obj$has_vdj),
    sprintf("(%.1f%%)\n", 100 * mean(obj$has_vdj)))

sample_meta <- obj@meta.data %>%
  distinct(sample_origin, category_plot, category, patient_id, condition, timepoint, batch) %>%
  arrange(category_plot)
stopifnot(!anyDuplicated(sample_meta$sample_origin))
CAT_LEVELS <- levels(droplevels(factor(sample_meta$category)))

patient_meta <- sample_meta %>% distinct(patient_id, category, condition)
stopifnot(!anyDuplicated(patient_meta$patient_id))

# a patient with one sequenced visit gives a structural 0% persistence and a
# "patient average" that is really a single sample. Flagged so it travels with
# every table.
patient_visits <- sample_meta %>%
  dplyr::count(patient_id, category, condition, name = "n_visits_seq") %>%
  mutate(single_visit = n_visits_seq < 2)
print(as.data.frame(patient_visits))

cat_of <- setNames(as.character(sample_meta$category_plot), sample_meta$sample_origin)

# drop_labels applied once, here. md_v is the analysis frame for everything below.
md   <- obj@meta.data
md_v <- md %>% filter(has_vdj, !cell_label %in% drop_labels)
md_v$cell_label <- factor(md_v$cell_label, levels = sort(unique(md_v$cell_label)))
PAL_STATE <- cat_pal(md_v$cell_label)
cat("Cells entering the analysis frame:", nrow(md_v), "\n")
```

# Helpers

Chao1 is not computed. On singleton-dominated TCR data it ran from 637 to 60,000
after rarefaction, and because it sat in the same BH family it inflated the
adjusted p-values of every other metric.

```{r helpers, results = FALSE}
# Gini inequality on the clone size vector. Nat Med 2023 use this as their
# clonality index. 0 = every clone the same size, 1 = one clone holds everything.
gini <- function(x) {
  x <- sort(x[!is.na(x)]); n <- length(x)
  if (n < 2 || sum(x) == 0) return(NA_real_)
  sum((2 * seq_len(n) - n - 1) * x) / (n * sum(x))
}

# Renyi entropy of order alpha. alpha = 0 -> log(richness), alpha -> 1 ->
# Shannon, alpha = 2 -> log(inverse Simpson), alpha -> Inf -> -log(max p).
renyi <- function(p, alpha) {
  p <- p[p > 0]
  if (is.infinite(alpha)) return(-log(max(p)))
  if (abs(alpha - 1) < 1e-9) return(-sum(p * log(p)))
  log(sum(p^alpha)) / (1 - alpha)
}

div_metrics <- function(clones) {
  clones <- clones[!is.na(clones)]
  tab <- table(clones)
  p   <- as.numeric(tab) / sum(tab)
  n   <- sum(tab)
  data.frame(
    n_cells      = n,
    n_clones     = length(tab),
    shannon      = -sum(p * log(p)),
    eff_clones   = exp(-sum(p * log(p))),   # e^shannon, the readable version
    norm_entropy = if (length(tab) > 1) -sum(p * log(p)) / log(length(tab)) else NA_real_,
    inv_simpson  = 1 / sum(p^2),
    gini_simpson = 1 - sum(p^2),
    gini         = gini(as.numeric(tab)),
    clonality    = if (length(tab) > 1) 1 + sum(p * log(p)) / log(length(tab)) else NA_real_,
    pct_expanded = 100 * sum(tab[tab > 1]) / n,
    top10_pct    = 100 * sum(sort(as.numeric(tab), decreasing = TRUE)[seq_len(min(10, length(tab)))]) / n,
    stringsAsFactors = FALSE)
}

div_rarefied <- function(clones, depth, n_boots = N_BOOTS) {
  clones <- clones[!is.na(clones)]
  if (length(clones) < depth) return(NULL)
  reps <- bind_rows(lapply(seq_len(n_boots), function(i)
    div_metrics(sample(clones, depth, replace = FALSE))))
  as.data.frame(lapply(reps, function(v) mean(v, na.rm = TRUE)))
}

# Renyi profile on repeated rarefactions, so the curve is comparable across
# samples of different depth
renyi_rarefied <- function(clones, depth, alphas = RENYI_ALPHA, n_boots = N_BOOTS) {
  clones <- clones[!is.na(clones)]
  if (length(clones) < depth) return(NULL)
  m <- vapply(seq_len(n_boots), function(i) {
    tab <- table(sample(clones, depth, replace = FALSE))
    p   <- as.numeric(tab) / sum(tab)
    vapply(alphas, function(a) renyi(p, a), numeric(1))
  }, numeric(length(alphas)))
  data.frame(alpha = alphas, renyi = rowMeans(m),
             renyi_sd = apply(m, 1, stats::sd), stringsAsFactors = FALSE)
}

# rank-biserial correlation, a rank effect size needing no distributional assumption
rb_effect <- function(x, y) {
  if (!length(x) || !length(y)) return(NA_real_)
  u <- sum(outer(x, y, ">")) + 0.5 * sum(outer(x, y, "=="))
  2 * u / (length(x) * length(y)) - 1
}

morisita <- function(a, b) {
  a <- a[!is.na(a)]; b <- b[!is.na(b)]
  ta <- table(a); tb <- table(b)
  u  <- union(names(ta), names(tb))
  x  <- as.numeric(ta[u]); x[is.na(x)] <- 0
  y  <- as.numeric(tb[u]); y[is.na(y)] <- 0
  X  <- sum(x); Y <- sum(y)
  if (X < 2 || Y < 2) return(NA_real_)
  la <- sum(x * (x - 1)) / (X * (X - 1))
  lb <- sum(y * (y - 1)) / (Y * (Y - 1))
  if ((la + lb) == 0) return(NA_real_)
  2 * sum(x * y) / ((la + lb) * X * Y)
}

# Tversky asymmetric similarity, a = 1 and b = 0, i.e. the fraction of A's
# clones that also appear in B. Asymmetric on purpose: a small state can be
# almost contained in a large one without the reverse holding.
tversky <- function(A, B, a = 1, b = 0) {
  A <- unique(A[!is.na(A)]); B <- unique(B[!is.na(B)])
  if (!length(A) || !length(B)) return(NA_real_)
  i <- length(intersect(A, B))
  i / (i + a * length(setdiff(A, B)) + b * length(setdiff(B, A)))
}
```

# Rarefaction curves

Observed clonotype richness as a function of how many cells are drawn. Two
samples are only comparable where their curves overlap, which is the whole
argument for rarefying rather than reporting richness as sequenced. The vertical
line is `RAREFY_TO`; every sample used in the models has a curve that reaches it.

```{r rarefaction, fig.width = 10, fig.height = 5}
rare_curve <- bind_rows(lapply(split(md_v, md_v$sample_origin), function(d) {
  cl <- d$CTstrict[!is.na(d$CTstrict)]
  if (!length(cl)) return(NULL)
  depths <- unique(pmin(c(seq(50, length(cl), by = 100), length(cl)), length(cl)))
  bind_rows(lapply(depths, function(k) {
    r <- mean(replicate(20, length(unique(sample(cl, k, replace = FALSE)))))
    data.frame(sample_origin = d$sample_origin[1], depth = k, richness = r)
  }))
})) %>% left_join(sample_meta, by = "sample_origin")

p_rarecurve <- ggplot(rare_curve, aes(depth, richness,
                                      group = category_plot, colour = condition)) +
  geom_line(alpha = 0.8, linewidth = 0.5) +
  geom_vline(xintercept = RAREFY_TO, linetype = "dashed", colour = "grey40") +
  facet_grid(~ condition) +
  scale_colour_manual(values = PAL_COND, guide = "none") +
  labs(x = "Cells sampled", y = "Unique clonotypes observed",
       title = paste0(SUBSET_TITLE, ": rarefaction curves"),
       subtitle = sprintf("One line per sample. Dashed line is the rarefaction depth (%d cells).",
                          RAREFY_TO)) +
  theme_tcr() + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
p_rarecurve
```

# Per sample repertoire metrics

Clones are counted among the cells of this subset only, so the denominator is T
cells that carried a recoverable TCR. Raw and rarefied are both written: raw
describes the sample as sequenced, rarefied is what the models are fitted on.
`trb_*` repeats the rarefied metrics on the beta-only clone id, which is the
sensitivity check against alpha-chain dropout.

```{r sample-metrics}
raw_metrics <- bind_rows(lapply(split(md_v$CTstrict, md_v$sample_origin), div_metrics),
                         .id = "sample_origin")

usable <- raw_metrics$sample_origin[raw_metrics$n_cells >= MIN_VDJ_CELLS]
depth  <- if (is.na(RAREFY_TO))
  min(raw_metrics$n_cells[raw_metrics$sample_origin %in% usable]) else RAREFY_TO
# RAREFY_TO used to equal MIN_VDJ_CELLS, so a threshold sample carried zero
# subsampling variance while a 5,000-cell sample got a real subsample.
stopifnot(depth <= MIN_VDJ_CELLS)
cat("Samples usable:", length(usable), "of", nrow(raw_metrics),
    "| rarefaction depth:", depth, "cells\n")
if (length(setdiff(raw_metrics$sample_origin, usable)))
  cat("Dropped below MIN_VDJ_CELLS:",
      paste(cat_of[setdiff(raw_metrics$sample_origin, usable)], collapse = ", "), "\n")

rare_metrics <- bind_rows(lapply(usable, function(s) {
  d <- div_rarefied(md_v$CTstrict[md_v$sample_origin == s], depth)
  if (is.null(d)) NULL else cbind(sample_origin = s, d)
}))
rare_trb <- bind_rows(lapply(usable, function(s) {
  d <- div_rarefied(md_v$clone_trb[md_v$sample_origin == s], depth)
  if (is.null(d)) NULL else cbind(sample_origin = s, d)
}))

# recovery and dropout recomputed from the object, so this notebook needs no contigs
recovery <- md %>%
  group_by(sample_origin) %>%
  summarise(n_cells_subset = n(),
            vdj_recovery_pct = round(100 * mean(has_vdj), 2), .groups = "drop")
dropout <- md %>% filter(has_vdj) %>%
  group_by(sample_origin) %>%
  summarise(pct_paired    = round(100 * mean(chain_status == "paired"), 2),
            pct_beta_only = round(100 * mean(chain_status == "beta_only"), 2),
            .groups = "drop") %>%
  mutate(dropout_flag = pct_beta_only > DROPOUT_FLAG_PCT)

sample_metrics <- raw_metrics %>%
  rename_with(~ paste0("raw_", .x), -sample_origin) %>%
  left_join(rare_metrics %>% rename_with(~ paste0("rare_", .x), -sample_origin),
            by = "sample_origin") %>%
  left_join(rare_trb %>% rename_with(~ paste0("trb_", .x), -sample_origin),
            by = "sample_origin") %>%
  left_join(recovery, by = "sample_origin") %>%
  left_join(dropout,  by = "sample_origin") %>%
  left_join(sample_meta, by = "sample_origin") %>%
  left_join(patient_visits %>% dplyr::select(patient_id, n_visits_seq, single_visit),
            by = "patient_id") %>%
  mutate(usable = sample_origin %in% usable) %>%
  dplyr::select(category_plot, sample_origin, category, patient_id, condition, timepoint,
                everything()) %>%
  arrange(category_plot)

print(as.data.frame(sample_metrics %>%
  dplyr::select(category_plot, raw_n_cells, raw_n_clones, raw_pct_expanded,
                rare_shannon, rare_clonality, rare_gini, trb_clonality, usable)))

sens <- sample_metrics %>% filter(usable, is.finite(rare_clonality), is.finite(trb_clonality))
cat(sprintf("Spearman rho, CTstrict vs beta-only clonality: %.3f (n = %d samples)\n",
            suppressWarnings(cor(sens$rare_clonality, sens$trb_clonality, method = "spearman")),
            nrow(sens)))
```

# Renyi entropy profiles

Shannon is the single point at `alpha = 1`. The profile says which part of the
distribution any difference lives in: `alpha = 0` is log richness and is driven
by singletons, `alpha = 1` weights clones by their frequency, and large `alpha`
is dominated by the single biggest clone. Two repertoires whose curves cross are
not orderable by any single diversity index, which is worth knowing before
reporting one.

Computed on the same rarefactions as everything else, so the curves are
comparable across samples of different depth.

```{r renyi, fig.width = 11, fig.height = 5}
renyi_tbl <- bind_rows(lapply(usable, function(s) {
  d <- renyi_rarefied(md_v$CTstrict[md_v$sample_origin == s], depth)
  if (is.null(d)) NULL else cbind(sample_origin = s, d)
})) %>% left_join(sample_meta, by = "sample_origin")

# a finite stand-in for Inf so the alpha axis can be drawn
alpha_levels <- c(as.character(RENYI_ALPHA[is.finite(RENYI_ALPHA)]), "Inf")
renyi_tbl <- renyi_tbl %>%
  mutate(alpha_lab = factor(ifelse(is.infinite(alpha), "Inf", as.character(alpha)),
                            levels = alpha_levels))

p_renyi <- ggplot(renyi_tbl, aes(alpha_lab, renyi, group = category_plot,
                                 colour = condition)) +
  geom_line(alpha = 0.8, linewidth = 0.5) + geom_point(size = 1) +
  facet_grid(~ condition) +
  geom_vline(xintercept = which(alpha_levels == "1"),
             linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = PAL_COND, guide = "none") +
  labs(x = expression(alpha), y = "Renyi entropy",
       title = paste0(SUBSET_TITLE, ": Renyi entropy profiles"),
       subtitle = "One line per sample, rarefied. Dashed line at alpha = 1 is Shannon entropy.") +
  theme_tcr() + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
p_renyi
```

```{r renyi-tests}
# a Wilcoxon at each alpha, reported the same way as every other condition test
renyi_tests <- bind_rows(lapply(RENYI_ALPHA, function(a) {
  d <- renyi_tbl %>% filter(alpha == a) %>%
    group_by(category, patient_id, condition) %>%
    summarise(value = mean(renyi, na.rm = TRUE), .groups = "drop")
  x <- d$value[d$condition == "CASE"];   x <- x[is.finite(x)]
  y <- d$value[d$condition == "CTRL"]; y <- y[is.finite(y)]
  if (min(length(x), length(y)) < MIN_UNITS_PER_GROUP) {
    log_row("renyi", "patient", paste0("alpha_", a), length(x), length(y),
            NA_integer_, "skipped")
    return(NULL)
  }
  w     <- suppressWarnings(wilcox.test(x, y))
  min_p <- min(2 / choose(length(x) + length(y), min(length(x), length(y))), 1)
  log_row("renyi", "patient", paste0("alpha_", a), length(x), length(y),
          length(x) + length(y), if (min(length(x), length(y)) < 3) "exploratory" else "ok")
  data.frame(alpha = a, n_CASE = length(x), n_CTRL = length(y),
             mean_CASE = mean(x), mean_CTRL = mean(y), diff = mean(x) - mean(y),
             rank_biserial = rb_effect(x, y), p_val = w$p.value,
             min_possible_p = min_p, powered = min_p <= 0.05)
}))
if (!is.null(renyi_tests) && nrow(renyi_tests)) {
  renyi_tests$p_val_adj <- p.adjust(renyi_tests$p_val, method = "BH")
  print(as.data.frame(renyi_tests))
}
```

# Per state repertoire metrics

```{r state-metrics}
state_metrics <- md_v %>%
  group_by(sample_origin, cell_label) %>%
  filter(n() >= MIN_CELLS_UNIT) %>%
  group_modify(~ div_metrics(.x$CTstrict)) %>%
  ungroup() %>%
  left_join(sample_meta, by = "sample_origin") %>%
  dplyr::select(category_plot, sample_origin, cell_label, condition, timepoint, everything()) %>%
  arrange(category_plot, cell_label)
```

```{r fig-gini, fig.width = 9, fig.height = 5}
# Nat Med 2023 Fig 3f: Gini as the headline clonality number, box + points,
# split by group. Ours keeps the sample as the point and the arm as the box.
gini_plot <- sample_metrics %>% filter(usable, is.finite(rare_gini))
p_gini <- ggplot(gini_plot, aes(condition, rare_gini, colour = condition)) +
  geom_boxplot(aes(fill = condition), alpha = 0.15, outlier.shape = NA,
               linewidth = 0.5, show.legend = FALSE) +
  geom_point(position = position_jitter(width = 0.12), size = 2, alpha = 0.85) +
  facet_grid(~ timepoint) +
  scale_colour_manual(values = PAL_COND, guide = "none") +
  scale_fill_manual(values = PAL_COND, guide = "none") +
  labs(x = NULL, y = "Gini index (rarefied)",
       title = paste0(SUBSET_TITLE, ": clonal inequality"),
       subtitle = "One point per sample. 0 = all clones equal, 1 = one clone holds everything.") +
  theme_tcr()
p_gini
```

# Tracking clones across timepoints within a patient

Overlap is computed inside a patient only. Morisita is the one to report: it is
abundance weighted, where Jaccard is dominated by singletons and can read 0.99
for two samples that share none of their dominant clones. A clone is persistent
when it is seen at two or more visits of the same patient.

```{r overlap}
overlap_tp <- bind_rows(lapply(unique(md_v$patient_id), function(pt) {
  d   <- md_v[md_v$patient_id == pt, ]
  tps <- intersect(TP_LEVELS, unique(as.character(d$timepoint)))
  if (length(tps) < 2) return(NULL)
  cmb <- t(combn(tps, 2))
  bind_rows(lapply(seq_len(nrow(cmb)), function(k) {
    a  <- d$CTstrict[d$timepoint == cmb[k, 1]]
    b  <- d$CTstrict[d$timepoint == cmb[k, 2]]
    at <- d$clone_trb[d$timepoint == cmb[k, 1]]
    bt <- d$clone_trb[d$timepoint == cmb[k, 2]]
    if (length(a) < MIN_CELLS_UNIT || length(b) < MIN_CELLS_UNIT) return(NULL)
    sa <- unique(a); sb <- unique(b)
    data.frame(category   = as.character(d$category[1]), patient_id = pt,
               condition  = as.character(d$condition[1]),
               tp_1 = cmb[k, 1], tp_2 = cmb[k, 2],
               n_cells_1 = length(a), n_cells_2 = length(b),
               n_clones_1 = length(sa), n_clones_2 = length(sb),
               n_shared = length(intersect(sa, sb)),
               jaccard  = length(intersect(sa, sb)) / length(union(sa, sb)),
               morisita = morisita(a, b),
               morisita_trb = morisita(at, bt),
               stringsAsFactors = FALSE)
  }))
})) %>% arrange(category, tp_1, tp_2)
print(as.data.frame(overlap_tp %>% dplyr::select(-patient_id, -morisita_trb)))
```

## Clone-count biplots between consecutive visits

Cell Rep Methods Fig 2C. Every clone of a patient plotted at its size in the
first visit against its size in the second. Shared clones sit on the diagonal,
clones private to one visit lie along an axis. It answers, per clone, what
Morisita answers as one number, and it shows whether an overlap value is carried
by a few large clones or by a long tail of singletons.

Sizes are on a log axis with a half-cell offset so absent clones can be drawn.

```{r fig-biplot, fig.width = 12, fig.height = 7}
consec <- overlap_tp %>%
  group_by(patient_id) %>%
  filter(match(tp_2, TP_LEVELS) == match(tp_1, TP_LEVELS) + 1 |
           row_number() == 1) %>%   # consecutive pairs, or the only pair
  ungroup()

biplot_dat <- bind_rows(lapply(seq_len(nrow(consec)), function(k) {
  r <- consec[k, ]
  d <- md_v %>% filter(patient_id == r$patient_id,
                       timepoint %in% c(r$tp_1, r$tp_2))
  w <- d %>% dplyr::count(CTstrict, timepoint, name = "n") %>%
    tidyr::pivot_wider(names_from = timepoint, values_from = n, values_fill = 0)
  if (!all(c(r$tp_1, r$tp_2) %in% names(w))) return(NULL)
  data.frame(category = r$category, condition = r$condition,
             pair = paste(r$tp_1, r$tp_2, sep = " -> "),
             x = w[[r$tp_1]], y = w[[r$tp_2]],
             stringsAsFactors = FALSE)
})) %>%
  mutate(status = ifelse(x > 0 & y > 0, "shared",
                  ifelse(x > 0, "first visit only", "second visit only")),
         mean_n = (x + y) / 2)

p_biplot <- ggplot(biplot_dat, aes(x + 0.5, y + 0.5, colour = log2(mean_n + 1))) +
  geom_abline(slope = 1, intercept = 0, colour = "grey70", linetype = "dashed") +
  geom_point(alpha = 0.6, size = 1.2) +
  facet_wrap(category ~ pair, scales = "free") +
  scale_x_log10() + scale_y_log10() +
  scale_colour_gradientn(colours = PAL_SEQ) +
  labs(x = "Clone size, earlier visit", y = "Clone size, later visit",
       colour = "log2(mean count)",
       title = paste0(SUBSET_TITLE, ": clone sizes between consecutive visits"),
       subtitle = "Shared clones lie on the diagonal, private clones along the axes. 0.5 added to allow the log axis.") +
  theme_tcr() + theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
                      strip.text = element_text(size = rel(0.7)))
p_biplot
```

```{r persistence}
persist <- md_v %>%
  distinct(patient_id, condition, CTstrict, timepoint) %>%
  dplyr::count(patient_id, condition, CTstrict, name = "n_visits")

persist_summary <- persist %>%
  group_by(patient_id, condition) %>%
  summarise(n_clones = n(), n_persistent = sum(n_visits >= 2),
            pct_persistent = round(100 * mean(n_visits >= 2), 2),
            max_visits = max(n_visits), .groups = "drop") %>%
  left_join(patient_visits, by = c("patient_id", "condition")) %>%
  mutate(pct_persistent = ifelse(single_visit, NA_real_, pct_persistent)) %>%
  dplyr::select(category, patient_id, condition, everything()) %>%
  arrange(category)
print(as.data.frame(persist_summary %>% dplyr::select(-patient_id)))

persistent_clones <- md_v %>%
  inner_join(persist %>% filter(n_visits >= 2) %>% dplyr::select(patient_id, CTstrict),
             by = c("patient_id", "CTstrict")) %>%
  group_by(category, patient_id, condition, CTstrict) %>%
  summarise(n_cells = n(),
            pct_of_patient = round(100 * dplyr::first(patient_clone_prop), 3),
            visits = paste(sort(unique(as.character(timepoint))), collapse = ","),
            states = paste(sort(unique(as.character(cell_label))), collapse = ","),
            CTaa   = names(sort(table(CTaa), decreasing = TRUE))[1],
            .groups = "drop") %>%
  arrange(category, desc(n_cells))
```

```{r tracking-plots, fig.width = 10, fig.height = 6}
if (nrow(overlap_tp)) {
  p_overlap <- ggplot(overlap_tp, aes(paste(tp_1, tp_2, sep = " vs "), morisita,
                                      colour = condition)) +
    geom_point(position = position_jitter(width = 0.12), size = 2) +
    ggrepel::geom_text_repel(aes(label = category), size = 2.5, max.overlaps = 20,
                             show.legend = FALSE) +
    facet_grid(~ condition) +
    scale_colour_manual(values = PAL_COND, guide = "none") +
    labs(x = NULL, y = "Morisita overlap",
         title = paste0(SUBSET_TITLE, ": repertoire overlap between visits, within patient")) +
    theme_tcr()
  print(p_overlap)
} else p_overlap <- NULL

# top clone trajectories per patient, the clearest single view of expansion and contraction
traj <- md_v %>%
  dplyr::count(category, patient_id, condition, timepoint, CTstrict, name = "n") %>%
  group_by(patient_id, timepoint) %>% mutate(prop = n / sum(n)) %>%
  group_by(patient_id, CTstrict) %>% mutate(peak = max(prop)) %>% ungroup()
top_clones <- traj %>% distinct(patient_id, CTstrict, peak) %>%
  group_by(patient_id) %>% slice_max(peak, n = 10, with_ties = FALSE) %>% ungroup()
traj_top <- traj %>% semi_join(top_clones, by = c("patient_id", "CTstrict"))
p_traj <- if (nrow(traj_top))
  ggplot(traj_top, aes(timepoint, 100 * prop, group = CTstrict, colour = CTstrict)) +
    geom_line(alpha = 0.7) + geom_point(size = 1) +
    facet_wrap(~ category, scales = "free_y") +
    labs(x = NULL, y = "% of TCR+ cells at that visit",
         title = paste0(SUBSET_TITLE, ": top 10 clones per patient")) +
    theme_tcr() + theme(legend.position = "none",
                        axis.text.x = element_text(angle = 0, hjust = 0.5)) else NULL
if (!is.null(p_traj)) print(p_traj)
```

# Clones shared between states

A clone found in two states is one lineage occupying both, which is the
differentiation question rather than the expansion question. Sharing is counted
inside a patient and then averaged over patients.

Two indices are written. `overlap_coef` (shared / smaller set) is symmetric and
is what the previous version plotted. `tversky` is the asymmetric containment
used in Nat Med 2023 Fig 3c: the value in row *i*, column *j* is the fraction of
state *i*'s clones that also appear in state *j*. Asymmetry is the point. A small
effector state can be almost entirely contained within a large memory state
while the reverse is nearly zero, and a symmetric index averages that away.

```{r state-sharing}
state_sharing <- bind_rows(lapply(unique(md_v$patient_id), function(pt) {
  d  <- md_v[md_v$patient_id == pt, ]
  st <- sort(unique(as.character(d$cell_label)))
  st <- st[vapply(st, function(s) sum(d$cell_label == s) >= MIN_CELLS_UNIT, logical(1))]
  if (length(st) < 2) return(NULL)
  # all ordered pairs, because tversky is not symmetric
  grd <- expand.grid(state_1 = st, state_2 = st, stringsAsFactors = FALSE) %>%
    filter(state_1 != state_2)
  bind_rows(lapply(seq_len(nrow(grd)), function(k) {
    a <- unique(d$CTstrict[d$cell_label == grd$state_1[k]])
    b <- unique(d$CTstrict[d$cell_label == grd$state_2[k]])
    data.frame(category = as.character(d$category[1]), patient_id = pt,
               condition = as.character(d$condition[1]),
               state_1 = grd$state_1[k], state_2 = grd$state_2[k],
               n_clones_1 = length(a), n_clones_2 = length(b),
               n_shared = length(intersect(a, b)),
               overlap_coef = length(intersect(a, b)) / min(length(a), length(b)),
               tversky = tversky(a, b),
               stringsAsFactors = FALSE)
  }))
})) %>% arrange(category, state_1, state_2)

state_sharing_mean <- state_sharing %>%
  group_by(condition, state_1, state_2) %>%
  summarise(n_patients = n(),
            mean_overlap = round(mean(overlap_coef, na.rm = TRUE), 3),
            mean_tversky = round(mean(tversky, na.rm = TRUE), 3), .groups = "drop")
```

```{r fig-tversky, fig.width = 11, fig.height = 5.5}
# Column order from clustering the mean Tversky matrix, so related states sit
# together instead of alphabetically. Diagonal masked, as in Nat Med 2023.
sq <- state_sharing_mean %>%
  group_by(state_1, state_2) %>%
  summarise(v = mean(mean_tversky, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = state_2, values_from = v) %>%
  as.data.frame()
rownames(sq) <- sq$state_1; sq$state_1 <- NULL
sq <- as.matrix(sq[, rownames(sq), drop = FALSE])
sq[is.na(sq)] <- 0
ord <- if (nrow(sq) > 2)
  rownames(sq)[stats::hclust(stats::dist(sq), method = "ward.D2")$order] else rownames(sq)

tv <- state_sharing_mean %>%
  mutate(state_1 = factor(state_1, levels = ord),
         state_2 = factor(state_2, levels = ord))

p_tversky <- ggplot(tv, aes(state_2, state_1, fill = mean_tversky)) +
  geom_tile(colour = "grey90", linewidth = 0.2) +
  facet_grid(~ condition) +
  scale_fill_gradientn(colours = c("#FFFFFF", "#FFF3B0", "#F0A03C", "#C1272D",
                                   "#7A1E5C", "#1A1033"),
                       na.value = "grey85") +
  coord_equal() +
  labs(x = "found in state", y = "fraction of this state's clones",
       fill = "Tversky index",
       title = paste0(SUBSET_TITLE, ": clonal sharing between states"),
       subtitle = "Asymmetric: row i, column j is the share of state i's clones also seen in state j. Diagonal excluded.") +
  theme_tcr() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = rel(0.75)),
        axis.text.y = element_text(size = rel(0.75)),
        panel.grid = element_blank())
p_tversky
```

```{r startrac}
# subset(), not obj[, cells]: the bracket form drops reductions in Seurat v5.
# Startrac does not need one, but obj_plot should behave the same way here as in
# the main notebook.
obj_plot <- subset(obj, cells = colnames(obj)[!obj$cell_label %in% drop_labels])
Idents(obj_plot) <- factor(obj_plot$cell_label)

startrac <- safe(StartracDiversity(obj_plot, type = "timepoint", group.by = "patient_id",
                                   exportTable = TRUE),
                 "startrac", "all", "startrac_indices")
if (!is.null(startrac)) {
  startrac <- as.data.frame(startrac)
  if ("group" %in% names(startrac))
    startrac <- startrac %>%
      left_join(patient_meta %>% dplyr::select(patient_id, category),
                by = c("group" = "patient_id")) %>%
      dplyr::select(category, everything())
  log_row("startrac", "all", "startrac_indices", NA_integer_, NA_integer_, nrow(startrac), "ok")
  if ("majorCluster" %in% names(startrac))
    cat("Startrac majorCluster values:",
        paste(head(unique(startrac$majorCluster), 5), collapse = ", "), "\n")
}
```

# Public clones

The same CDR3 amino acid sequence in more than one patient is convergence, not
shared lineage, and is reported separately from anything tracked. `CTaa` is used
here and nowhere else.

```{r public}
public_clones <- md_v %>%
  filter(!is.na(CTaa)) %>%
  distinct(patient_id, CTaa) %>%
  dplyr::count(CTaa, name = "n_patients") %>%
  filter(n_patients >= 2) %>%
  left_join(md_v %>% filter(!is.na(CTaa)) %>%
              group_by(CTaa) %>%
              summarise(n_cells = n(),
                        patients   = paste(sort(unique(as.character(category))), collapse = ","),
                        conditions = paste(sort(unique(as.character(condition))), collapse = ","),
                        batches    = paste(sort(unique(batch)), collapse = ","),
                        states     = paste(sort(unique(as.character(cell_label))), collapse = ","),
                        .groups = "drop"), by = "CTaa") %>%
  mutate(same_batch = !grepl(",", batches)) %>%
  arrange(desc(n_patients), desc(n_cells))
cat("Public CDR3aa clonotypes (>=2 patients):", nrow(public_clones),
    "| confined to one batch:", sum(public_clones$same_batch), "\n")

# if sharing concentrates in one patient pair AND that pair sits in one
# sequencing batch, index hopping is more likely than convergent selection.
pat_batch <- sample_meta %>% group_by(patient_id, category) %>%
  summarise(batches = paste(sort(unique(batch)), collapse = ","), .groups = "drop")
public_pairs <- md_v %>%
  filter(!is.na(CTaa)) %>%
  distinct(patient_id, CTaa) %>%
  inner_join(., ., by = "CTaa", relationship = "many-to-many") %>%
  filter(patient_id.x < patient_id.y) %>%
  dplyr::count(patient_id.x, patient_id.y, name = "n_shared_aa") %>%
  left_join(pat_batch, by = c("patient_id.x" = "patient_id")) %>%
  rename(category_x = category, batches_x = batches) %>%
  left_join(pat_batch, by = c("patient_id.y" = "patient_id")) %>%
  rename(category_y = category, batches_y = batches) %>%
  mutate(shared_batch = batches_x == batches_y) %>%
  dplyr::select(category_x, category_y, n_shared_aa, batches_x, batches_y, shared_batch,
                patient_id.x, patient_id.y) %>%
  arrange(desc(n_shared_aa))
print(as.data.frame(head(public_pairs, 10)))
```

# Family 1: CASE vs CTRL, patient level

Timepoints of a patient are averaged into one number so the unit is the patient,
exactly as counts are pooled per patient in the DEG notebook. A sensitivity fit
excluding single-visit patients runs alongside.

```{r family1}
METRICS <- c("rare_shannon", "rare_eff_clones", "rare_inv_simpson", "rare_norm_entropy",
             "rare_clonality", "rare_gini", "rare_n_clones", "rare_pct_expanded",
             "rare_top10_pct")
sm <- sample_metrics %>% filter(usable)

pat_level <- sm %>%
  group_by(category, patient_id, condition) %>%
  summarise(across(all_of(METRICS), ~ mean(.x, na.rm = TRUE)),
            n_samples = n(), .groups = "drop") %>%
  left_join(patient_visits %>% dplyr::select(patient_id, single_visit), by = "patient_id") %>%
  arrange(category)

test_condition <- function(dat, tag) {
  bind_rows(lapply(METRICS, function(m) {
    x <- dat[[m]][dat$condition == "CASE"];   x <- x[is.finite(x)]
    y <- dat[[m]][dat$condition == "CTRL"]; y <- y[is.finite(y)]
    if (min(length(x), length(y)) < MIN_UNITS_PER_GROUP) {
      log_row("condition", tag, m, length(x), length(y), NA_integer_, "skipped")
      return(NULL)
    }
    w <- suppressWarnings(wilcox.test(x, y))
    min_p <- min(2 / choose(length(x) + length(y), min(length(x), length(y))), 1)
    log_row("condition", tag, m, length(x), length(y), length(x) + length(y),
            if (min(length(x), length(y)) < 3) "exploratory" else "ok")
    data.frame(analysis = tag, metric = m, n_CASE = length(x), n_CTRL = length(y),
               mean_CASE = mean(x), mean_CTRL = mean(y), diff = mean(x) - mean(y),
               rank_biserial = rb_effect(x, y), p_val = w$p.value,
               min_possible_p = min_p, powered = min_p <= 0.05,
               stringsAsFactors = FALSE)
  }))
}

res_cond <- bind_rows(
  test_condition(pat_level, "all_patients"),
  test_condition(pat_level %>% filter(!single_visit), "multi_visit_only")) %>%
  group_by(analysis) %>% mutate(p_val_adj = p.adjust(p_val, method = "BH")) %>% ungroup()
print(as.data.frame(res_cond))
```

# Family 2: timepoint trajectories within a condition, paired

Timepoint varies inside a patient, so the sample is the unit and the patient is a
random intercept. `lmer(metric ~ timepoint + (1 | patient_id))` is the paired
equivalent of the `~ patient_id + timepoint` design used for counts, fitted
inside each condition separately. `TP_LEVELS` sets the reference, so every
`timepointD14` coefficient means "T1 minus T0".

```{r family2}
fit_lmm <- function(d, form, family, unit, item) {
  fit <- tryCatch(suppressMessages(lmerTest::lmer(form, data = d)), error = function(e) NULL)
  if (is.null(fit)) {
    log_row(family, unit, item, NA_integer_, NA_integer_, nrow(d), "skipped", "fit failed")
    return(NULL)
  }
  cf <- as.data.frame(coef(summary(fit)))
  cf$term <- rownames(cf); rownames(cf) <- NULL
  names(cf)[names(cf) == "Pr(>|t|)"]  <- "p_val"
  names(cf)[names(cf) == "Std. Error"] <- "se"
  names(cf)[names(cf) == "Estimate"]   <- "estimate"
  log_row(family, unit, item, NA_integer_, NA_integer_, nrow(d),
          if (length(unique(d$patient_id)) < 3) "exploratory" else "ok")
  cf %>% filter(term != "(Intercept)") %>%
    dplyr::select(term, estimate, se, p_val) %>%
    mutate(item = item, n_patients = length(unique(d$patient_id)), .before = 1)
}

res_within <- bind_rows(lapply(COND_LEVELS, function(cnd) {
  d0 <- sm %>% filter(condition == cnd)
  bind_rows(lapply(METRICS, function(m) {
    d <- d0 %>% dplyr::select(patient_id, timepoint, value = all_of(m)) %>%
      filter(is.finite(value))
    d$timepoint <- droplevels(factor(as.character(d$timepoint), levels = TP_LEVELS))
    keep <- names(which(table(d$patient_id) >= 2))   # one visit carries no within-patient info
    d <- d %>% filter(patient_id %in% keep)
    if (nlevels(d$timepoint) < 2 || length(unique(d$patient_id)) < MIN_UNITS_PER_GROUP) {
      log_row("within_condition_timepoint", cnd, m, NA_integer_, NA_integer_, nrow(d), "skipped")
      return(NULL)
    }
    out <- fit_lmm(d, value ~ timepoint + (1 | patient_id),
                   "within_condition_timepoint", cnd, m)
    if (is.null(out)) return(NULL)
    cbind(condition = cnd, reference = levels(d$timepoint)[1], out)
  }))
}))
if (!is.null(res_within) && nrow(res_within)) {
  res_within$p_val_adj <- p.adjust(res_within$p_val, method = "BH")
  print(as.data.frame(res_within))
}
```

# Family 3: CASE vs CTRL at each timepoint, unpaired

```{r family3}
res_between <- bind_rows(lapply(TP_LEVELS, function(tp) {
  d0 <- sm %>% filter(timepoint == tp)
  bind_rows(lapply(METRICS, function(m) {
    x <- d0[[m]][d0$condition == "CASE"];   x <- x[is.finite(x)]
    y <- d0[[m]][d0$condition == "CTRL"]; y <- y[is.finite(y)]
    if (min(length(x), length(y)) < MIN_UNITS_PER_GROUP) {
      log_row("between_condition_timepoint", tp, m, length(x), length(y),
              length(x) + length(y), "skipped")
      return(NULL)
    }
    w <- suppressWarnings(wilcox.test(x, y))
    min_p <- min(2 / choose(length(x) + length(y), min(length(x), length(y))), 1)
    log_row("between_condition_timepoint", tp, m, length(x), length(y),
            length(x) + length(y), if (min(length(x), length(y)) < 3) "exploratory" else "ok")
    data.frame(timepoint = tp, metric = m, n_CASE = length(x), n_CTRL = length(y),
               mean_CASE = mean(x), mean_CTRL = mean(y), diff = mean(x) - mean(y),
               rank_biserial = rb_effect(x, y), p_val = w$p.value,
               min_possible_p = min_p, powered = min_p <= 0.05,
               stringsAsFactors = FALSE)
  }))
}))
if (!is.null(res_between) && nrow(res_between)) {
  res_between$p_val_adj <- p.adjust(res_between$p_val, method = "BH")
  print(as.data.frame(res_between))
}
```

# Family 4: condition x timepoint interaction

Whether the CASE repertoire *changes differently* from CTRL. Patient is nested
inside condition, so it enters as a random intercept rather than a fixed effect,
for the same reason as in the DEG interaction model. A timepoint is included
only if both arms have at least `MIN_UNITS_PER_GROUP` patients there.

```{r family4}
pat_tp <- sm %>% distinct(patient_id, condition, timepoint) %>%
  dplyr::count(condition, timepoint, name = "n_pat")
tp_intx <- intersect(TP_LEVELS, Reduce(intersect, lapply(COND_LEVELS, function(cnd)
  as.character(pat_tp$timepoint[pat_tp$condition == cnd & pat_tp$n_pat >= MIN_UNITS_PER_GROUP]))))
cat("Timepoints in the interaction model:", paste(tp_intx, collapse = ", "), "\n")

res_intx <- if (length(tp_intx) >= 2) bind_rows(lapply(METRICS, function(m) {
  d <- sm %>% filter(timepoint %in% tp_intx) %>%
    dplyr::select(patient_id, condition, timepoint, value = all_of(m)) %>%
    filter(is.finite(value))
  d$timepoint <- droplevels(factor(as.character(d$timepoint), levels = tp_intx))
  d$condition <- droplevels(factor(as.character(d$condition), levels = COND_LEVELS))
  if (nlevels(d$timepoint) < 2 || nlevels(d$condition) < 2 ||
      length(unique(d$patient_id)) < 3) {
    log_row("condition_by_timepoint_interaction", "all", m, NA_integer_, NA_integer_,
            nrow(d), "skipped")
    return(NULL)
  }
  out <- fit_lmm(d, value ~ condition * timepoint + (1 | patient_id),
                 "condition_by_timepoint_interaction", "all", m)
  if (is.null(out)) return(NULL)
  cbind(reference = levels(d$timepoint)[1], out)
})) else NULL

if (!is.null(res_intx) && nrow(res_intx)) {
  res_intx$p_val_adj <- p.adjust(res_intx$p_val, method = "BH")
  res_intx$is_interaction <- grepl(":", res_intx$term)
  print(as.data.frame(res_intx %>% filter(is_interaction)))
}
```

```{r metric-plots, fig.width = 11, fig.height = 6}
long <- sm %>%
  dplyr::select(category_plot, category, patient_id, condition, timepoint, all_of(METRICS)) %>%
  pivot_longer(all_of(METRICS), names_to = "metric", values_to = "value")
p_metric_traj <- ggplot(long, aes(timepoint, value, group = category, colour = condition)) +
  geom_line(alpha = 0.6) + geom_point(size = 1.5) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_colour_manual(values = PAL_COND) +
  labs(x = NULL, y = NULL, colour = NULL,
       title = paste0(SUBSET_TITLE, ": rarefied repertoire metrics")) +
  theme_tcr() + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
print(p_metric_traj)

p_metric_box <- ggplot(long, aes(condition, value, colour = condition)) +
  geom_boxplot(aes(fill = condition), alpha = 0.15, outlier.shape = NA, linewidth = 0.4) +
  geom_point(position = position_jitter(width = 0.15), size = 1.5, alpha = 0.85) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_colour_manual(values = PAL_COND, guide = "none") +
  scale_fill_manual(values = PAL_COND, guide = "none") +
  labs(x = NULL, y = NULL, title = paste0(SUBSET_TITLE, ": metrics by condition, all samples")) +
  theme_tcr()
print(p_metric_box)
```

# Write the figures

```{r figures-pdf, results = FALSE}
pdf(file.path(fig_dir, "subset_data_diversity.pdf"), width = 11, height = 6)
print(p_rarecurve); print(p_renyi); print(p_gini)
print(p_metric_traj); print(p_metric_box)
invisible(dev.off())

pdf(file.path(fig_dir, "subset_data_clonal_tracking.pdf"), width = 12, height = 7)
if (!is.null(p_overlap)) print(p_overlap)
if (!is.null(p_traj))    print(p_traj)
print(p_biplot)
invisible(dev.off())

pdf(file.path(fig_dir, "subset_data_clone_sharing.pdf"), width = 11, height = 6)
print(p_tversky)
invisible(dev.off())
```

# Write the workbook

Every table is listed once, here, after all of it has been computed. Nothing
accumulates in a global as the notebook runs, so re-running a chunk cannot
corrupt the workbook. Analyses that were skipped leave a NULL and are reported
by name rather than silently vanishing; the manifest is still the full record.

```{r write}
manifest <- bind_rows(MANIFEST)

sheet <- function(d, desc) {
  if (is.null(d)) return(NULL)
  d <- as.data.frame(d)
  if (!nrow(d)) return(NULL)
  list(d = d, desc = desc)
}

BOOK <- list(
  rarefaction_curves = sheet(
    rare_curve %>% mutate(richness = round(richness, 1)) %>%
      dplyr::select(category_plot, sample_origin, condition, timepoint, depth, richness),
    "Mean observed clonotype richness at each subsampling depth, 20 draws per point."),
  sample_metrics = sheet(sample_metrics,
    "One row per sample. raw_* = as sequenced, rare_* = rarefied (the models use these), trb_* = beta-only sensitivity."),
  beta_only_sensitivity = sheet(
    sens %>% dplyr::select(category_plot, sample_origin, condition, timepoint,
                           pct_beta_only, rare_n_clones, trb_n_clones,
                           rare_clonality, trb_clonality),
    "The two clone calls side by side. Large disagreement means alpha dropout is driving a result."),
  renyi_profiles = sheet(
    renyi_tbl %>% mutate(renyi = round(renyi, 4), renyi_sd = round(renyi_sd, 4)) %>%
      dplyr::select(category_plot, sample_origin, condition, timepoint,
                    alpha, renyi, renyi_sd),
    "Rarefied Renyi entropy at each alpha, per sample. alpha=0 is log richness, alpha=1 is Shannon, alpha=Inf is -log of the largest clone's share."),
  renyi_condition_tests = sheet(renyi_tests,
    "CASE vs CTRL at each Renyi alpha, patient as the unit. alpha=1 is the Shannon comparison."),
  state_metrics_by_sample = sheet(state_metrics,
    "Diversity within each cell state of each sample. Not rarefied, so compare within a sample rather than across."),
  overlap_between_visits = sheet(overlap_tp,
    "Repertoire overlap between two visits of the same patient. Report morisita; morisita_trb is the beta-only check."),
  clone_sizes_between_visits = sheet(
    biplot_dat %>% dplyr::select(category, condition, pair, x, y, status),
    "Every clone's size at two consecutive visits of one patient. The data behind the biplots."),
  persistence_by_patient = sheet(persist_summary,
    "Percent of a patient's clones seen at 2+ visits. NA where the patient has one sequenced visit: a structural zero, not a finding."),
  persistent_clones = sheet(persistent_clones,
    "Every clone seen at 2+ visits of one patient, with its size, visits and states."),
  top_clone_trajectories = sheet(traj_top,
    "Per-visit frequency of each patient's ten largest clones. The data behind the trajectory figure."),
  sharing_between_states = sheet(state_sharing,
    "Clones shared between two cell states within a patient. overlap_coef is symmetric; tversky is the fraction of state_1's clones found in state_2."),
  sharing_states_mean = sheet(state_sharing_mean,
    "The same matrices averaged over patients. This is the version plotted."),
  startrac_indices = sheet(startrac,
    "expa = expansion, migr = cross-timepoint movement, tran = cross-state transition, per state per patient."),
  public_clones_aa = sheet(public_clones,
    "CDR3aa shared by 2+ patients. same_batch TRUE means every cell came from one sequencing batch, which is what index hopping looks like."),
  public_sharing_by_pair = sheet(public_pairs,
    "CDR3aa shared by each pair of patients, and whether that pair sits in the same batch. Concentration in one same-batch pair is a technical warning, not a finding."),
  patient_level_metrics = sheet(pat_level,
    "One row per patient, timepoints averaged. single_visit flags a patient whose average is really one sample."),
  condition_patient_level = sheet(res_cond,
    "CASE vs CTRL, patient as the unit. Read diff and rank_biserial; powered FALSE means no p here could have reached 0.05."),
  within_cond_timepoint = sheet(res_within,
    "Trajectory inside each arm. Coefficients are differences from the reference timepoint in the reference column."),
  between_cond_per_timepoint = sheet(res_between,
    "CASE vs CTRL separately at each visit, unpaired. Same power caveat as condition_patient_level."),
  design_patients_per_cell = sheet(pat_tp,
    "Patients contributing to each condition x timepoint cell. The design behind every skipped row in the manifest."),
  cond_by_timepoint_intx = sheet(res_intx,
    "Full coefficient table. Filter is_interaction TRUE for the terms that answer the question."),
  run_manifest = sheet(manifest,
    "Every step attempted, with status ok / exploratory / skipped. A skipped analysis is otherwise indistinguishable from an empty one.")
)

wb <- createWorkbook()
for (nm in names(BOOK)) {
  s <- BOOK[[nm]]
  if (is.null(s)) { cat("empty or skipped, not written:", nm, "\n"); next }
  stopifnot(nchar(nm) <= 31)
  addWorksheet(wb, nm)
  writeData(wb, nm, s$desc, startRow = 1, startCol = 1, colNames = FALSE)
  writeData(wb, nm, s$d,    startRow = 3, rowNames = FALSE)
  freezePane(wb, nm, firstActiveRow = 4)
}
f <- file.path(clone_root, "subset_data_clonal_TCR_stats.xlsx")
saveWorkbook(wb, f, overwrite = TRUE)
cat("wrote", basename(f), "-", sum(!vapply(BOOK, is.null, logical(1))), "sheets\n")

print(as.data.frame(manifest))
cat("\nSteps run:", sum(manifest$status != "skipped"),
    " skipped:", sum(manifest$status == "skipped"), "\n")
```
